### 2017-11-15 talk.Rmd
add kaggle survey as data example

---
title: "Interpretable Machine Learning"
author: "Christoph Molnar"
date: "December 11, 2017"
output: ioslides_presentation
---

## What is Machine Learning

##
![](images/machine-learning-xkcd.png)

## Interpretability
- The degree to which an observer can understand the cause of a decision.
- Current state in machine learning: Minimize loss, get best prediction, ignore interpretation
- Current state in statistics: Fit model, interpret output, ignore predictive performance
- Ideal world: Get best predictive performance + interpretability

## When is it important

## Different entry points
- Understand data better
- Use only interpretable models (sparse linear models, short decision trees, rules, ...)
- **Apply model-agnostic methods to make black boxes interpretable**

## Model-agnostic methods
- Not bound to model
- Use always the same type of explanation
- ...
- Disadvantage: Can't take advantage of internal structures

## Predicting house prices
TODO: Add example dataset

## PDP

## ICE

## LIME

## Shapley value

## The big picture

## Book

### 2017-11-15 code/prepare-kaggle-data.R

@@ -1,0 +1,56 @@
+## Prepare survey data
+library('dplyr')
+data.dir = 'data/'
+
+survey.raw = read.csv(sprintf('%smultipleChoiceResponses.csv', data.dir), na.strings = c('NA', '-1', '-99', ''))
+exchange = read.csv(sprintf('%sconversionRates.csv', data.dir), row.names = 1)
+survey = survey.raw
+
+usd.to.eur = 1/exchange[exchange$originCountry == 'EUR','exchangeRate']
+# Choose variables of interest
+# GenderSelect, Country, Age, EmploymentStatus, StudentStatus, 
+# CodeWriter, CareerSwitsche, CurrentJobTitleSelect, TitleFit, 
+survey$CompensationAmount = as.numeric(as.character(survey$CompensationAmount))
+
+survey = left_join(survey, exchange, by = c('CompensationCurrency' = 'originCountry'))
+survey$CompensationAmount = usd.to.eur * survey$CompensationAmount * survey$exchangeRate
+survey$CompensationCurrency = NULL
+survey$exchangeRate = NULL
+
+
+## Remove missing and exorbitant salaries
+survey = filter(survey, !is.na(CompensationAmount) & CompensationAmount < 200000)
+
+## Choose some variables 
+
+
+mean.missing = lapply(survey, function(x) mean(is.na(x)))
+mean.missing[mean.missing < 0.1]
+
+survey = select(survey, GenderSelect, Country, Age, EmploymentStatus, CodeWriter, CurrentJobTitleSelect, TitleFit, 
+                CompensationAmount)
+
+nunique = function(x){length(unique(x))}
+lapply(survey, nunique)
+
+
+library(mlr)
+library('ggplot2')
+survey.imp = impute(survey, classes = list(integer = imputeMean(), factor = imputeMode()))
+task = makeRegrTask(data = survey.imp$data, target = 'CompensationAmount')
+
+lrn = makeLearner('regr.randomForest')
+mod = train(lrn, task)
+
+
+pdp1 = generatePartialDependenceData(mod, task, features = c('Age'))
+plotPartialDependence(pdp1) + scale_y_continuous(limits = c(0, NA))
+
+
+pdp1 = generatePartialDependenceData(mod, task, features = c('GenderSelect'))
+ggplot(pdp1$data) + geom_point(aes(x = GenderSelect, y = CompensationAmount), stat='identity') + scale_y_continuous(limits = c(0, NA))
+
+
+
+
+

### 2017-11-15 talk.Rmd
add thoughts on data example

@@
 
-## Predicting house prices
-TODO: Add example dataset
+## Predicting data scientist salaries
+Kaggle survey data
+n  = ...
+Questions asked: ...
+
+## Predicting data scientist scenario
+TODO: Show code to fit the model
+
+## Hypothetical scenario
+We are an HR consulting company that want to use the model to suggest data scientist compensations for companies. 
+We have the model now. 
+Do we use it?
 

### 2017-11-22 talk.Rmd
change order, continue slides

--- 2017-11-15 talk.Rmd
@@
 ## What is Machine Learning
+
+Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
+
+# Let's predict data scientist salaries
+
+## Step 1: Find a problem
+Hypothetical scenario: For a customer, an HR company, we want to predict data scientist salaries.
+
+## Step 2: Find some data
+Kaggle conducted an industry-wide survey of data scientists. 
+Questions asked: 
+- Demographic
+- Salary
+- Are you currently enrolled as a student at a degree granting school?	
+- How adequately do you feel your title describes what you do (or what you did if retired)?	
+- How long have you been learning data science?	
+TODO: Add only relevant questions here for analysis
+
+```{r load-data, warning=FALSE, include=FALSE}
+library('mlr')
+library('ggplot2')
+source('code/prepare-kaggle-data.R')
+```
+
+## Step 3: Throw ML on your data
+```{r learn}
+set.seed(42)
+task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
+n = getTaskSize(task)
+lrn = makeLearner('regr.randomForest')
+mod = train(lrn, task, subset = seq(1, n, 2))
+pred = predict(mod, task = task, subset = seq(2, n, 2))
+## TODO: Find better variables https://www.kaggle.com/kaggle/kaggle-survey-2017/data
+performance(pred, measures = list('mae'=mae))
+```
+## We are done! {.center}
+
+```{r, echo=FALSE}
+knitr::include_graphics("images/done-here.gif")
+```
 
@@
 ## When is it important
+http://people.dbmi.columbia.edu/noemie/papers/15kdd.pdf
+Asthma example
 
 ## Different entry points
-- Understand data better
-- Use only interpretable models (sparse linear models, short decision trees, rules, ...)
-- **Apply model-agnostic methods to make black boxes interpretable**
+- Understand data better $\rightarrow$ do it!
+- Use only interpretable models (sparse linear models, short decision trees, rules, ...) $\rightarrow$ does not always work. 
+- **Apply model-agnostic methods to make black boxes interpretable** $\rightarrow$ cool, tell me more
 
@@
 
-## Predicting house prices
-TODO: Add example dataset
+## Partial dependence plots
+TODO: Fix Tenure variable ordering
+TODO: Rotate labels
+```{r}
+pdp1 = generatePartialDependenceData(mod, task, features =c('Tenure'))
+ggplot(pdp1$data) + geom_point(aes(x=Tenure, y=CompensationAmount)) + 
+  scale_y_continuous(limits=c(0, NA))
+```
 
-## PDP
+## By age
+
+```{r}
+pdp1 = generatePartialDependenceData(mod, task, features =c('Age'))
+plotPartialDependence(pdp1) + scale_y_continuous(limits = c(0, NA))
+```
+
+## By education
+
+```{r}
+pdp1 = generatePartialDependenceData(mod, task, features =c('FormalEducation'))
+plotPartialDependence(pdp1) + scale_y_continuous(limits = c(0, NA))
+```
 

### 2017-11-22 code/countries-to-continents.R

@@ -1,0 +1,49 @@
+## Kudos to https://www.kaggle.com/khsamaha/programming-languages-by-continents
+
+vec.Countries <- c("Algeria","Angola","Benin","Botswana","Burkina","Burundi","Cameroon","Cape Verde",
+                   "Central African Republic","Chad","Comoros","Congo","Congo, Democratic Republic of","Djibouti",
+               "Egypt","Equatorial Guinea","Eritrea","Ethiopia","Gabon","Gambia","Ghana","Guinea","Guinea-Bissau","Ivory Coast",
+               "Kenya","Lesotho","Liberia","Libya","Madagascar","Malawi","Mali","Mauritania","Mauritius","Morocco",
+               "Mozambique","Namibia","Niger","Nigeria","Rwanda","Sao Tomeand Principe","Senegal","Seychelles",
+               "SierraLeone","Somalia","South Africa","South Sudan","Sudan","Swaziland","Tanzania","Taiwan",
+               "Togo","Tunisia","Uganda","Zambia","Zimbabwe","Afghanistan","Bahrain","Bangladesh","Bhutan","Brunei",
+               "Burma(Myanmar)","Cambodia","People's Republic of China","Hong Kong","EastTimor","India","Indonesia","Iran",
+               "Iraq","Israel","Japan","Jordan","Kazakhstan","NorthKorea","South Korea","Kuwait","Kyrgyzstan","Laos",
+               "Lebanon","Malaysia","Maldives","Mongolia","Nepal","Oman","Pakistan","Philippines","Qatar","Russia",
+               "Saudi Arabia","Singapore","SriLanka","Syria","Tajikistan","Thailand","Turkey","Turkmenistan","United Arab Emirates","Uzbekistan",
+               "Vietnam","Yemen","Albania","Andorra","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia",
+               "Cyprus","Czech Republic","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland",
+               "Italy","Latvia","Liechtenstein","Lithuania","Luxembourg","Macedonia","Malta","Moldova","Monaco","Montenegro",
+               "Netherlands","Norway","Poland","Portugal","Romania","SanMarino","Serbia","Slovakia","Slovenia","Spain",
[…]

### 2017-11-22 code/prepare-kaggle-data.R

@@ -6,6 +6,8 @@ survey.raw = read.csv(sprintf('%smultipleChoiceResponses.csv', data.dir), na.str
 exchange = read.csv(sprintf('%sconversionRates.csv', data.dir), row.names = 1)
 survey = survey.raw
 
+survey = filter(survey, CompensationCurrency %in% c('EUR', 'USE'))
+
 usd.to.eur = 1/exchange[exchange$originCountry == 'EUR','exchangeRate']
 # Choose variables of interest
 # GenderSelect, Country, Age, EmploymentStatus, StudentStatus, 
@@ -18,39 +20,44 @@ survey$CompensationCurrency = NULL
 survey$exchangeRate = NULL
 
 
+## Fix country
+source('code/countries-to-continents.R')
+survey$Country = as.character(survey$Country)
+survey$Country[survey$Country == "People 's Republic of China"] <- "People's Republic of China"
+survey$Country[survey$Country == "Republic of China"] <- "People's Republic of China"
+survey$Continent <- Countries.Continents$Continent[match(survey$Country,
+                                                              Countries.Continents$Country)]
+survey$Continent = droplevels(as.factor(survey$Continent))
+
 ## Remove missing and exorbitant salaries
 survey = filter(survey, !is.na(CompensationAmount) & CompensationAmount < 200000)
 
 ## Choose some variables 
-
-
+# GenderSelect:  Select your gender identity. - Selected Choice	
+# Country: Select the country you currently live in.	
+# Age: What's your age?	
+# EmploymentStatus: What's your current employment status?	
+# LanguageRecommendationSelect: What programming language would you recommend a new data scientist learn first? (Select one option) - Selected Choice	
+# LearningDataScienceTime: How long have you been learning data science?	
+# Tenure: How long have you been writing code to analyze data?	
+# FormalEducation: Which level of formal education have you attained?	
 mean.missing = lapply(survey, function(x) mean(is.na(x)))
 mean.missing[mean.missing < 0.1]
-
-survey = select(survey, GenderSelect, Country, Age, EmploymentStatus, CodeWriter, CurrentJobTitleSelect, TitleFit, 
-                CompensationAmount)
+survey$EmploymentStatus = droplevels(survey$EmploymentStatus)
[…]

### 2017-11-23 talk.Rmd
create narrative

--- 2017-11-15 talk.Rmd
@@
 
+##
+
+![](images/wow.jpg)
+
+# Let's predict data scientist salaries
+
 ## What is Machine Learning
 
+Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
+
+![](images/magic.jpg)
+
+## Step 1: Find some data
+Kaggle conducted an industry-wide survey of data scientists. 
+Questions asked:
+
+- Demographic
+- Salary
+- Are you currently enrolled as a student at a degree granting school?	
+- How adequately do you feel your title describes what you do (or what you did if retired)?	
+- How long have you been learning data science?
+
+TODO: Add only relevant questions here for analysis
+
+```{r load-data, warning=FALSE, include=FALSE}
+library('mlr')
+library('ggplot2')
+source('code/prepare-kaggle-data.R')
+```
+
+## Step 2: Throw ML on your data
+```{r learn, warning=FALSE}
+set.seed(42)
+task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
+n = getTaskSize(task)
+lrn = makeLearner('regr.randomForest', importance=TRUE)
+mod = train(lrn, task, subset = seq(1, n, 2))
+pred = predict(mod, task = task, subset = seq(2, n, 2))
+## TODO: Find better variables https://www.kaggle.com/kaggle/kaggle-survey-2017/data
+```
+## Step 3: Profit. We are done! {.center}
+
+```{r, echo=FALSE}
+knitr::include_graphics("images/done-here.gif")
+```
+
 ##
-![](images/machine-learning-xkcd.png)
+
+![](images/Hide-the-pain-harold-phone.jpg)
+
+## "There is a problem with the model!"
+![](images/age.jpeg)
+
+## Individual Conditional Expectation
+
+```{r}
+ice = generatePartialDependenceData(mod, task, features =c('Age'), 
+                                     individual = TRUE)
+plotPartialDependence(ice) + scale_y_continuous(limits=c(0, NA))
+```
+
+## ICE, centered
+
+```{r}
+ice.centered = generatePartialDependenceData(mod, task, 
+                features =c('Age'), individual = TRUE, center = list(Age=20))
+plotPartialDependence(ice.centered)
+```
+
+## Partial dependence plots
+```{r}
+pdp = generatePartialDependenceData(mod, task, features =c('Age'))
+plotPartialDependence(pdp) + scale_y_continuous(limits=c(0, NA))
+```
+
+## "We want to understand the model better!"
+
+## Permutation feature importance
+
+```{r, warning=FALSE}
+feat.imp = getFeatureImportance(mod, type=1)
+knitr::kable(t(feat.imp$res))
+```
+
+## Gender?
+
+```{r}
+pdp = generatePartialDependenceData(mod, task, features =c('GenderSelect'))
+knitr::kable(pdp$data)
+```
+
+##
+
+![](images/hidden-pain-bias.jpg)
+
+##
+
+TODO: Add zoomed in version of Hidden Pain Harolds face
+
+## "We want to understand each decision!"
+
+## Shapley value
+
+Based on Game theory. 
+TODO: Add visualisation
+
+## Shapley value
+
+TODO: Calculate shapley value for one instance.
 
@@
 
-## Predicting house prices
-TODO: Add example dataset
+## Book
+Interested in learning more!
 
-## PDP
-
-## ICE
-
-## LIME
-
-## Shapley value
-
-## The big picture
-
-## Book
+Read my book: https://christophm.github.io/xai-book/

### 2017-11-27 talk.Rmd
improve narrative
improving talk

@@
 ## 
+Scenario: A company approaches you to predict data scientist salaries with machine learning.
 
@@
 library('ggplot2')
+library('tidyr')
+library('lime')
 source('code/prepare-kaggle-data.R')
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, out.width='70%', fig.align='center'}
 knitr::include_graphics("images/done-here.gif")
+```
+
+## "There is a problem with the model!"
+```{r, echo=FALSE, out.width='100%', fig.align='center'}
+knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
+```
+
+## "The older the applicants, the higher the predicted salary, regardless of skills."
+```{r, echo=FALSE, out.width='100%', fig.align='center'}
+knitr::include_graphics("images/age.jpeg")
+```
+
+# Individual Conditional Expectation
+
+##
+
+```{r}
+ice = generatePartialDependenceData(mod, task, features ='Age', 
+                                    individual = TRUE)
+plotPartialDependence(ice) + scale_y_continuous(limits=c(0, NA))
 ```
@@
 
-![](images/Hide-the-pain-harold-phone.jpg)
-
-## "There is a problem with the model!"
-![](images/age.jpeg)
-
-## Individual Conditional Expectation
-
 ```{r}
-ice = generatePartialDependenceData(mod, task, features =c('Age'), 
-                                     individual = TRUE)
-plotPartialDependence(ice) + scale_y_continuous(limits=c(0, NA))
+ice.c = generatePartialDependenceData(mod, task, features ='Age', 
+          individual = TRUE, center = list(Age=20))
+plotPartialDependence(ice.c)
 ```
 
-## ICE, centered
+# Partial dependence plots
 
-```{r}
-ice.centered = generatePartialDependenceData(mod, task, 
-                features =c('Age'), individual = TRUE, center = list(Age=20))
-plotPartialDependence(ice.centered)
-```
+##
 
-## Partial dependence plots
 ```{r}
@@
 
-## Permutation feature importance
+# Permutation feature importance
+
+##
 
 ```{r, warning=FALSE}
-feat.imp = getFeatureImportance(mod, type=1)
-knitr::kable(t(feat.imp$res))
+feat.imp = getFeatureImportance(mod, type=1)$res
+dat = gather(feat.imp, key='Feature', value='Importance') %>% arrange(Importance)
+dat$Feature = factor(dat$Feature, levels = dat$Feature)
+ggplot(dat)  + geom_point(aes(y=Feature, x = Importance))
 ```
 
-## Gender?
+## Gender?!
+```{r, echo=FALSE, fig.align='center', out.width='60%'}
+knitr::include_graphics("images/big-mistake.png")
+```
+
+##
 
@@
 
-![](images/hidden-pain-bias.jpg)
-
-##
-
-TODO: Add zoomed in version of Hidden Pain Harolds face
+```{r, echo=FALSE, fig.align='center'}
+knitr::include_graphics("images/hidden-pain-bias.jpg")
+```
 
@@
 
-## Shapley value
+# LIME
 
-Based on Game theory. 
-TODO: Add visualisation
-
-## Shapley value
-
-TODO: Calculate shapley value for one instance.
+## 
+```{r, echo=FALSE, fig.align='center'}
+set.seed(44)
+```
+```{r, echo=TRUE, fig.align='center'}
+dat = getTaskData(task)
+explanation <- lime(dat, mod)
+# Explain new instance
+explainer <- lime::explain(dat[3, ], explanation, n_features = 3)
+plot_features(explainer, ncol=1)
+```
 
@@
 ## When is it important
-http://people.dbmi.columbia.edu/noemie/papers/15kdd.pdf
-Asthma example
-
-## Different entry points
-- Understand data better $\rightarrow$ do it!
-- Use only interpretable models (sparse linear models, short decision trees, rules, ...) $\rightarrow$ does not always work. 
-- **Apply model-agnostic methods to make black boxes interpretable** $\rightarrow$ cool, tell me more
-
-## Model-agnostic methods
-- Not bound to model
-- Use always the same type of explanation
-- ...
-- Disadvantage: Can't take advantage of internal structures
+Interpretability of machine learning models is when the problem is fundamentally underspecified.
 
@@
 
-Read my book: https://christophm.github.io/xai-book/
+Read my book about "Interpretable Machine Learning"
+https://christophm.github.io/xai-book/

### 2017-11-29 talk.Rmd
finish first draft of slides

@@
 author: "Christoph Molnar"
-date: "December 11, 2017"
-output: ioslides_presentation
+date: "December 4, 2017"
+output:
+  ioslides_presentation:
+    css: styles.css
 ---
 
-## 
+##  {.emphasizedabit}
 Scenario: A company approaches you to predict data scientist salaries with machine learning.
 
-![](images/wow.jpg)
+##
+
+```{r, echo=FALSE, fig.align='center'}
+knitr::include_graphics("images/wow.jpg")
+```
 
@@
 
-![](images/magic.jpg)
+```{r, echo=FALSE, fig.align='center'}
+knitr::include_graphics("images/magic.jpg")
+```
 
@@
 Kaggle conducted an industry-wide survey of data scientists. 
-Questions asked:
+https://www.kaggle.com/kaggle/kaggle-survey-2017
 
-- Demographic
-- Salary
-- Are you currently enrolled as a student at a degree granting school?	
-- How adequately do you feel your title describes what you do (or what you did if retired)?	
-- How long have you been learning data science?
+Information asked:
 
-TODO: Add only relevant questions here for analysis
+- Compensation
+- Demographics
+- Job title
+- Experience
+- ...
 
@@
 task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
-n = getTaskSize(task)
 lrn = makeLearner('regr.randomForest', importance=TRUE)
-mod = train(lrn, task, subset = seq(1, n, 2))
-pred = predict(mod, task = task, subset = seq(2, n, 2))
-## TODO: Find better variables https://www.kaggle.com/kaggle/kaggle-survey-2017/data
+mod = train(lrn, task)
 ```
@@
 
-## "There is a problem with the model!"
+##  {.emphasizedabit}
+
+"There is a problem with the model!"
+
+## What problem?
+
 ```{r, echo=FALSE, out.width='100%', fig.align='center'}
@@
 
-## "We want to understand the model better!"
+##  {.emphasizedabit}
+
+"We want to understand the model better!"
 
@@
 ## Gender?!
-```{r, echo=FALSE, fig.align='center', out.width='60%'}
+```{r, echo=FALSE, fig.align='center', out.width='80%'}
 knitr::include_graphics("images/big-mistake.png")
@@
 ```{r}
-pdp = generatePartialDependenceData(mod, task, features =c('GenderSelect'))
-knitr::kable(pdp$data)
+pdp = generatePartialDependenceData(mod, task, features =c('Gender'))
+ggplot(pdp$data) + geom_point(aes(x=Gender, y=CompensationAmount)) + 
+  geom_segment(aes(x=Gender, xend=Gender, yend=CompensationAmount), y=0) + 
+  scale_y_continuous(limits=c(0, NA)) + 
+  theme(axis.text.x = element_text(angle = 10, hjust = 1))
+```
+
+##
+
+```{r, echo=FALSE, fig.align='center', out.width='100%'}
+knitr::include_graphics("images/facepalm.gif")
 ```
@@
 
-## "We want to understand each decision!"
+##
+
+```{r, echo=FALSE, fig.align='center', out.width='40%'}
+knitr::include_graphics("images/angry.gif")
+```
 
@@
 
-## Interpretability
-- The degree to which an observer can understand the cause of a decision.
-- Current state in machine learning: Minimize loss, get best prediction, ignore interpretation
-- Current state in statistics: Fit model, interpret output, ignore predictive performance
-- Ideal world: Get best predictive performance + interpretability
-
-## When is it important
-Interpretability of machine learning models is when the problem is fundamentally underspecified.
-
-## Book
-Interested in learning more!
+## Interested in learning more?
 
@@
 https://christophm.github.io/xai-book/
+
+```{r, echo=FALSE, fig.align='center', out.width='80%'}
+knitr::include_graphics("images/book.png")
+```

### 2017-11-29 code/countries-to-continents.R

@@ -1,49 +1,0 @@
-## Kudos to https://www.kaggle.com/khsamaha/programming-languages-by-continents
-
-vec.Countries <- c("Algeria","Angola","Benin","Botswana","Burkina","Burundi","Cameroon","Cape Verde",
-                   "Central African Republic","Chad","Comoros","Congo","Congo, Democratic Republic of","Djibouti",
-               "Egypt","Equatorial Guinea","Eritrea","Ethiopia","Gabon","Gambia","Ghana","Guinea","Guinea-Bissau","Ivory Coast",
-               "Kenya","Lesotho","Liberia","Libya","Madagascar","Malawi","Mali","Mauritania","Mauritius","Morocco",
-               "Mozambique","Namibia","Niger","Nigeria","Rwanda","Sao Tomeand Principe","Senegal","Seychelles",
-               "SierraLeone","Somalia","South Africa","South Sudan","Sudan","Swaziland","Tanzania","Taiwan",
-               "Togo","Tunisia","Uganda","Zambia","Zimbabwe","Afghanistan","Bahrain","Bangladesh","Bhutan","Brunei",
-               "Burma(Myanmar)","Cambodia","People's Republic of China","Hong Kong","EastTimor","India","Indonesia","Iran",
-               "Iraq","Israel","Japan","Jordan","Kazakhstan","NorthKorea","South Korea","Kuwait","Kyrgyzstan","Laos",
-               "Lebanon","Malaysia","Maldives","Mongolia","Nepal","Oman","Pakistan","Philippines","Qatar","Russia",
-               "Saudi Arabia","Singapore","SriLanka","Syria","Tajikistan","Thailand","Turkey","Turkmenistan","United Arab Emirates","Uzbekistan",
-               "Vietnam","Yemen","Albania","Andorra","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia",
-               "Cyprus","Czech Republic","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland",
-               "Italy","Latvia","Liechtenstein","Lithuania","Luxembourg","Macedonia","Malta","Moldova","Monaco","Montenegro",
-               "Netherlands","Norway","Poland","Portugal","Romania","SanMarino","Serbia","Slovakia","Slovenia","Spain",
[…]

### 2017-11-29 code/prepare-kaggle-data.R

@@ -1,12 +1,13 @@
 ## Prepare survey data
 library('dplyr')
+library('mlr')
 data.dir = 'data/'
 
 survey.raw = read.csv(sprintf('%smultipleChoiceResponses.csv', data.dir), na.strings = c('NA', '-1', '-99', ''))
 exchange = read.csv(sprintf('%sconversionRates.csv', data.dir), row.names = 1)
 survey = survey.raw
 
-survey = filter(survey, CompensationCurrency %in% c('EUR', 'USE'))
+survey = filter(survey, CompensationCurrency %in% c('EUR'))
 
 usd.to.eur = 1/exchange[exchange$originCountry == 'EUR','exchangeRate']
 # Choose variables of interest
@@ -18,16 +19,6 @@ survey = left_join(survey, exchange, by = c('CompensationCurrency' = 'originCoun
 survey$CompensationAmount = usd.to.eur * survey$CompensationAmount * survey$exchangeRate
 survey$CompensationCurrency = NULL
 survey$exchangeRate = NULL
-
-
-## Fix country
-source('code/countries-to-continents.R')
-survey$Country = as.character(survey$Country)
-survey$Country[survey$Country == "People 's Republic of China"] <- "People's Republic of China"
-survey$Country[survey$Country == "Republic of China"] <- "People's Republic of China"
-survey$Continent <- Countries.Continents$Continent[match(survey$Country,
-                                                              Countries.Continents$Country)]
-survey$Continent = droplevels(as.factor(survey$Continent))
 
 ## Remove missing and exorbitant salaries
 survey = filter(survey, !is.na(CompensationAmount) & CompensationAmount < 200000)
@@ -44,20 +35,17 @@ survey = filter(survey, !is.na(CompensationAmount) & CompensationAmount < 200000
 mean.missing = lapply(survey, function(x) mean(is.na(x)))
 mean.missing[mean.missing < 0.1]
 survey$EmploymentStatus = droplevels(survey$EmploymentStatus)
-survey = select(survey, GenderSelect, Continent, Age, EmploymentStatus, CurrentJobTitleSelect, TitleFit, 
+survey = select(survey, Gender=GenderSelect, Age, CodeWriter, EmploymentStatus, JobTitle=CurrentJobTitleSelect, 
                 CompensationAmount, LanguageRecommendationSelect, LearningDataScienceTime,
-                FormalEducation, Tenure)
[…]

### 2017-11-30 talk.Rmd
fix one slide

@@
 ```{r learn, warning=FALSE}
+library('mlr')
 set.seed(42)

### 2017-12-04 talk.Rmd
update link
add kaggle obd license disclaimer

@@
 - ...
+
+<font size="2">Contains information from Kaggle ML and Data Science Survey, 2017, which is made available here under the Open Database License (ODbL).</font>
 
@@
 Read my book about "Interpretable Machine Learning"
-https://christophm.github.io/xai-book/
+https://christophm.github.io/interpretable-ml-book/
 

### 2017-12-04 talk.Rmd
add references

@@
 ```
-## Step 3: Profit. We are done! {.center}
+## Step 3: Profit. {.center}
 
@@
 
+<font size="2">Goldstein, A., Kapelner, A., Bleich, J., & Pitkin, E.
+(2013).
+Peeking Inside the Black Box: Visualizing Statistical Learning with Plots of Individual Conditional Expectation, 1–22.
+https://doi.org/10.1080/10618600.2014.907095 </font>
+
 ##
@@
 
+<font size="2">Friedman, J. H. (1999). Greedy Function Approximation : A Gradient Boosting Machine. North, 1(3), 1–10. https://doi.org/10.2307/2699986 </font>
+
 ##  {.emphasizedabit}
@@
 ```
+
+<font size="2">Breiman, Leo. "Random forests." Machine learning 45.1 (2001): 5-32. </font>
 
@@
 set.seed(44)
+dat = getTaskData(task)
 ```
 ```{r, echo=TRUE, fig.align='center'}
-dat = getTaskData(task)
 explanation <- lime(dat, mod)
-# Explain new instance
 explainer <- lime::explain(dat[3, ], explanation, n_features = 3)
@@
 ```
+
+<font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
 

### 2018-03-27 talk.Rmd
pdates slides with iml package

@@
 task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
-lrn = makeLearner('regr.randomForest', importance=TRUE)
+lrn = makeLearner('regr.gbm', interaction.depth = 4, n.trees = 500)
 mod = train(lrn, task)
@@
 ```{r}
-ice = generatePartialDependenceData(mod, task, features ='Age', 
-                                    individual = TRUE)
-plotPartialDependence(ice) + scale_y_continuous(limits=c(0, NA))
+library("iml")
+X = survey.dat[-which(names(survey.dat) == "CompensationAmount")]
+predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
+ice = Ice$new(predictor, feature ='Age')
+ice$plot()
 ```
@@
 ```{r}
-ice.c = generatePartialDependenceData(mod, task, features ='Age', 
-          individual = TRUE, center = list(Age=20))
-plotPartialDependence(ice.c)
+ice$center(20)$plot()
 ```
@@
 ```{r}
-pdp = generatePartialDependenceData(mod, task, features =c('Age'))
-plotPartialDependence(pdp) + scale_y_continuous(limits=c(0, NA))
+pdp = PartialDependence$new(predictor, feature = "Age")
+pdp$plot()
 ```
@@
 ```{r, warning=FALSE}
-feat.imp = getFeatureImportance(mod, type=1)$res
-dat = gather(feat.imp, key='Feature', value='Importance') %>% arrange(Importance)
-dat$Feature = factor(dat$Feature, levels = dat$Feature)
-ggplot(dat)  + geom_point(aes(y=Feature, x = Importance))
+feat.imp = FeatureImp$new(predictor, loss = 'mae')
+feat.imp$plot()
 ```
@@
 ```{r}
-pdp = generatePartialDependenceData(mod, task, features =c('Gender'))
-ggplot(pdp$data) + geom_point(aes(x=Gender, y=CompensationAmount)) + 
-  geom_segment(aes(x=Gender, xend=Gender, yend=CompensationAmount), y=0) + 
-  scale_y_continuous(limits=c(0, NA)) + 
-  theme(axis.text.x = element_text(angle = 10, hjust = 1))
+pdp = PartialDependence$new(predictor, feature="Gender")
+pdp$plot() 
 ```
@@
 set.seed(44)
-dat = getTaskData(task)
 ```
-```{r, echo=TRUE, fig.align='center'}
-explanation <- lime(dat, mod)
-explainer <- lime::explain(dat[3, ], explanation, n_features = 3)
-plot_features(explainer, ncol=1)
+
+```{r, echo=TRUE, fig.align='center', message = FALSE}
+explanation = LocalModel$new(predictor)
+explanation$explain(X[5, ])$plot()
 ```
@@
 <font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
+
+# Shapley
+
+## 
+```{r, echo=FALSE, fig.align='center'}
+set.seed(44)
+```
+
+```{r, echo=TRUE, fig.align='center'}
+explanation = Shapley$new(predictor)
+explanation$explain(X[5, ])$plot()
+```
 

### 2018-05-29 talk-40min.Rmd
continues talk

---
title: "Interpretable Machine Learning"
author: "Christoph Molnar"
date: "December 4, 2017"
output: ioslides_presentation
css: styles.css
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
  warning = FALSE,
  message = FALSE, 
  fig.align='center')
```

##  {.emphasizedabit}
A client wants you to predict data scientist salaries with machine learning.

##

```{r, echo=FALSE, fig.align='center'}
knitr::include_graphics("images/wow.jpg")
```

# Let's predict data scientist salaries

## What is Machine Learning

Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.

```{r, echo=FALSE, fig.align='center'}
knitr::include_graphics("images/magic.jpg")
```

## Step 1: Find some data
Kaggle conducted an industry-wide survey of data scientists. 
https://www.kaggle.com/kaggle/kaggle-survey-2017

Information asked:

- Compensation
- Demographics
- Job title
- Experience
- ...

<font size="2">Contains information from Kaggle ML and Data Science Survey, 2017, which is made available here under the Open Database License (ODbL).</font>

```{r load-data,include=FALSE}
library('mlr')
library('ggplot2')
library('tidyr')
library('lime')
source('code/prepare-kaggle-data.R')
```

## Step 2: Throw ML on your data
```{r learn}
library('mlr')
set.seed(42)
task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
lrn = makeLearner('regr.randomForest')
mod = train(lrn, task)
```
## Step 3: Profit. {.center}

```{r, echo=FALSE, out.width='70%'}
knitr::include_graphics("images/done-here.gif")
```

##  {.emphasizedabit}

"There is a problem with the model!"

## What problem?

```{r, echo=FALSE}
knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
```

## "The older the applicants, the higher the predicted salary, regardless of skills."
```{r, echo=FALSE, out.width='100%'}
knitr::include_graphics("images/age.jpeg")
```

# Partial Dependence Plots

##

```{r}
library("iml")
X = survey.dat[-which(names(survey.dat) == "CompensationAmount")]
predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
pd = Partial$new(predictor, feature ='Age')
pd$plot() 
```

<font size="2">Goldstein, A., Kapelner, A., Bleich, J., & Pitkin, E. (2013). Peeking Inside the Black Box: Visualizing Statistical Learning with Plots of Individual Conditional Expectation, 1–22. https://doi.org/10.1080/10618600.2014.907095 </font>

<font size="2">Friedman, J. H. (1999). Greedy Function Approximation : A Gradient Boosting Machine. North, 1(3), 1–10. https://doi.org/10.2307/2699986 </font>

##  {.emphasizedabit}

Client: "We want to understand the model better!"

# Permutation feature importance

##

```{r, warning=FALSE}
feat.imp = FeatureImp$new(predictor, loss = 'mae')
feat.imp$plot()
```

<font size="2">Breiman, Leo. "Random forests." Machine learning 45.1 (2001): 5-32. </font>

## Gender?!
```{r, echo=FALSE, out.width='80%'}
knitr::include_graphics("images/big-mistake.png")
```

##

```{r}
pdp = Partial$new(predictor, feature="Gender")
pdp$plot() 
```

```{r}
pdp$results[pdp$results$.type == "pdp", c("Gender", ".y.hat")]
```

##

```{r, echo=FALSE}
knitr::include_graphics("images/hidden-pain-bias.jpg")
```

##

```{r, echo=FALSE, out.width='40%'}
knitr::include_graphics("images/angry.gif")
```

# LIME

## 
```{r, echo=FALSE}
set.seed(44)
```

```{r}
explanation = LocalModel$new(predictor)
explanation$explain(X[15, ])
explanation$plot()
```

<font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>

# Shapley

## 
```{r}
set.seed(44)
```

```{r}
explanation = Shapley$new(predictor)
explanation$explain(X[15, ])
explanation$plot()
```

# Some Theory

## What is interpretability?

Interpretability is the degree to which a human can understand the cause of a decision. 
But we don't really have a good way to measure that really. Not as easy as benchmarking ML algorithms.

Some things:
sparsity, contrastiveness, social, focus on the abnormal, truthful, coherent with prior beliefs, general and probable.

<font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>

## When do we need interpretability?

We need interpretability when the loss function does not cover all constraints.

(like, everything)

## When don't we need interpretability?

When we can capture everything in the loss function and the data collection.
only causal relationships.
perfect operationalization of features.

## What tools do we have?

- Intrinsically interpretable models
- Tools for analyzing specific black box models
- Model-agnostic tools: analysing any black box model
- Interpretability through making data explainable

### Interpretable Models

TODO: Graphic with schematic interpretable model

### Interpretable Models

TODO: Graphic with tree+if-then-rules+decision tree

## Model-agnostic methods
TODO: Graphic with schematic Model-agnostic methods

## Model-agnostic methods

- Partial dependence plots
- LIME
- ...

## Model-specific methods
TODO: Graphic with schematic Model-specific methods

TODO: add images from papers

## Model-specific methods

- RandomForestExplainer
- Visualizing activations of neural networks
-

## Interested in learning more?

Read my book about "Interpretable Machine Learning"
https://christophm.github.io/interpretable-ml-book/

```{r, echo=FALSE, fig.align='center', out.width='30%'}
knitr::include_graphics("images/book.jpg")
```

### 2018-05-29 code/prepare-kaggle-data.R

@@ -9,7 +9,7 @@ survey = survey.raw
 
 survey = filter(survey, CompensationCurrency %in% c('EUR'))
 
-# usd.to.eur = 1/exchange[exchange$originCountry == 'EUR','exchangeRate']
+usd.to.eur = 1/exchange[exchange$originCountry == 'EUR','exchangeRate']
 # Choose variables of interest
 # GenderSelect, Country, Age, EmploymentStatus, StudentStatus, 
 # CodeWriter, CareerSwitsche, CurrentJobTitleSelect, TitleFit, 

### 2018-05-30 talk-40min.Rmd
continues talk

@@
 ```
+
+TODO: Insert image
+
 ## Step 3: Profit. {.center}
@@
 
+## The Problem
+
+- Opaque decision making by machine learning algorithms
+- Examples: AI doctor, self-driving cars, creditworthiness
+- Trust as a user: Should I start severe therapy, because the machine said so?
+- Debugging as a practitioner: Why did the algorithm miss-classify sample X? Did it learn generalizable features?
+
+##
+
+![](images/black-box.png)
+
 # Some Theory
@@
 
+- Debugging
+- Increasing trust
+
+### Example: Doctor and ML
+
+### Example: Predictive CV screening
+
+- Nasty feedback loops
+- Biased data
+- Proxies instead of causality
+-
+
+## Example:
+
 ## When don't we need interpretability?
@@
 perfect operationalization of features.
+
+Things that work well. 
+Well defined problems
+
+## Example: Digit recognition
+
+![](images/mnist.jpeg)
 
@@
 
-### Interpretable Models
+## Interpretable Models
 
-TODO: Graphic with schematic interpretable model
+![](images/black-box.png)
 
-### Interpretable Models
+## Interpretable Models
 
-TODO: Graphic with tree+if-then-rules+decision tree
+![](images/white-box.png)
+
+## Interpretable Models
+
+- Decision trees
+- Decision rules
+- Generalized regression models
+- ...
+
+## Model-specific methods
+
+![](images/black-box.png)
+
+## Model-specific methods
+
+![](images/specific-black-box.png)
+
+## Model-specific methods
+
+- RandomForestExplainer
+- Visualizing activations of neural networks
+-
+
+TODO: add images from papers
+## Model-agnostic methods
+![](images/black-box.png)
 
 ## Model-agnostic methods
-TODO: Graphic with schematic Model-agnostic methods
+![](images/agnostic-black-box.png)
 
@@
 
-## Model-specific methods
-TODO: Graphic with schematic Model-specific methods
+## Data-focused methods
+![](images/black-box.png)
 
-TODO: add images from papers
+## Data-focused methods
+![](images/data-box.png)
 
-## Model-specific methods
+## Data-focused methods
 
-- RandomForestExplainer
-- Visualizing activations of neural networks
--
+- Prototypes: What is the prototype of a certain class
+- Criticisms: Which data points are unusual for a certain class. 
+- Counterfactuals: What do I need to change for an instance to change the prediction?
+- Anchors: What features do I have to anchor to keep prediction the same?
+
+TODO: Cite
+
+## Outcomes of interpretability method:
+
+- Model
+- Data point
+- Interpretable unit (like a weight)
+- Summary statistic for a feature
 

### 2018-06-04 talk-40min.Rmd
continues talk

@@
 author: "Christoph Molnar"
-date: "December 4, 2017"
-output: ioslides_presentation
+date: "June 7, 2018"
+output: 
+  ioslides_presentation:
+    widescreen: false
+    smaller: false
 css: styles.css
@@
   message = FALSE, 
-  fig.align='center')
+  fig.align='center', 
+  out.width = '100%')
 ```
@@
 
-```{r, echo=FALSE, fig.align='center'}
+```{r, echo=FALSE, out.height = '100%', out.width=NULL}
 knitr::include_graphics("images/wow.jpg")
@@
 
-```{r, echo=FALSE, fig.align='center'}
+```{r, echo=FALSE, out.height = '100%', out.width=NULL}
 knitr::include_graphics("images/magic.jpg")
@@
 
-TODO: Insert image
+```{r, echo=FALSE}
+knitr::include_graphics("images/comp-dog.gif")
+```
 
@@
 
-```{r, echo=FALSE, out.width='70%'}
+```{r, echo=FALSE, out.height=NULL, out.width='100%'}
 knitr::include_graphics("images/done-here.gif")
@@
 
-"There is a problem with the model!"
+Client: "There is a problem with the model!"
 
-## What problem?
+## {.emphasizedabit}
 
-```{r, echo=FALSE}
+"What problem?""
+
+```{r, echo=FALSE, out.height='50%'}
 knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
@@
 
-## "The older the applicants, the higher the predicted salary, regardless of skills."
-```{r, echo=FALSE, out.width='100%'}
+## {.emphasizedabit}
+
+ Client: "Model predicts high salaries for old yet unskilled people."
+
+```{r, echo=FALSE, out.height='30%'}
 knitr::include_graphics("images/age.jpeg")
@@
 
-```{r}
-pdp$results[pdp$results$.type == "pdp", c("Gender", ".y.hat")]
-```
-
 ##
@@
 
+# Some Theory
+
+## The Problem
+
+```{r, echo=FALSE, fig.align='center'}
+knitr::include_graphics("images/black-box.png")
+```
+
 ## The Problem
@@
 - Opaque decision making by machine learning algorithms
-- Examples: AI doctor, self-driving cars, creditworthiness
-- Trust as a user: Should I start severe therapy, because the machine said so?
+- Bias and mistakes in the data
+- Trust as a user: Should I start a therapy with sever side effects, because the machine said so?
 - Debugging as a practitioner: Why did the algorithm miss-classify sample X? Did it learn generalizable features?
-
-##
-
-![](images/black-box.png)
-
-# Some Theory
 
@@
 
-Some things:
-sparsity, contrastiveness, social, focus on the abnormal, truthful, coherent with prior beliefs, general and probable.
+Interpretability is also a mean to look at the data and possible issues with them.
 
@@
 
-(like, everything)
+- Decisions about humans.
+- Critical applications that decide about life and death.
+- Newly developed systems with unknown consequences.
+- Models using proxies instead of causal inputs.
+- Debugging the models.
+- Increasing trust.
 
-- Debugging
-- Increasing trust
+### Examples where we need interpretability
 
-### Example: Doctor and ML
-
-### Example: Predictive CV screening
-
-- Nasty feedback loops
-- Biased data
-- Proxies instead of causality
--
-
-## Example:
+- Credit scores
+- Autonomous vehicles
+- Medical recommendations
+- Automated diagnoses
 
@@
 
-![](images/black-box.png)
+```{r, echo = FALSE}
+knitr::include_graphics("images/black-box.png")
+```
 
@@
 
-![](images/white-box.png)
+```{r, echo=FALSE}
+knitr::include_graphics("images/white-box.png")
+```
 
@@
 
-![](images/black-box.png)
+```{r, echo=FALSE}
+knitr::include_graphics("images/black-box.png")
+```
 
@@
 
-![](images/specific-black-box.png)
+```{r, echo=FALSE}
+knitr::include_graphics("images/specific-black-box.png")
+```
 
@@
 TODO: add images from papers
-## Model-agnostic methods
-![](images/black-box.png)
 
 ## Model-agnostic methods
-![](images/agnostic-black-box.png)
+
+```{r, echo=FALSE}
+knitr::include_graphics("images/black-box.png")
+```
+
+## Model-agnostic methods
+
+```{r, echo=FALSE}
+knitr::include_graphics("images/agnostic-black-box.png")
+```
 
@@
 
-## Data-focused methods
-![](images/black-box.png)
+## Example-focused methods
 
-## Data-focused methods
-![](images/data-box.png)
+```{r, echo=FALSE}
+knitr::include_graphics("images/black-box.png")
+```
 
-## Data-focused methods
+## Example-focused methods
+```{r, echo=FALSE}
+knitr::include_graphics("images/data-box.png")
+```
+
+## Example-focused methods
 

### 2018-08-17 talk-40min.Rmd
improves talk

@@
 
-##
-
-```{r, echo=FALSE, out.height = '100%', out.width=NULL}
-knitr::include_graphics("images/wow.jpg")
-```
+## {.center data-background=images/wow.jpg data-background-size=contain}
 
@@
 
-```{r, echo=FALSE, out.height = '100%', out.width=NULL}
-knitr::include_graphics("images/magic.jpg")
-```
+## {data-background=images/magic.jpg data-background-size=contain}
+
+# Step 1: Find some data
 
 ## Step 1: Find some data
+
 Kaggle conducted an industry-wide survey of data scientists. 
@@
 
-## Step 2: Throw ML on your data
+# Step 2: Throw ML on your data
+
 ```{r learn}
@@
 
-```{r, echo=FALSE}
-knitr::include_graphics("images/comp-dog.gif")
-```
+## {.center data-background=images/comp-dog.gif data-background-size=contain}
 
-## Step 3: Profit. {.center}
+<div class="notes">
+- Random Forest
+- Target: Compensation
+- All features that you found in the data
+- Train, test, ship
+</div>
 
-```{r, echo=FALSE, out.height=NULL, out.width='100%'}
-knitr::include_graphics("images/done-here.gif")
+# Step 3: Profit. {.center}
+
+## {data-background=images/done-here.gif data-background-size=contain}
+```{r, echo=FALSE, out.width='80%', include = FALSE}
+knitr::include_graphics("images/done-here.gif" )
 ```
@@
 
-## {.emphasizedabit}
+##  {.emphasizedabit .center data-background=images/Hide-the-pain-harold-phone.jpg data-background-size=cover}
 
-"What problem?""
+"What problem?"
 
-```{r, echo=FALSE, out.height='50%'}
+```{r, echo=FALSE, out.height='50%', include=FALSE}
 knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
@@
 
-```{r, echo=FALSE, out.height='30%'}
+```{r, echo=FALSE, fig.width=16, include=FALSE}
 knitr::include_graphics("images/age.jpeg")
@@
 
-# Partial Dependence Plots
+##  {.emphasizedabit .center data-background=images/black-box.gif data-background-size=cover}
 
-##
+<div class="white">
+Looking inside the black box
+</div>
 
-```{r}
+# How do the features influence my predictions?
+
+## Partial Dependence Plot
+
+```{r out.width="80%"}
 library("iml")
@@
 
+<div class="notes">
+- PDP answers: How does the prediction change when a feature input changes?
+- Average of ICE curves
+</div>
+
 ##  {.emphasizedabit}
@@
 
-# Permutation feature importance
+# What are the most important features?
 
-##
+## Permutation feature importance
 
@@
 
-## Gender?!
-```{r, echo=FALSE, out.width='80%'}
+## Gender?! {.emphasizedabit .center data-background=images/big-mistake.png data-background-size=contain}
+
+```{r, echo=FALSE, out.width='80%', include=FALSE}
 knitr::include_graphics("images/big-mistake.png")
 ```
+
+# What's the influence of gender on the prediction?
 
@@
 
-##
+## {.center data-background=images/hidden-pain-bias.jpg data-background-size=contain}
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, include = FALSE}
 knitr::include_graphics("images/hidden-pain-bias.jpg")
@@
 
-##
+## {.center data-background=images/angry.gif data-background-size=contain}
 
-```{r, echo=FALSE, out.width='40%'}
+```{r, echo=FALSE, out.width='40%', include = FALSE}
 knitr::include_graphics("images/angry.gif")
@@
 
+# Explaining individual predictions
+
 # LIME
+
+## {.center data-background=images/lime-fitting-1.png data-background-size=contain}
 
@@
 
-# Shapley
+# Shapley Value
+
+##
+
+TODO: find image for explaining shapley value
 
@@
 ```
-
-## The Problem
-
+<div class="notes">
 - Opaque decision making by machine learning algorithms
@@
 - Debugging as a practitioner: Why did the algorithm miss-classify sample X? Did it learn generalizable features? 
+</div>
 
@@
 
+<div class="notes">
 Interpretability is the degree to which a human can understand the cause of a decision. 
@@
 <font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>
+</div>
 
@@
 
+<div class="notes">
 We need interpretability when the loss function does not cover all constraints.
@@
 - Increasing trust.
+</div>
 
-### Examples where we need interpretability
+## Examples where we need interpretability
 
+TODO: Insert images
 - Credit scores
@@
 
-## Example: Digit recognition
-
-![](images/mnist.jpeg)
+## Example: Digit recognition {.center data-background=images/mnist.jpeg data-background-size=cover}
 
 ## What tools do we have?
+
+TODO: replace with image
 
@@
 
-## Interpretable Models
-
+<div class="notes">
 - Decision trees
@@
 - ...
+</div>
 
@@
 
-## Model-specific methods
-
-- RandomForestExplainer
+<div class="notes">
 - Visualizing activations of neural networks
--
-
-TODO: add images from papers
+- Random Forest specific feature importance
+- xgboost explainer
+- problem: new models, not comparable between models, missing flexibility
+</div>
 
@@
 
-## Model-agnostic methods
-
+<div class="notes">
 - Partial dependence plots
@@
 
-## Example-focused methods
-
+<div class="notes">
 - Prototypes: What is the prototype of a certain class
@@
 - Anchors: What features do I have to anchor to keep prediction the same?
-
-TODO: Cite
-
-## Outcomes of interpretability method:
-
-- Model
-- Data point
-- Interpretable unit (like a weight)
-- Summary statistic for a feature
+</div>
 

### 2018-08-17 talk.Rmd
improves talk

@@
 ## "The older the applicants, the higher the predicted salary, regardless of skills."
-```{r, echo=FALSE, out.width='100%', fig.align='center'}
+```{r, echo=FALSE, out.height='100%', fig.align='center'}
 knitr::include_graphics("images/age.jpeg")
@@
 predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
-ice = Ice$new(predictor, feature ='Age')
+ice = Partial$new(predictor, feature ='Age')
 ice$plot()

### 2018-08-20 talk-40min.Rmd
finishes first draft of slides

@@
 
-## What is Machine Learning
+# What is Machine Learning?
 
+<div class="notes">
 Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
+</div>
 
@@
 
-# Step 3: Profit. {.center}
+# Step 3: Profit {.center}
 
@@
 
-# LIME
+# Local Models (LIME)
 
@@
 
-##
-
-TODO: find image for explaining shapley value
+## {.center data-background=images/team.jpg data-background-size=contain}
 
@@
 
-## What is interpretability?
+# When do we need interpretability?
 
-<div class="notes">
-Interpretability is the degree to which a human can understand the cause of a decision. 
-But we don't really have a good way to measure that really. Not as easy as benchmarking ML algorithms.
+## {.center data-background=images/credit.jpg data-background-size=cover}
 
-Interpretability is also a mean to look at the data and possible issues with them.
+## {.center data-background=images/pill.jpg data-background-size=cover}
 
-<font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>
-</div>
+## {.center data-background=images/death.jpg data-background-size=cover}
 
-## When do we need interpretability?
+## {.center data-background=images/people.jpg data-background-size=cover}
 
@@
 
-## Examples where we need interpretability
+# When do we NOT need interpretability?
 
-TODO: Insert images
-- Credit scores
-- Autonomous vehicles
-- Medical recommendations
-- Automated diagnoses
+##  {.center data-background=images/mnist.jpeg data-background-size=cover}
 
-## When don't we need interpretability?
-
+<div class="notes">
 When we can capture everything in the loss function and the data collection.
@@
 Well defined problems
+</div>
 
-## Example: Digit recognition {.center data-background=images/mnist.jpeg data-background-size=cover}
+# What tools do we have?
 
-## What tools do we have?
-
-TODO: replace with image
-
+<div class="notes">
 - Intrinsically interpretable models
@@
 - Interpretability through making data explainable
+</div>
 
@@
 ```
+
+# Backup slides
+
+# What is interpretability?
+
+<div class="notes">
+Interpretability is the degree to which a human can understand the cause of a decision. 
+But we don't really have a good way to measure that really. Not as easy as benchmarking ML algorithms.
+
+Interpretability is also a mean to look at the data and possible issues with them.
+
+<font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>
+</div>
+
+# Statistical Modeling: The Two Cultures
+
+<div class="notes">
+Paper by Leo Breiman (Random Forest)
+data modeling culture (statistics) vs algorithmic modeling culture (ml)
+tradeoff between interpretability and predictive performance
+example: tree vs. random forest, logistic regression vs. neural network
+</div>
+
+# What makes an explanation human-friendly?
+
+<div class="notes">
+short: 1-2 causes
+counterfactual: compares to other output
+truthful
+social context
+consistent with prior beliefs
+probable
+focuses on abnormal causes
+</div>

### 2018-09-28 poster/poster_2018.Rnw
starts poster

\documentclass[portrait, a0paper]{tikzposter}
\usepackage{url}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{pdfpages}
\usepackage{xcolor}
\usepackage{dsfont}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{csquotes}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{enumerate}
\usepackage[absolute,overlay]{textpos}
\usepackage{psfrag}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{eqnarray}
\usepackage{arydshln}
\usepackage{tabularx}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{colortbl}
\usepackage{mathtools}
\usepackage{wrapfig}
\usepackage{xargs}
\usepackage{subfig}

\definecolorpalette{BlueGrayOrange}{
    \definecolor{colorOne}{HTML}{C0C0C0}
    \definecolor{colorTwo}{HTML}{CCCCCC}
    \definecolor{colorThree}{HTML}{009440}
}
\usepackage{mathtools}
\usepackage{hyperref}
\usetheme{Desert}
\usecolorstyle[colorPalette=BlueGrayOrange]{Russia}
\usebackgroundstyle{Rays}
\title{Interpretable Machine Learning}
\author{Christoph Molnar, Prof. Bernd Bischl}
\institute{Department of Statistics, Ludwig Maximilians University\\[0.5cm] \url{christoph.molnar@stat.uni-muenchen.de}}
%\titlegraphic{\hspace{-3200px}\includegraphics[scale=0.3]{figures/logo.png}}
\usetitlestyle{Empty}
\colorlet{titlefgcolor}{black}
\colorlet{titlebgcolor}{colorOne}
\makeatletter
\renewcommand\TP@maketitle{%
   %\hspace{-200px}\includegraphics[scale=0.35]{figures/logo.png}\hspace{100px}
     \tikz[remember picture, overlay]\node[scale=0.8,anchor=east,xshift=0.65\linewidth,yshift=4cm,inner sep=0pt] {%
      \includegraphics[scale=6.5]{figures/logo.pdf} %green 1.3
   };
   \hspace{500px}
   \centering
   \begin{minipage}[b]{0.7\linewidth}
        \centering
        \color{titlefgcolor}
        {\bfseries \huge \sc \@title \par}
        \vspace*{1em}
        {\Large \@author \par}
        \vspace*{1em}
        {\Large \@institute}
    \end{minipage}%
    %   \tikz[remember picture,overlay]\node[scale=0.8,anchor=east,xshift=0.45\linewidth,yshift=3cm,inner sep=0pt] {%
    %    %\@titlegraphic
    %    \includegraphics[width=8cm]{figures/LMU}
    % };
    %
}
\makeatother

<<setup-child, include = FALSE, echo=FALSE>>=
library(mlr)
library(ggplot2)
library(gridExtra)
library(data.table)
library(ggExtra)
library(knitr)

options(digits = 3, width = 65, str = strOptions(strict.width = "cut", vec.len = 3))

opts_chunk$set(
  echo        = FALSE,
  prompt      = FALSE,
  keep.source = TRUE,
  strip.white = TRUE,
  cache       = TRUE,
  tidy        = FALSE,
  concordance = TRUE,
  message     = FALSE,
  warning     = FALSE,

  size        = 'scriptsize',

  fig.height  = 5.8,
  fig.width   = 8,
  fig.pos     = "h!",
  small.mar   = TRUE,
  eps         = FALSE,
  crop        = TRUE,
  fig.align   = "center",
  out.width   = "0.4\\textwidth"
  # fig.path    = "knit-figure/prada1-"
)

theme_update(axis.line = element_line(colour = "black"),
  panel.grid.major = element_line(colour = "grey80"),
  panel.grid.minor = element_line(colour = "grey80"),
  panel.border = element_blank(),
  panel.background = element_rect(fill = "transparent"),
  plot.background = element_rect(fill = "transparent"))
@

\makeatletter
\renewcommand\section{%
   \@startsection
     {section}% name
     {1}% level
     {\z@}% indent
     {0ex \@plus -1ex \@minus -.2ex}% beforeskip, changed!
     {2.3ex \@plus.2ex}% afterskip
     {\normalfont\Large\bfseries}% style
}
\makeatother
\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

\begin{columns}

\column{0.5}
\block{Introduction}{
Let's predict how much rent you should pay for your flat, based on some data. [footnote]
We got some data on it, contains rent in Euros, the size of the living area, the location (inner city, outer ring, industry, party area).
We use machine learning methods to let the computer learn by itself to prediction. 
Machine learning is a set of techniques that lets the computer learn from data to make predictions. 
The good thing is, that we don't have to find out the exact relationship between the rent and the attributes of the flat, but the computer will. 
The bad thing is that the computer will not easily tell us the relationships it found out.

Except when you use methods from interpretable machine learning, which are presented in this poster. 
FOOTNOTE: the example is completely made up, don't use for actual life decisions.
}

\block{Background}{
\section*{Feature effect}

\section*{Feature importance}
}

\column{0.5}

\block{Method}{
Model-agnostic global permutation feature importance (PFI):
}

\block{References}{
\begingroup
  \renewcommand{\section}[2]{}%
  \small
  \bibliographystyle{plain}
  \bibliography{Bib}
\endgroup
}
\end{columns}
\end{document}

### 2018-09-28 poster/poster_2018.Rnw
structures poster

@@
 \title{Interpretable Machine Learning}
-\author{Christoph Molnar, Prof. Bernd Bischl}
-\institute{Department of Statistics, Ludwig Maximilians University\\[0.5cm] \url{christoph.molnar@stat.uni-muenchen.de}}
+\author{Christoph Molnar \\ }
+\institute{
+LMU Munich / Department of Statistics/ \url{christoph.molnar@stat.uni-muenchen.de} \\
+Prof. Dr. Bernd Bischl}
 %\titlegraphic{\hspace{-3200px}\includegraphics[scale=0.3]{figures/logo.png}}
@@
 \colorlet{titlebgcolor}{colorOne}
-\makeatletter
-\renewcommand\TP@maketitle{%
-   %\hspace{-200px}\includegraphics[scale=0.35]{figures/logo.png}\hspace{100px}
-     \tikz[remember picture, overlay]\node[scale=0.8,anchor=east,xshift=0.65\linewidth,yshift=4cm,inner sep=0pt] {%
-      \includegraphics[scale=6.5]{figures/logo.pdf} %green 1.3
-   };
-   \hspace{500px}
-   \centering
-   \begin{minipage}[b]{0.7\linewidth}
-        \centering
-        \color{titlefgcolor}
-        {\bfseries \huge \sc \@title \par}
-        \vspace*{1em}
-        {\Large \@author \par}
-        \vspace*{1em}
-        {\Large \@institute}
-    \end{minipage}%
-    %   \tikz[remember picture,overlay]\node[scale=0.8,anchor=east,xshift=0.45\linewidth,yshift=3cm,inner sep=0pt] {%
-    %    %\@titlegraphic
-    %    \includegraphics[width=8cm]{figures/LMU}
-    % };
-    %
-}
-\makeatother
 
@@
 
-\column{0.5}
-\block{Introduction}{
+\column{0.33}
+\block{What did you learn, computer?}{
 Let's predict how much rent you should pay for your flat, based on some data. [footnote]
 We got some data on it, contains rent in Euros, the size of the living area, the location (inner city, outer ring, industry, party area).
+TODO: Create graphic with features and output as icons.
 We use machine learning methods to let the computer learn by itself to prediction. 
@@
 The bad thing is that the computer will not easily tell us the relationships it found out. 
+Now we have the following problem:
+TODO: Insert black box graphic
 
@@
 
-\block{Background}{
-\section*{Feature effect}
+\block{Interpretable Models}{
+One approach: Just use machine learning models that only allow relationships with simple structures, like a decision tree, or a weighted sum of the inputs.
+Advantage: Already interpretable
+Disadvantage: Usually you pay with performance of the model.
 
-\section*{Feature importance}
+\section*{Linear regression models}
+\section*{Decision trees}
+\section*{Decision rules}
+
+All of the above techniques can also be used on a more meta level:
+When you alread have a black box machine learning model, you can approximate it with an interpretable model.
+TODO: Visualizaton.
 }
 
-\column{0.5}
+\column{0.33}
 
-\block{Method}{
-Model-agnostic global permutation feature importance (PFI):
+\block{How do features affect the predictions?}{
+\section*{Individual Conditional Expectation}
+\section*{Partial Dependence Plots}
+\section*{Average Accumulated Effects}
 }
+
+\column{0.33}
+\block{Which features were important?}{
+How important was a feature
+}
+
+\block{Why was this prediction made?}
 

### 2018-10-03 poster/poster_2018.Rnw
adds first content

@@
 
-\definecolorpalette{BlueGrayOrange}{
+\definecolorpalette{GrayOrangeBlue}{
 \definecolor{colorOne}{HTML}{C0C0C0}
@@
   warning     = FALSE,
-
   size        = 'scriptsize',
-
   fig.height  = 5.8,
@@
   fig.align   = "center",
-  out.width   = "0.4\\textwidth"
+  out.width   = "0.3\\textwidth"
   # fig.path    = "knit-figure/prada1-"
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+%\SweaveOpts{concordance=TRUE}
 \maketitle
@@
 Let's predict how much rent you should pay for your flat, based on some data. [footnote]
+This serves as an example for a machine learning task.
+But it could be anything, really, like predicting loan defaults, when some robot will need to be repaired, how good a drug will work and so on.
 We got some data on it, contains rent in Euros, the size of the living area, the location (inner city, outer ring, industry, party area).
@@
 We use machine learning methods to let the computer learn by itself to prediction. 
+For this example I used random forests, which is the majority vote of hundreds of different decision trees.
+
+<<create-data, echo = FALSE>>=
+set.seed(42)
+n = 500
+df = data.frame(location = sample(c("Inner city", "Outer city", "Industrial"), prob = c(0.2, 0.7, 0.1), size = n, replace = TRUE))
+df$cat.allowed = factor(sample(c("Yes", "No"), size = n, replace = TRUE))
+df$living.area = runif(n, min = 20, max = 100) + (df$location != "Inner city") * abs(round(rnorm(n, sd = 20), 1))
+df$rent = (10 * (df$location == "Industrial") + 13 * (df$location == "Outer city") + 15 * (df$location == "Inner city")) * df$living.area + 100 * (df$cat.allowed == "Yes") + 50 * (df$cat.allowed == "Yes" & df$location == "Inner city")
+df$useless.feature = rnorm(n)
+@
+
+<<train-model, echo=FALSE>>=
+library(randomForest)
+library(mlr)
+library(iml)
+
+tsk = makeRegrTask(data = df, target = "rent")
+lrn = makeLearner("regr.svm")
+rf = train(lrn, tsk)
+pred = Predictor$new(rf, data = df, y = "rent")
+# 
+# 
+# eff = FeatureEffect$new(pred, "living.area", method = "ale")
+# 
+# eff = FeatureEffect$new(pred, "cat.allowed", method = "ale")
+# 
+# eff = FeatureEffect$new(pred, "location", method = "ale")
+# 
+# eff = FeatureEffect$new(pred, "location", method = "pdp")
+# 
+# 
+# 
+# interact = Interaction$new(pred)
+# 
+# interact.la = Interaction$new(pred, "living.area")
+@
+
+% \textbf{What is machine learning?}
 Machine learning is a set of techniques that lets the computer learn from data to make predictions.
+
+% \textbf{Machine learning has a problem.}
 The good thing is, that we don't have to find out the exact relationship between the rent and the attributes of the flat, but the computer will. 
@@
 Now we have the following problem:
-TODO: Insert black box graphic
+\includegraphics[width=0.25\textwidth]{figure/black-box.png}
 
@@
 }
+
+\block{Which features were important?}{
+Let's start with a simple question: Which are the most important features to know for making predictions?
+The machine learning model usually won't even tell us that.
+We measure importance as the drop in performance of the model if we would not know the features.
+Not knowing the feature can be simulated by shuffling the feature in the data: We break the assocation between the feature and the target.
+If the performance of the model drops a lot, the features was important.
+<<feature-importance>>=
+imp = FeatureImp$new(pred, "mae")
+plot(imp)
+@
+In our little rent example, the most important feature was the size of the living area.
+
+Caveats: Doesn't work well when the features are highly correlated.
+}
+
+\column{0.33}
+
+\block{How do features affect the predictions?}{
+We want to know what relationships the model learned between the input features and the output.
+In our example, we ask:
+Is the predicted rent higher or lower when cats are allowed?
+How does the relationship between the living area and the predicted rent look like?
+
+Let's start with the most simple technique.
+
+The effect of a feature on the prediction can become complex, depending on the underlying model used.
+But it's rather simple for one data instance (here an appartment).
+We know the prediction for this point, and we can visualize how the prediction changes when we vary the input.
+For example we can vary the size of the living area and observe how the prediction of the model changes. 
+If we do this for all data points and plot this, we get the individual conditional expectation plot.
+When we average those lines, the we get the partial dependence plot:
+
+<<ice>>=
+eff = FeatureEffect$new(pred, "living.area", method = "pdp+ice")
+plot(eff)
+@
+
+Caveat: Has trouble when the inputs are strongly correlated, because ICE and PDP rely on replacing values of the feature with other grid values, and when correlated the new data instances become unrealistic.
+In this case, use Accumulated Local Effects instead.
+
+Accumulated Local Effects work like XXXXXX
+}
+
+\block{Why was this prediction made?}{
+We have a couple of options to explain why a particular prediction was made by a machine learning model.
+
+\begin{itemize}
+\item Create a counterfactual explanation
+\item Fit a local, interpretable model
+\item Use game theory to attribute the prediction to the features
+\end{itemize}
+
+\section*{Counterfactual explanations}
+
+\section*{Local surrogate models}
+
+\section*{Shapley Values}
+
+}
+
+\column{0.33}
 
@@
 \section*{Linear regression models}
+The linear regression model models the outcome as a weighted sum:
+
+\[y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p\]
+
+$x_1$ to $x_p$ are the p features that are used to predict the outcome y. 
+The goal is to find the best possible weights, which can be found for example via the least squares algorithm.
+
+In our rent example, the folowing weights are found:
+<<train-model-lm>>=
+mod.lm = lm(rent ~ ., data = df)
+knitr::kable(coef(mod.lm))
+@
+A positive weight means that the feature increases the prediction, negative decreases the prediction.
+An increase of the living area increases the predicted rent by \Sexpr{coef(mod.lm)["living.area"]}.
+Changing the location from the industrial complex (the reference category) to the inner city increases the predicted rent by \Sexpr{coef(mod.lm)["locationInner city"]}.
+Linear models have a huge amount of extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity. CITE and CITE
+
 \section*{Decision trees}
+Decision trees are like the decision trees we know from other areas, except that they are learned based on data in machine learning.
+Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.
+
+Let's train a decision tree on the rent data:
+<<train-model-tree>>=
+library(partykit)
+mod.tree = ctree(rent ~ ., data = df, control = ctree_control(maxdepth = 3))
+plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
+@
+The living area is selected as the first split feature.
+
 \section*{Decision rules}
+<<train-model-rules, eval = FALSE>>=
+
+library("RWeka")
+library(rJava)
+
+extract.rules.jrip = function (rule) {
+rules = scan(text=.jcall(rule$classifier, "S", "toString"), sep="\n", what="")
+# removes text
+rules = rules[-c(1, 2, length(rules))]
+rules = gsub("\\([0-9]*\\.[0-9]\\/[0-9]*\\.[0-9]\\)", "", rules)
+rules = as.matrix(rules)[-c(1:2, 6), ,drop=FALSE]
+rules  = data.frame(rules)
+if (nrow(rules) == 0) {
+return(NULL)
+} else {
+knitr::kable(rules)
+}
+}
+
+df2 = df
+df2$rent = cut(df$rent, breaks = quantile(df$rent, probs = seq(from = 0, to = 1, length.out = 5)))
+df2$living.area = cut(df$living.area, breaks = quantile(df$living.area, probs = seq(from = 0, to = 1, length.out = 5)))
+df2$useless.feature = cut(df$useless.feature, breaks = quantile(df$useless.feature, probs = seq(from = 0, to = 1, length.out = 5)))
+rule = JRip(rent ~ ., data = df2)
+extract.rules.jrip(rule)
+@
+
+Other approaches: 
+RuleFit CITE, SLIM, \ldots
+
+\section*{Surrogate models}
 
@@
 }
-
-\column{0.33}
-
-\block{How do features affect the predictions?}{
-\section*{Individual Conditional Expectation}
-\section*{Partial Dependence Plots}
-\section*{Average Accumulated Effects}
-}
-
-\column{0.33}
-\block{Which features were important?}{
-How important was a feature
-}
-
-\block{Why was this prediction made?}
 

### 2018-10-19 poster/poster_2018.Rnw
Continues poster

--- 2018-09-28 poster/poster_2018.Rnw
@@
 \usepackage{psfrag}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\usepackage{eqnarray}
+%\usepackage{algorithm}
+%\usepackage{algpseudocode}
+%\usepackage{eqnarray}
 \usepackage{arydshln}
@@
 \usepackage{colortbl}
-\usepackage{mathtools}
 \usepackage{wrapfig}
@@
 }
-\usepackage{mathtools}
 \usepackage{hyperref}
@@
   strip.white = TRUE,
-  cache       = TRUE,
+  cache       = FALSE,
   tidy        = FALSE,
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
 \maketitle
@@
 \column{0.33}
-\block{What did you learn, computer?}{
-Let's predict how much rent you should pay for your flat, based on some data. [footnote]
-We got some data on it, contains rent in Euros, the size of the living area, the location (inner city, outer ring, industry, party area).
-TODO: Create graphic with features and output as icons.
-We use machine learning methods to let the computer learn by itself to prediction. 
-Machine learning is a set of techniques that lets the computer learn from data to make predictions. 
-The good thing is, that we don't have to find out the exact relationship between the rent and the attributes of the flat, but the computer will. 
-The bad thing is that the computer will not easily tell us the relationships it found out. 
-Now we have the following problem:
-TODO: Insert black box graphic
+\block{Machine learning}{
+Machine learning is a set of techniques that lets a computer learn from data to make predictions by finding patterns in the data, in contrast to of telling the computer explicitly how to make predictions.
+Machine learning can be used to predict credit default, risk for cancer, number of sales in the next week and much more.
+The big disadnvantage of the machine learning approach is that the program now is a black box for us.
 
-Except when you use methods from interpretable machine learning, which are presented in this poster. 
-FOOTNOTE: the example is completely made up, don't use for actual life decisions.
+\includegraphics[width=0.25\textwidth]{figure/black-box.png}
+
+This poster presents methods for interpretability that are model-agnostic (i.e.
+can be applied to any black box model), which help to explain the behaviour of machine learning models and individual predictions.
+It also presents models that are considered intrinsically interpretable.
+
 }
 
-\block{Interpretable Models}{
-One approach: Just use machine learning models that only allow relationships with simple structures, like a decision tree, or a weighted sum of the inputs.
-Advantage: Already interpretable
-Disadvantage: Usually you pay with performance of the model.
+\block{Example: Predict Rent}{
+To illustrate the intepretability methods, we use a machine learning model that predicts the rent of an appartment given the size of the living area, the location and if cats are allowed.
+The data is simulated and any other data would have worked as well.
+For this example I used random forests, but other models, like a neural network, a linear regression model or boosting would have been possible as well, because the interpretability methods are all model-agnostic.
 
-\section*{Linear regression models}
-\section*{Decision trees}
-\section*{Decision rules}
+<<create-data, echo = FALSE>>=
+set.seed(42)
+n = 500
+df = data.frame(location = sample(c("Inner city", "Outer city"), prob = c(0.4, 0.6), size = n, replace = TRUE))
+df$cat.allowed = factor(sample(c("Yes", "No"), size = n, replace = TRUE))
+df$size = round(runif(n, min = 20, max = 100) + (df$location != "Inner city") * abs(round(rnorm(n, sd = 20), 1)), 0)
+df$rent = (12 * (df$location == "Outer city") + 17 * (df$location == "Inner city")) * df$size + 100 * (df$cat.allowed == "Yes") + 50 * (df$cat.allowed == "Yes" & df$location == "Inner city")
+@
 
-All of the above techniques can also be used on a more meta level:
-When you alread have a black box machine learning model, you can approximate it with an interpretable model.
-TODO: Visualizaton.
+<<train-model, echo=FALSE>>=
+library(randomForest)
+library(mlr)
+library(iml)
+
+tsk = makeRegrTask(data = df, target = "rent")
+lrn = makeLearner("regr.randomForest")
+rf = train(lrn, tsk)
+pred = Predictor$new(rf, data = df, y = "rent")
+@
+
+}
+
+\block{Why explain?}{
+We need to explain machine learning models to\cite{adadi2018peeking}:
+TODO: COPY FROM SLIDES
+\begin{itemize}
+\item \textbf{Explain to justify}: Machine learning models make errors, can be unfair (e.g racial bias.) and make unexpected predictions. In those cases, we require a model to justify predictions.
+\item \textbf{Explain to control}: Understand more about the systems vulnerabilities and conditions for failure. e.g. know for what type of data it performs worse. 
+\item \textbf{Explain to improve}: Insights about model behaviour helps to improve the model.
+\item \textbf{Explain to discover}: A black box model is optimized to make predictions, not to give insights.
+Interpretability methods enable to extract insights from black box models.
+\end{itemize}
+}
+
+\block{Feature importance}{
+We can measure importance of the inputs as the drop in performance of the model if we would not know the features.
+Not knowing the feature can be simulated by shuffling the feature in the data: We break the assocation between the feature and the target.
+If the performance of the model drops a lot, the features was important.
+<<feature-importance, echo=FALSE>>=
+imp = FeatureImp$new(pred, "mae")
+plot(imp)
+@
+In our little rent example, the most important feature was the size of the living area.
+
+Caveats: Doesn't work well when the features are highly correlated.
 }
@@
 
-\block{How do features affect the predictions?}{
-\section*{Individual Conditional Expectation}
-\section*{Partial Dependence Plots}
-\section*{Average Accumulated Effects}
+\block{Feature effects}{
+How does, on average, a feature input affect the prediction?
+The effect of a feature on the prediction can become complex, depending on the underlying model used.
+
+The simplest way to analyze this for an individual prediction is called Individual Conditional Expectations: 
+For an individual data point, we vary one of the features and observe how the prediction changes.
+If we do this for all data points and plot this, we get the individual conditional expectation plot \cite{goldstein2015peeking}.
+When we average those lines (black), the we get the partial dependence plot (yellow line) \cite{friedman2001greedy}:
+
+We know the prediction for this point, and we can visualize how the prediction changes when we vary the input.
+For example we can vary the size of the living area and observe how the prediction of the model changes.
+
+<<ice, echo=FALSE>>=
+eff = FeatureEffect$new(pred, "size", method = "pdp+ice")
+plot(eff, rug = FALSE)  + ggtitle("Effect of size on predicted rent") +
+  scale_x_continuous("Size of appartment") + 
+  scale_y_continuous("Predicted rent")
+@
+
+Caveat: ICE and PDP have trouble when the inputs are strongly correlated.
+In this case, use Accumulated Local Effects instead \cite{apley2016visualizing}.
+}
+
+\block{Individual prediction explanation}{
+
+<<appartment, echo=FALSE,include=FALSE>>=
+# Choose appartment
+appartment = df[10,]
+pred.rent = round(pred$predict(appartment)$.prediction, 0)
+@
+
+We have a couple of options to explain why a particular prediction was made by a machine learning model.
+Suppose the rent model predicts \Sexpr{pred.rent} Euros rent, for an appartment in the inner city with a living area of 85 $m^2$ and where cats are allowed. 
+How can we explain this prediction?
+
+<<counterfact, echo=FALSE, include = FALSE>>=
+appartment.mod = appartment
+appartment.mod$cat.allowed = factor("No", levels = c("No", "Yes"))
+pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
+@
+\textbf{Counterfactual explanations}: 
+A counterfactual explanation of a prediction describes the smallest change to the feature values (inputs) that changes the prediction to a predefined output.
+For example: How does the input have to change so that the prediction changes below 1200 Euros?
+Answer: If cats were not allowed, the predicted rent would be \Sexpr{pred.new} Euros.
+
+\textbf{Shapley Values}: 
+A prediction can be seen as a cooperative game. 
+The feature values (inputs) are players in a game and cooperate to achieve a game payout, which is the prediction.
+We want to fairly distribute the difference of the prediction to the average prediction among the features.
+Game theory has developed a method called Shapley values that can fairly do this type of distribution.
+For machine learning, it works by repeatedly simulating that only a subset of feature values is known and computing how the prediction would look like.
+
+<<shapley, echo=FALSE>>=
+shap = Shapley$new(pred)
+shap$explain(appartment[,1:3])
+plot(shap)
+@
+
+\textbf{Local surrogate models}: 
+The idea behind local surrogate models is to replace the complex model with an interpretable model (e.g. regression), but only locally. 
+This means we fit an individual interpretable model for the data point for which we want to explain the prediction.
+The closer data points that are to the data point of interest the higher the weight in the model calculation.
+Interpretation depends on the chosen model, see poster box on the top right.
+
+<<lime, echo=FALSE>>=
+# locmod = LocalModel$new(pred, k = 2)
+# locmod$explain(df[1,1:3])
+# plot(locmod)
+@
+
 }
@@
 \column{0.33}
-\block{Which features were important?}{
-How important was a feature
+
+\block{Interpretable Models}{
+Some models are intrinsically interpretable, because they learn simple relationships between input and output like a decision tree or a weighted sum of the inputs.
+The big disadvantage of intrinsically interpretable models is that they often predict the data less well.
+
+\section*{Linear Regression Models}
+The linear regression model models the outcome as a weighted sum:
+
+\[y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p\]
+
+$x_1$ to $x_p$ are the p features that are used to predict the outcome y. 
+The goal is to find the best possible weights, which can be found for example via the least squares algorithm.
+
+In our rent example, the folowing weights are found:
+<<train-model-lm, echo=FALSE, include=FALSE>>=
+mod.lm = lm(rent ~ ., data = df)
+cf = round(coef(mod.lm))
+knitr::kable(cf, format = 'latex')
+@
+
+$$rent = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot size  \Sexpr{cf["locationOuter city"]} \cdot I_{\text{outer city}} +  \Sexpr{cf["cat.allowedYes"]} \cdot I_{\text{cats allowed}}$$
+
+A positive weight means that the feature increases the prediction, negative decreases the prediction.
+An increase of the living area increases the predicted rent by \Sexpr{coef(mod.lm)["size"]}.
+Changing the location from the industrial complex (the reference category) to the inner city increases the predicted rent by \Sexpr{coef(mod.lm)["locationInner city"]}.
+Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.
+
+\section*{Decision Trees}
+Decision trees are like the decision trees we know from other areas, except that they are learned based on data in machine learning.
+Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.
+
+Let's train a decision tree on the rent data:
+<<train-model-tree, echo=FALSE>>=
+library(partykit)
+mod.tree = ctree(rent ~ ., data = df, control = ctree_control(maxdepth = 2))
+plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
+@
+The living area is selected as the first split feature.
+
+\section*{Decision rules}
+A decision rule is a simple IF-THEN statement consisting of a condition and a prediction.
+For example: IF it rains today AND if it’s April (condition), THEN it will rain tomorrow (prediction).
+A single decision rule or a combination of several rules can be used to make predictions.
+
+<<train-model-rules, echo=FALSE, includee=FALSE, eval=FALSE>>=
+
+library("RWeka")
+library(rJava)
+
+extract.rules.jrip = function (rule) {
+rules = scan(text=.jcall(rule$classifier, "S", "toString"), sep="\n", what="")
+# removes text
+rules = rules[-c(1, 2, length(rules))]
+rules = gsub("\\([0-9]*\\.[0-9]\\/[0-9]*\\.[0-9]\\)", "", rules)
+rules = as.matrix(rules)[-c(1:2, 6), ,drop=FALSE]
+rules  = data.frame(rules)
+if (nrow(rules) == 0) {
+return(NULL)
+} else {
+knitr::kable(rules, format = 'latex')
+}
 }
 
-\block{Why was this prediction made?}
+df2 = df
+df2$rent = cut(df$rent, breaks = quantile(df$rent, probs = seq(from = 0, to = 1, length.out = 20)))
+df2$size = cut(df$size, breaks = quantile(df$size, probs = seq(from = 0, to = 1, length.out = 5)))
+rule = JRip(rent ~ ., data = df2)
+extract.rules.jrip(rule)
+@
+A rule could be:
+
+IF living area $\in 90 \text{ and } 110 m^2$  AND location $=$ inner city THEN the rent is between 1540 and 1890 EUR
+
+Other approaches: 
+RuleFit CITE, SLIM, \ldots
+
+\section*{Surrogate Models}
+Often these intrinsically interpretable models are much worse than more flexible models, but they can still be used to increase interpretabilit by approximating a complex model with an interpretable model.
+Predict the predictions of the complex model with an intrinsically interpretable model and interpret the latter.
+For example, interpret a neural network by approximating its predictions with a decision tree.
+
+}
 

### 2018-10-19 poster/poster_2018.Rnw
Improves poster

@@
 
-\includegraphics[width=0.25\textwidth]{figure/black-box.png}
+\includegraphics[width=0.25\textwidth]{figure/teaser.png}
 
@@
 \block{Why explain?}{
-We need to explain machine learning models to\cite{adadi2018peeking}:
-TODO: COPY FROM SLIDES
+We need to explain machine learning models to \cite{adadi2018peeking}:
 \begin{itemize}
@@
 If the performance of the model drops a lot, the features was important.
-<<feature-importance, echo=FALSE>>=
+<<feature-importance, echo=FALSE, fig.width=12, fig.height=3>>=
 imp = FeatureImp$new(pred, "mae")
-plot(imp)
+
+impres = imp$results
+
+ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
+  theme(aspect.ratio=1/4, text = element_text(size=20),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 @
 In our little rent example, the most important feature was the size of the living area. 
-
-Caveats: Doesn't work well when the features are highly correlated.
 }
@@
 
-<<ice, echo=FALSE>>=
+<<ice, echo=FALSE, fig.width=12, fig.height=3>>=
 eff = FeatureEffect$new(pred, "size", method = "pdp+ice")
@@
   scale_x_continuous("Size of appartment") + 
-  scale_y_continuous("Predicted rent")
+  scale_y_continuous("Predicted rent") +
+  theme(aspect.ratio=1/4, text = element_text(size=20),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+
 @
@@
 # Choose appartment
-appartment = df[10,]
+appartment = df[3,]
 pred.rent = round(pred$predict(appartment)$.prediction, 0)
@@
 We have a couple of options to explain why a particular prediction was made by a machine learning model.
-Suppose the rent model predicts \Sexpr{pred.rent} Euros rent, for an appartment in the inner city with a living area of 85 $m^2$ and where cats are allowed. 
+Suppose the rent model predicts \Sexpr{pred.rent} Euros rent, for an appartment in the outer city with a living area of 92 $m^2$ and where cats are not allowed. 
 How can we explain this prediction?
@@
 appartment.mod = appartment
-appartment.mod$cat.allowed = factor("No", levels = c("No", "Yes"))
+appartment.mod$cat.allowed = factor("Yes", levels = c("No", "Yes"))
 pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
@@
 A counterfactual explanation of a prediction describes the smallest change to the feature values (inputs) that changes the prediction to a predefined output.
-For example: How does the input have to change so that the prediction changes below 1200 Euros?
-Answer: If cats were not allowed, the predicted rent would be \Sexpr{pred.new} Euros.
+For example: How does the input have to change so that the prediction changes above 1100 Euros?
+Answer (one of many possible ones): If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.
 
@@
 
-<<shapley, echo=FALSE>>=
+<<shapley, echo=FALSE, fig.width=12, fig.height=3>>=
 shap = Shapley$new(pred)
 shap$explain(appartment[,1:3])
-plot(shap)
+plot(shap) +
+  theme(aspect.ratio=1/4, text = element_text(size=20),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  scale_y_continuous("Contribution to prediction") +
+  scale_x_discrete("")
+
 @
@@
 \block{Interpretable Models}{
-Some models are intrinsically interpretable, because they learn simple relationships between input and output like a decision tree or a weighted sum of the inputs.
-The big disadvantage of intrinsically interpretable models is that they often predict the data less well.
+Some models are intrinsically interpretable, because they learn simple relationships between input and output.
+Their big disadvantage is that they often predict the data less well.
 
@@
 
-$x_1$ to $x_p$ are the p features that are used to predict the outcome y. 
-The goal is to find the best possible weights, which can be found for example via the least squares algorithm.
-
-In our rent example, the folowing weights are found:
+$x_1$ to $x_p$ are the features that are used to predict the outcome y. 
+The goal is to find the best possible weights $\beta$.
+In our rent example, the best weights are:
 <<train-model-lm, echo=FALSE, include=FALSE>>=
@@
 
-$$rent = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot size  \Sexpr{cf["locationOuter city"]} \cdot I_{\text{outer city}} +  \Sexpr{cf["cat.allowedYes"]} \cdot I_{\text{cats allowed}}$$
+$$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  \Sexpr{cf["locationOuter city"]} \cdot I_{\text{outer city}} +  \Sexpr{cf["cat.allowedYes"]} \cdot I_{\text{cats allowed}}$$
 
@@
 
-\section*{Decision rules}
+\section*{Decision Rules}
 A decision rule is a simple IF-THEN statement consisting of a condition and a prediction.
@@
 
-Other approaches: 
-RuleFit CITE, SLIM, \ldots
-
 \section*{Surrogate Models}

### 2018-10-21 poster/poster_2018.Rnw
Update poster_2018.Rnw

@@
 \usepackage{hyperref}
-\usetheme{Desert}
-\usecolorstyle[colorPalette=BlueGrayOrange]{Russia}
+%\usetheme{Desert}
+%\usecolorstyle[colorPalette=BlueGrayOrange]{Russia}
 \usebackgroundstyle{Rays}
@@
 \block{Machine learning}{
-Machine learning is a set of techniques that lets a computer learn from data to make predictions by finding patterns in the data, in contrast to of telling the computer explicitly how to make predictions.
-Machine learning can be used to predict credit default, risk for cancer, number of sales in the next week and much more.
-The big disadnvantage of the machine learning approach is that the program now is a black box for us.
+Machine learning is a set of methods that let a computer learn from data to make predictions by finding patterns.
+Machine learning can be used to predict credit default, risk for cancer, weekly sales ...
+}
+\block{Black Box Problem}{
+The big disadvantage of the machine learning approach is that the  learned program is intransparent to us.
 
@@
 
-This poster presents methods for interpretability that are model-agnostic (i.e.
-can be applied to any black box model), which help to explain the behaviour of machine learning models and individual predictions.
-It also presents models that are considered intrinsically interpretable.
+Interpretable machine learning is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.
+}
 
+\block{Why explain?}{
+We need to explain machine learning models to \cite{adadi2018peeking}:
+\begin{itemize}
+\item \textbf{Explain to justify}: Machine learning models make errors, can be unfair (e.g racial bias.) and make unexpected predictions. In those cases, we require a model to justify predictions.
+\item \textbf{Explain to control}: Understand more about the systems vulnerabilities and conditions for failure. e.g. know for what type of data it performs worse. 
+\item \textbf{Explain to improve}: Insights about model behaviour helps to improve the model.
+\item \textbf{Explain to discover}: A black box model is optimized to make predictions, not to give insights.
+Interpretability methods enable to extract insights from black box models.
+\end{itemize}
 }
@@
 \block{Example: Predict Rent}{
-To illustrate the intepretability methods, we use a machine learning model that predicts the rent of an appartment given the size of the living area, the location and if cats are allowed.
-The data is simulated and any other data would have worked as well.
-For this example I used random forests, but other models, like a neural network, a linear regression model or boosting would have been possible as well, because the interpretability methods are all model-agnostic.
+As example we use a machine learning model that predicts the rent of an appartment given the size of its living area, whether the location is good or bad and if cats are allowed.
 
@@
 n = 500
-df = data.frame(location = sample(c("Inner city", "Outer city"), prob = c(0.4, 0.6), size = n, replace = TRUE))
-df$cat.allowed = factor(sample(c("Yes", "No"), size = n, replace = TRUE))
-df$size = round(runif(n, min = 20, max = 100) + (df$location != "Inner city") * abs(round(rnorm(n, sd = 20), 1)), 0)
-df$rent = (12 * (df$location == "Outer city") + 17 * (df$location == "Inner city")) * df$size + 100 * (df$cat.allowed == "Yes") + 50 * (df$cat.allowed == "Yes" & df$location == "Inner city")
+df = data.frame(location = sample(c("Good", "Bad"), prob = c(0.4, 0.6), size = n, replace = TRUE))
+df$cats = factor(sample(c("Yes", "No"), size = n, replace = TRUE))
+df$size = round(runif(n, min = 20, max = 140) )
+df$rent = (12 * (df$location == "Bad") + 17 * (df$location == "Good")) * df$size + 100 * (df$cats == "Yes") + 50 * (df$cats == "Yes" & df$location == "Good") + rnorm(n, sd = 10)
+@
+
+<<rent>>=
+knitr::include_graphics("figure/rent.png")
 @
@@
 
-\block{Why explain?}{
-We need to explain machine learning models to \cite{adadi2018peeking}:
-\begin{itemize}
-\item \textbf{Explain to justify}: Machine learning models make errors, can be unfair (e.g racial bias.) and make unexpected predictions. In those cases, we require a model to justify predictions.
-\item \textbf{Explain to control}: Understand more about the systems vulnerabilities and conditions for failure. e.g. know for what type of data it performs worse. 
-\item \textbf{Explain to improve}: Insights about model behaviour helps to improve the model.
-\item \textbf{Explain to discover}: A black box model is optimized to make predictions, not to give insights.
-Interpretability methods enable to extract insights from black box models.
-\end{itemize}
-}
-
 \block{Feature importance}{
-We can measure importance of the inputs as the drop in performance of the model if we would not know the features.
-Not knowing the feature can be simulated by shuffling the feature in the data: We break the assocation between the feature and the target.
-If the performance of the model drops a lot, the features was important.
-<<feature-importance, echo=FALSE, fig.width=12, fig.height=3>>=
+The feature importance tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
+The bigger the drop, the more important.
+<<feature-importance, echo=FALSE, fig.width=12, fig.height=8>>=
 imp = FeatureImp$new(pred, "mae")
@@
 ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
-  theme(aspect.ratio=1/4, text = element_text(size=20),
+  theme(aspect.ratio=2/4, text = element_text(size=30),
     panel.grid.major = element_blank(), panel.grid.minor = element_blank())
 @
-In our little rent example, the most important feature was the size of the living area. 
+In the rent example the most important feature was the size of the living area. 
 }
@@
 \block{Feature effects}{
-How does, on average, a feature input affect the prediction?
-The effect of a feature on the prediction can become complex, depending on the underlying model used.
+The feature effects describe how changing one of the features (on average) changes the prediction.
+The effects can be visualized for individual data points (called Individual Conditional Expectations \cite{goldstein2015peeking}) or as an averaged curve (called Partial Dependence Plots \cite{friedman2001greedy}).
 
-The simplest way to analyze this for an individual prediction is called Individual Conditional Expectations: 
-For an individual data point, we vary one of the features and observe how the prediction changes.
-If we do this for all data points and plot this, we get the individual conditional expectation plot \cite{goldstein2015peeking}.
-When we average those lines (black), the we get the partial dependence plot (yellow line) \cite{friedman2001greedy}:
-
-We know the prediction for this point, and we can visualize how the prediction changes when we vary the input.
+<<ice, echo=FALSE, fig.width=12, fig.height=6>>=
+eff = FeatureEffect$new(pred, "size", method = "ale")
+eff.res = eff$results
+ggplot(eff.res)  + geom_line(aes(x=size, y = .ale), size = 3) +
+  scale_x_continuous("Feature Effect") + 
+  scale_y_continuous("Predicted rent") +
+  theme(aspect.ratio=2/4, text = element_text(size=35),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+@
 For example we can vary the size of the living area and observe how the prediction of the model changes.
 
-<<ice, echo=FALSE, fig.width=12, fig.height=3>>=
-eff = FeatureEffect$new(pred, "size", method = "pdp+ice")
-plot(eff, rug = FALSE)  + ggtitle("Effect of size on predicted rent") +
-  scale_x_continuous("Size of appartment") + 
-  scale_y_continuous("Predicted rent") +
-  theme(aspect.ratio=1/4, text = element_text(size=20),
-    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-
-@
-
-Caveat: ICE and PDP have trouble when the inputs are strongly correlated.
-In this case, use Accumulated Local Effects instead \cite{apley2016visualizing}.
+%Caveat: ICE and PDP have trouble when the inputs are strongly correlated.
+%In this case, use Accumulated Local Effects instead \cite{apley2016visualizing}.
 }
 
-\block{Individual prediction explanation}{
+\block{Individual predictions}{
 
@@
 # Choose appartment
-appartment = df[3,]
+appartment = data.frame(location = factor("Good", levels = c("Bad", "Good")),
+    cats = factor("Yes", levels = c("No", "Yes")),
+  size = 50)
 pred.rent = round(pred$predict(appartment)$.prediction, 0)
@@
 We have a couple of options to explain why a particular prediction was made by a machine learning model.
-Suppose the rent model predicts \Sexpr{pred.rent} Euros rent, for an appartment in the outer city with a living area of 92 $m^2$ and where cats are not allowed. 
-How can we explain this prediction?
+Suppose we want to explain the prediction of the following appartment. 
+<<rent2>>=
+knitr::include_graphics("figure/rent.png")
+@
 
@@
 appartment.mod = appartment
-appartment.mod$cat.allowed = factor("Yes", levels = c("No", "Yes"))
+appartment.mod$cats = factor("No", levels = c("No", "Yes"))
 pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
 @
-\textbf{Counterfactual explanations}: 
+\section*{Counterfactual explanations}
 A counterfactual explanation of a prediction describes the smallest change to the feature values (inputs) that changes the prediction to a predefined output.
-For example: How does the input have to change so that the prediction changes above 1100 Euros?
+For example: How does the inputs have to change so that the prediction changes below 900 Euros?
 Answer (one of many possible ones): If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.
 
-\textbf{Shapley Values}: 
-A prediction can be seen as a cooperative game. 
-The feature values (inputs) are players in a game and cooperate to achieve a game payout, which is the prediction.
-We want to fairly distribute the difference of the prediction to the average prediction among the features.
-Game theory has developed a method called Shapley values that can fairly do this type of distribution.
-For machine learning, it works by repeatedly simulating that only a subset of feature values is known and computing how the prediction would look like.
+\section*{Shapley Values}
+From a game theoretical perspective, the feature values are like players in a game and cooperate to achieve a game payout, which is the prediction.
+A method called Shapley Values fairly attributes the difference between the prediction and the average prediction among the feature values to explain a prediction.
 
-<<shapley, echo=FALSE, fig.width=12, fig.height=3>>=
+<<shapley, echo=FALSE, fig.width=12, fig.height=6>>=
 shap = Shapley$new(pred)
-shap$explain(appartment[,1:3])
+shap$explain(appartment)
 plot(shap) +
-  theme(aspect.ratio=1/4, text = element_text(size=20),
+  theme(aspect.ratio=2/4, text = element_text(size=30),
     panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
@@
 
-\textbf{Local surrogate models}: 
+\section*{Local surrogate models}
 The idea behind local surrogate models is to replace the complex model with an interpretable model (e.g. regression), but only locally. 
@@
 \block{Interpretable Models}{
-Some models are intrinsically interpretable, because they learn simple relationships between input and output.
-Their big disadvantage is that they often predict the data less well.
+Intrinsically interpretable models learn simple relationships between input and output,
+but they often predict the data less well.
 
 \section*{Linear Regression Models}
-The linear regression model models the outcome as a weighted sum:
+The linear regression model models the outcome y as a weighted sum of the input features.
 
-\[y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p\]
+\[y = \beta_0 + \beta_1 \cdot \text{Feature}_1 + \ldots + \beta_p \cdot \text{Feature}_p\]
 
-$x_1$ to $x_p$ are the features that are used to predict the outcome y. 
 The goal is to find the best possible weights $\beta$.
-In our rent example, the best weights are:
+In the rent example, the following model is learned.
 <<train-model-lm, echo=FALSE, include=FALSE>>=
@@
 
-$$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  \Sexpr{cf["locationOuter city"]} \cdot I_{\text{outer city}} +  \Sexpr{cf["cat.allowedYes"]} \cdot I_{\text{cats allowed}}$$
+$$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  + \Sexpr{cf["locationGood"]} \cdot I_{\text{good location}} +  \Sexpr{cf["catsYes"]} \cdot I_{\text{cats allowed}}$$
 
-A positive weight means that the feature increases the prediction, negative decreases the prediction.
 An increase of the living area increases the predicted rent by \Sexpr{coef(mod.lm)["size"]}.
-Changing the location from the industrial complex (the reference category) to the inner city increases the predicted rent by \Sexpr{coef(mod.lm)["locationInner city"]}.
-Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.
+Changing the location from "Bad" to the "Good" increases the predicted rent by \Sexpr{coef(mod.lm)["locationGood"]}.
+%Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.
 
 \section*{Decision Trees}
-Decision trees are like the decision trees we know from other areas, except that they are learned based on data in machine learning.
 Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features. 
+The prediction is based on the leaf node we end up in.
 
-Let's train a decision tree on the rent data:
+A decision tree trained on the rent data:
 <<train-model-tree, echo=FALSE>>=
@@
 @
-The living area is selected as the first split feature.
+For example, the tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.
 
@@
 A decision rule is a simple IF-THEN statement consisting of a condition and a prediction.
-For example: IF it rains today AND if it’s April (condition), THEN it will rain tomorrow (prediction).
-A single decision rule or a combination of several rules can be used to make predictions.
+One or many rules can be used to make predictions.
 
@@
 @
-A rule could be:
 
-IF living area $\in 90 \text{ and } 110 m^2$  AND location $=$ inner city THEN the rent is between 1540 and 1890 EUR
+One decision rule learned for the rent prediction is:
+
+IF living area $\in 90 \text{ and } 110 m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR
 
 \section*{Surrogate Models}
-Often these intrinsically interpretable models are much worse than more flexible models, but they can still be used to increase interpretabilit by approximating a complex model with an interpretable model.
-Predict the predictions of the complex model with an intrinsically interpretable model and interpret the latter.
-For example, interpret a neural network by approximating its predictions with a decision tree.
-
+Intrinsically interpretable models can also be used as to increase interpretability of a black box model by approximating its predictions, acting as a surrogate model which can be interpreted. 
+Example: Neural network as black box model, decision tree as surrogate model.
 }

### 2018-10-22 poster/poster_2018.Rnw
continues simplification

@@
   fig.align   = "center",
-  out.width   = "0.3\\textwidth"
+  out.width   = "0.27\\textwidth"
   # fig.path    = "knit-figure/prada1-"
@@
 Machine learning is a set of methods that let a computer learn from data to make predictions by finding patterns.
-Machine learning can be used to predict credit default, risk for cancer, weekly sales ...
+Can be used to predict credit default, cancer, weekly sales ...
+<<ml>>=
+knitr::include_graphics("figure/ml.png")
+@
 }
@@
 
-Interpretable machine learning is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.
+Interpretable machine learning is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.\cite{molnar}
 }
@@
 
-\block{Feature importance}{
-The feature importance tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
-The bigger the drop, the more important.
-<<feature-importance, echo=FALSE, fig.width=12, fig.height=8>>=
-imp = FeatureImp$new(pred, "mae")
-
-impres = imp$results
-
-ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
-  theme(aspect.ratio=2/4, text = element_text(size=30),
-    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
-@
-In the rent example the most important feature was the size of the living area. 
-}
-
 \column{0.33}
@@
 }
+\block{Feature importance}{
+The feature importance tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
+The bigger the drop, the more important.
+<<feature-importance, echo=FALSE, fig.width=12, fig.height=8>>=
+imp = FeatureImp$new(pred, "mae")
 
+impres = imp$results
+
+ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
+  theme(aspect.ratio=2/4, text = element_text(size=30),
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+@
+In the rent example the most important feature was the size of the living area. 
+}
 \block{Individual predictions}{
@@
 From a game theoretical perspective, the feature values are like players in a game and cooperate to achieve a game payout, which is the prediction.
-A method called Shapley Values fairly attributes the difference between the prediction and the average prediction among the feature values to explain a prediction.
+A method called Shapley Values TODO:CITE fairly attributes the difference between the prediction and the average prediction among the feature values to explain a prediction.
 
@@
 
-@
-
-\section*{Local surrogate models}
-The idea behind local surrogate models is to replace the complex model with an interpretable model (e.g. regression), but only locally. 
-This means we fit an individual interpretable model for the data point for which we want to explain the prediction.
-The closer data points that are to the data point of interest the higher the weight in the model calculation.
-Interpretation depends on the chosen model, see poster box on the top right.
-
-<<lime, echo=FALSE>>=
-# locmod = LocalModel$new(pred, k = 2)
-# locmod$explain(df[1,1:3])
-# plot(locmod)
 @
@@
 A decision tree trained on the rent data:
-<<train-model-tree, echo=FALSE>>=
+<<train-model-tree, echo=FALSE, out.width="0.28\\textwidth">>=
 library(partykit)
 mod.tree = ctree(rent ~ ., data = df, control = ctree_control(maxdepth = 2))
-plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
+#plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
+knitr::include_graphics("figure/tree.png")
 @
@@
 Example: Neural network as black box model, decision tree as surrogate model.
+
+\section*{Local surrogate models}
+The idea behind local surrogate models is to replace the complex model with an individual, locally weighted interpretable model (e.g.
+regression) for the data point for which we want to explain the prediction.
+
+<<lime, echo=FALSE>>=
+# locmod = LocalModel$new(pred, k = 2)
+# locmod$explain(df[1,1:3])
+# plot(locmod)
+knitr::include_graphics("figure/local-surrogate.png")
+@
 }
@@
 }
+
 \end{columns}

### 2018-10-23 poster/poster_2018.Rnw
adds images and references

@@
 \block{Machine learning}{
-Machine learning is a set of methods that let a computer learn from data to make predictions by finding patterns.
-Can be used to predict credit default, cancer, weekly sales ...
+Machine learning is a set of methods that let a computer learn from data to make predictions (e.g. cancer, weekly sales, credit default) by finding patterns.
 <<ml>>=
-knitr::include_graphics("figure/ml.png")
+knitr::include_graphics("figure/programing-ml.png")
 @
@@
 \block{Black Box Problem}{
-The big disadvantage of the machine learning approach is that the  learned program is intransparent to us.
+<<iml>>=
+knitr::include_graphics("figure/iml.png")
+@
 
-\includegraphics[width=0.25\textwidth]{figure/teaser.png}
-
-Interpretable machine learning is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.\cite{molnar}
+Solution: \textbf{Interpretable machine learning} is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.\cite{molnar}
 }
@@
 We need to explain machine learning models to \cite{adadi2018peeking}:
-\begin{itemize}
-\item \textbf{Explain to justify}: Machine learning models make errors, can be unfair (e.g racial bias.) and make unexpected predictions. In those cases, we require a model to justify predictions.
-\item \textbf{Explain to control}: Understand more about the systems vulnerabilities and conditions for failure. e.g. know for what type of data it performs worse. 
-\item \textbf{Explain to improve}: Insights about model behaviour helps to improve the model.
-\item \textbf{Explain to discover}: A black box model is optimized to make predictions, not to give insights.
-Interpretability methods enable to extract insights from black box models.
-\end{itemize}
+% \begin{itemize}
+% \item \textbf{Explain to justify}: Machine learning models make errors, can be unfair (e.g racial bias.) and make unexpected predictions. In those cases, we require a model to justify predictions.
+% \item \textbf{Explain to control}: Understand more about the systems vulnerabilities and conditions for failure. e.g. know for what type of data it performs worse. 
+% \item \textbf{Explain to improve}: Insights about model behaviour helps to improve the model.
+% \item \textbf{Explain to discover}: A black box model is optimized to make predictions, not to give insights.
+% Interpretability methods enable to extract insights from black box models.
+% \end{itemize}
+<<explain-to>>=
+knitr::include_graphics("figure/explain.png")
+@
 }
@@
 \block{Example: Predict Rent}{
-As example we use a machine learning model that predicts the rent of an appartment given the size of its living area, whether the location is good or bad and if cats are allowed.
+As example we train a machine learning model (here a random forest) that predicts the rent of an appartment given the size of its living area, the location ("Good or "Bad") and if cats are allowed.
 
@@
 \block{Feature effects}{
-The feature effects describe how changing one of the features (on average) changes the prediction.
-The effects can be visualized for individual data points (called Individual Conditional Expectations \cite{goldstein2015peeking}) or as an averaged curve (called Partial Dependence Plots \cite{friedman2001greedy}).
-
+The feature effect describe how changing a features changes the prediction of all data points on average. \cite{friedman2001greedy}
 <<ice, echo=FALSE, fig.width=12, fig.height=6>>=
-eff = FeatureEffect$new(pred, "size", method = "ale")
+eff = FeatureEffect$new(pred, "size", method = "pdp")
 eff.res = eff$results
-ggplot(eff.res)  + geom_line(aes(x=size, y = .ale), size = 3) +
-  scale_x_continuous("Feature Effect") + 
+ggplot(eff.res)  + geom_line(aes(x=size, y = .y.hat), size = 3) +
+  scale_x_continuous("Size of living area") + 
   scale_y_continuous("Predicted rent") +
@@
 \block{Feature importance}{
-The feature importance tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
+The feature importance \cite{Fisher2018} tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
 The bigger the drop, the more important.
@@
   theme(aspect.ratio=2/4, text = element_text(size=30),
-    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
+    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
+  scale_x_discrete("") +
+  scale_y_continuous("Feature Importance")
 @
@@
 @
+\vspace{1cm}
 \section*{Counterfactual explanations}
@@
 \section*{Shapley Values}
-From a game theoretical perspective, the feature values are like players in a game and cooperate to achieve a game payout, which is the prediction.
-A method called Shapley Values TODO:CITE fairly attributes the difference between the prediction and the average prediction among the feature values to explain a prediction.
+From a game theoretical perspective, the feature values can be seen as players in a game that cooperate to achieve the prediction as a payout.
+The Shapley Values method \cite{vstrumbelj2014explaining} fairly attributes the difference between the prediction and the average prediction among the feature values.
 
@@
     panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
-  scale_y_continuous("Contribution to prediction") +
+  scale_y_continuous("Contribution to prediction minus avg. prediction") +
   scale_x_discrete("")
@@
 Intrinsically interpretable models learn simple relationships between input and output,
-but they often predict the data less well.
+but often predict the data less well than flexible, yet less interpretable, models.
 
 \section*{Linear Regression Models}
-The linear regression model models the outcome y as a weighted sum of the input features.
+%The linear regression model models the outcome y as a weighted sum of the input features.
 
-\[y = \beta_0 + \beta_1 \cdot \text{Feature}_1 + \ldots + \beta_p \cdot \text{Feature}_p\]
+%\[y = \beta_0 + \beta_1 \cdot \text{Feature}_1 + \ldots + \beta_p \cdot \text{Feature}_p\]
 
-The goal is to find the best possible weights $\beta$.
+%The goal is to find the best possible weights $\beta$.
+<<lm>>=
+knitr::include_graphics("figure/lm.png")
+@
+
 In the rent example, the following model is learned.
@@
 Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.
-The prediction is based on the leaf node we end up in.
 
-A decision tree trained on the rent data:
 <<train-model-tree, echo=FALSE, out.width="0.28\\textwidth">>=
@@
 
-\section*{Surrogate Models}
+\section*{Global Surrogate Models}
 Intrinsically interpretable models can also be used as to increase interpretability of a black box model by approximating its predictions, acting as a surrogate model which can be interpreted. 
@@
 
-\section*{Local surrogate models}
-The idea behind local surrogate models is to replace the complex model with an individual, locally weighted interpretable model (e.g.
-regression) for the data point for which we want to explain the prediction.
+\section*{Local Surrogate Models}
+The idea behind local surrogate models \cite{ribeiro2016should} is to replace the complex model with an individual, locally weighted interpretable model (e.g.
+decision tree) for the data point for which we want to explain the prediction.
 

### 2018-10-23 poster/poster_2018.Rnw
adds more images

@@
 <<rent>>=
-knitr::include_graphics("figure/rent.png")
+knitr::include_graphics("figure/rent-ml.png")
 @
@@
 @
-For example we can vary the size of the living area and observe how the prediction of the model changes.
+\textit{For example we can vary the size of the living area and observe how the prediction of the model changes.}
 
@@
 @
-In the rent example the most important feature was the size of the living area. 
+\textit{In the rent example the most important feature was the size of the living area. }
 }
@@
 Suppose we want to explain the prediction of the following appartment. 
-<<rent2>>=
+<<rent2, out.width="0.2\\textwidth">>=
 knitr::include_graphics("figure/rent.png")
@@
 @
-\vspace{1cm}
-\section*{Counterfactual explanations}
-A counterfactual explanation of a prediction describes the smallest change to the feature values (inputs) that changes the prediction to a predefined output.
+\\ \\
+\textbf{Counterfactual explanations} \\
+A counterfactual explanation of a prediction describes the smallest changes to the feature values (inputs) that change the prediction to a predefined output.
 For example: How does the inputs have to change so that the prediction changes below 900 Euros?
-Answer (one of many possible ones): If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.
-
-\section*{Shapley Values}
+Answer (one of many possible ones): \\
+\textit{If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.}
+\\ \\
+\textbf{Shapley Values} \\
 From a game theoretical perspective, the feature values can be seen as players in a game that cooperate to achieve the prediction as a payout.
@@
     panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
-  scale_y_continuous("Contribution to prediction minus avg. prediction") +
+  scale_y_continuous("Contribution to difference of \nprediction minus avg. prediction") +
   scale_x_discrete("")
@@
 but often predict the data less well than flexible, yet less interpretable, models.
-
-\section*{Linear Regression Models}
+\\ \\
+\textbf{Linear Regression Models}\\
 %The linear regression model models the outcome y as a weighted sum of the input features.
@@
 @
-
-In the rent example, the following model is learned.
+\textit{In the rent example, the following weights were learned:}
 <<train-model-lm, echo=FALSE, include=FALSE>>=
@@
 
-$$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  + \Sexpr{cf["locationGood"]} \cdot I_{\text{good location}} +  \Sexpr{cf["catsYes"]} \cdot I_{\text{cats allowed}}$$
-
-An increase of the living area increases the predicted rent by \Sexpr{coef(mod.lm)["size"]}.
-Changing the location from "Bad" to the "Good" increases the predicted rent by \Sexpr{coef(mod.lm)["locationGood"]}.
+$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  + \Sexpr{cf["locationGood"]} \cdot I_{\text{good location}} +  \Sexpr{cf["catsYes"]} \cdot I_{\text{cats allowed}}$
+\textit{The predicted rent increases by \Sexpr{cf["size"]} Euros per each $m^2$.}
+%Changing the location from "Bad" to the "Good" increases the predicted rent by \Sexpr{coef(mod.lm)["locationGood"]}.
 %Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.
-
-\section*{Decision Trees}
+\\ \\
+\textbf{Decision Trees}\\
 Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.
@@
 @
-For example, the tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.
-
-\section*{Decision Rules}
-A decision rule is a simple IF-THEN statement consisting of a condition and a prediction.
-One or many rules can be used to make predictions.
+\textit{The tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.}
+\\ \\
+\textbf{Decision Rules} \\
+Decision rules are IF-THEN statements consisting of a condition and a prediction.
+One or many rules can be used to make predictions. For example:
 
@@
 
-One decision rule learned for the rent prediction is:
+\textit{
+IF $90m^2\leq \text{size} < 110m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR}
 
-IF living area $\in 90 \text{ and } 110 m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR
+}
+\block{Surrogate Models}{
+\textbf{Global Surrogate Models} \\
+Intrinsically interpretable models (e.g.
+tree)  can also be used as to increase interpretability of a black box model (e.g.
+neural network) by approximating its predictions, acting as a surrogate model which can be interpreted.
+<<surrogate>>=
+knitr::include_graphics("figure/global-surrogate.png")
+@
+\\ \\
+\textbf{Local Surrogate Models}\cite{ribeiro2016should} \\
+Replace a  complex model with an individual, locally weighted interpretable model (e.g. decision tree) for the data point for which we want to explain the prediction.
 
-\section*{Global Surrogate Models}
-Intrinsically interpretable models can also be used as to increase interpretability of a black box model by approximating its predictions, acting as a surrogate model which can be interpreted. 
-Example: Neural network as black box model, decision tree as surrogate model.
-
-\section*{Local Surrogate Models}
-The idea behind local surrogate models \cite{ribeiro2016should} is to replace the complex model with an individual, locally weighted interpretable model (e.g.
-decision tree) for the data point for which we want to explain the prediction.
-
-<<lime, echo=FALSE>>=
+<<lime, echo=FALSE, out.width="0.2\\textwidth">>=
 # locmod = LocalModel$new(pred, k = 2)

### 2018-10-23 poster/poster_2018.Rnw
adds improvements

@@
 
-\definecolorpalette{GrayOrangeBlue}{
-\definecolor{colorOne}{HTML}{C0C0C0}
-\definecolor{colorTwo}{HTML}{CCCCCC}
-\definecolor{colorThree}{HTML}{009440}
-}
 \usepackage{hyperref}
-%\usetheme{Desert}
-%\usecolorstyle[colorPalette=BlueGrayOrange]{Russia}
-\usebackgroundstyle{Rays}
+\usetheme{Board}
+\usecolorstyle[colorPalette=PurpleGrayBlue]{Germany}
+
+\usebackgroundstyle{Default}
 \title{Interpretable Machine Learning}
-\author{Christoph Molnar \\ }
-\institute{
-LMU Munich / Department of Statistics/ \url{christoph.molnar@stat.uni-muenchen.de} \\
-Prof. Dr. Bernd Bischl}
-%\titlegraphic{\hspace{-3200px}\includegraphics[scale=0.3]{figures/logo.png}}
+\author{Christoph Molnar  \url{christoph.molnar@stat.uni-muenchen.de}}
+\institute{LMU Munich / Department of Statistics / Prof. Dr. Bernd Bischl}
+%\titlegraphic{\includegraphics[scale=0.1]{figure/logo.png}}
 \usetitlestyle{Empty}
-\colorlet{titlefgcolor}{black}
-\colorlet{titlebgcolor}{colorOne}
+% \colorlet{titlefgcolor}{black}
+% \colorlet{titlebgcolor}{colorOne}
 
@@
 
-\makeatletter
-\renewcommand\section{%
-\@startsection
-{section}% name
-{1}% level
-{\z@}% indent
-{0ex \@plus -1ex \@minus -.2ex}% beforeskip, changed!
-{2.3ex \@plus.2ex}% afterskip
-{\normalfont\Large\bfseries}% style
-}
-\makeatother
 \begin{document}
@@
 \column{0.33}
-\block{Machine learning}{
-Machine learning is a set of methods that let a computer learn from data to make predictions (e.g. cancer, weekly sales, credit default) by finding patterns.
+\block{Machine Learning}{
+Machine learning is a set of methods that allow computers to learn from data to make and improve predictions (e.g. cancer, weekly sales, credit default).
 <<ml>>=
@@
 
-Solution: \textbf{Interpretable machine learning} is a set of methods and models that enable to explain the behaviour and predictions of machine learning systems.\cite{molnar}
+Solution: \textbf{Interpretable Machine Learning} is a set of methods and models that make it possible to explain the behaviour and predictions of machine learning systems.\cite{molnar}
 }
 
-\block{Why explain?}{
-We need to explain machine learning models to \cite{adadi2018peeking}:
+\block{Why Explain?}{
+We need to explain machine learning models to ... \cite{adadi2018peeking}:
 % \begin{itemize}
@@
 \block{Example: Predict Rent}{
-As example we train a machine learning model (here a random forest) that predicts the rent of an appartment given the size of its living area, the location ("Good or "Bad") and if cats are allowed.
+
+As an application example we train a machine learning model (here a random forest), which predicts the rent of an apartment based on the size of the living space, the location ("good" or "bad") and whether cats are allowed.
 
@@
 
-\block{Feature effects}{
-The feature effect describe how changing a features changes the prediction of all data points on average. \cite{friedman2001greedy}
+\block{Feature Effect}{
+
+The feature effect describes how changing a feature changes on average the prediction of all data points. \cite{friedman2001greedy}
+
 <<ice, echo=FALSE, fig.width=12, fig.height=6>>=
@@
 @
-\textit{For example we can vary the size of the living area and observe how the prediction of the model changes.}
+
+\textit{If we vary the size of the living area, we observe how the predicted rente increases with increasing size.}
 
@@
 }
-\block{Feature importance}{
-The feature importance \cite{Fisher2018} tells us how much the model accuracy drops when we don't know a feature, which is simulated by shuffling the feature values.
-The bigger the drop, the more important.
+\block{Feature Importance}{
+Feature importance \cite{Fisher2018} tells us how much model error increases when we shuffle its values in the data (= "destroying" the relationship between the feature and the outcome).
+The greater the increase in error, the more important the feature is.
 <<feature-importance, echo=FALSE, fig.width=12, fig.height=8>>=
@@
 @
-\textit{In the rent example the most important feature was the size of the living area. }
+\textit{In the rent example the most important feature was the size of the living area. Shuffling the size feature increases the model error by a factor of 2.5.}
 }
-\block{Individual predictions}{
-
-<<appartment, echo=FALSE,include=FALSE>>=
-# Choose appartment
-appartment = data.frame(location = factor("Good", levels = c("Bad", "Good")),
-    cats = factor("Yes", levels = c("No", "Yes")),
+<<apartment, echo=FALSE,include=FALSE>>=
+# Choose apartment
+apartment = data.frame(location = factor("Good", levels = c("Bad", "Good")),
+    cats = factor("No", levels = c("No", "Yes")),
   size = 50)
-pred.rent = round(pred$predict(appartment)$.prediction, 0)
+pred.rent = round(pred$predict(apartment)$.prediction, 0)
 @
-
-We have a couple of options to explain why a particular prediction was made by a machine learning model.
-Suppose we want to explain the prediction of the following appartment. 
+\block{Individual Predictions}{
+Sometimes we want to explain why a certain prediction was made by a machine learning model.
+\textit{Suppose we want to explain the 989Euro rent prediction for a $50m^2$ apartment in a good location and where cats are forbidden.}
 <<rent2, out.width="0.2\\textwidth">>=
@@
 <<counterfact, echo=FALSE, include = FALSE>>=
-appartment.mod = appartment
-appartment.mod$cats = factor("No", levels = c("No", "Yes"))
-pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
+apartment.mod = apartment
+apartment.mod$cats = factor("Yes", levels = c("No", "Yes"))
+pred.new = round(pred$predict(apartment.mod[,1:3, drop=FALSE]), 0)
 @
-\\ \\
-\textbf{Counterfactual explanations} \\
-A counterfactual explanation of a prediction describes the smallest changes to the feature values (inputs) that change the prediction to a predefined output.
-For example: How does the inputs have to change so that the prediction changes below 900 Euros?
-Answer (one of many possible ones): \\
-\textit{If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.}
+
+\textbf{Counterfactual Explanation} \\
+A counterfactual explanation of a prediction describes the smallest changes to the feature values (inputs) that change the prediction to a predefined output. \\
+\textit{How do the inputs for the 989 Euro apartment have to change so that the predicted rent is over 1100 Euros?
+Answer (one of many possible ones): If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.}
 \\ \\
 \textbf{Shapley Values} \\
-From a game theoretical perspective, the feature values can be seen as players in a game that cooperate to achieve the prediction as a payout.
-The Shapley Values method \cite{vstrumbelj2014explaining} fairly attributes the difference between the prediction and the average prediction among the feature values.
+From a game theory point of view, the feature values are players in a cooperative game who receive the value of the prediction as a payout.
+The Shapley Values method \cite{vstrumbelj2014explaining} splits the difference between the prediction and the average prediction fairly among the feature values.
 
@@
 shap = Shapley$new(pred)
-shap$explain(appartment)
+shap$explain(apartment)
+avg.pred  = round(mean(pred$predict(df)[[1]]))
 plot(shap) +
@@
 @
+\textit{The predicted value of the apartment is 989Euros, the average prediction of all apartments is \Sexpr{avg.pred}.
+The negative difference is explained by its small size and the ban on cats.
+The good location has a positive effect on the predicted rent.}
 
@@
 \block{Interpretable Models}{
-Intrinsically interpretable models learn simple relationships between input and output,
-but often predict the data less well than flexible, yet less interpretable, models.
+Intrinsically interpretable models learn simple relationships between input and output.
 \\ \\
@@
 \textbf{Decision Trees}\\
-Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.
+Decision trees divide the data into smaller subsets based on 'decisions' made on input features.
 
@@
 @
-\textit{The tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.}
+\textit{The tree predicts a rent of 1908 Euros for an apartment larger than 77 $m^2$ in a good location.}
 \\ \\
 \textbf{Decision Rules} \\
-Decision rules are IF-THEN statements consisting of a condition and a prediction.
-One or many rules can be used to make predictions. For example:
-
+Decision rules are IF-THEN statements that consist of a condition and a prediction.
+One or more rules can be used to make predictions. For example:
 <<train-model-rules, echo=FALSE, includee=FALSE, eval=FALSE>>=
@@
 @
-
 \textit{
@@
 \textbf{Global Surrogate Models} \\
+%Flexible but less interpretable models often predict the data better than interpretable models.
 Intrinsically interpretable models (e.g.
-tree)  can also be used as to increase interpretability of a black box model (e.g.
-neural network) by approximating its predictions, acting as a surrogate model which can be interpreted.
+a tree) can be used to increase interpretability of a black box model (e.g.
+a neural network) by approximating its predictions by acting as a surrogate model that can be interpreted.
 <<surrogate>>=
@@
 @
-\\ \\
+
 \textbf{Local Surrogate Models}\cite{ribeiro2016should} \\
-Replace a  complex model with an individual, locally weighted interpretable model (e.g. decision tree) for the data point for which we want to explain the prediction.
+To explain an individual prediction, replace a  complex model with a locally weighted, interpretable model (e.g. decision tree).
+Data points that are close to the data point of interest get a high weight.
 
-<<lime, echo=FALSE, out.width="0.2\\textwidth">>=
+<<lime, echo=FALSE>>=
 # locmod = LocalModel$new(pred, k = 2)

### 2018-10-23 poster/poster_2018.Rnw
final proofread

@@
 \usepackage{hyperref}
+% \definecolorpalette{BlueGrayOrange}{
+%     \definecolor{colorOne}{HTML}{C0C0C0}
+%     \definecolor{colorTwo}{HTML}{CCCCCC}
+%     \definecolor{colorThree}{HTML}{009440}
+% }
+\definecolorpalette{LMU}{
+    \definecolor{colorTwo}{HTML}{CCCCCC}
+    \definecolor{colorThree}{HTML}{C0C0C0}
+    \definecolor{colorOne}{HTML}{009440}
+}
+
 \usetheme{Board}
-\usecolorstyle[colorPalette=PurpleGrayBlue]{Germany}
+\usecolorstyle[colorPalette=LMU]{Germany}
 
-\usebackgroundstyle{Default}
+% \usebackgroundstyle{Default}
+
 \title{Interpretable Machine Learning}
@@
 %\titlegraphic{\includegraphics[scale=0.1]{figure/logo.png}}
-\usetitlestyle{Empty}
-% \colorlet{titlefgcolor}{black}
-% \colorlet{titlebgcolor}{colorOne}
+\usetitlestyle{Autumn}
+\colorlet{titlefgcolor}{black}
+\colorlet{titlebgcolor}{white}
 
@@
   scale_x_continuous("Size of living area") + 
-  scale_y_continuous("Predicted rent") +
+  scale_y_continuous("Average predicted rent") +
   theme(aspect.ratio=2/4, text = element_text(size=35),
@@
 
-\textit{If we vary the size of the living area, we observe how the predicted rente increases with increasing size.}
+\textit{If we vary the size of the living area, we observe how the predicted rent increases with increasing size.}
 
@@
 \block{Feature Importance}{
-Feature importance \cite{Fisher2018} tells us how much model error increases when we shuffle its values in the data (= "destroying" the relationship between the feature and the outcome).
+Feature importance \cite{Fisher2018} tells us how much the model error increases when we shuffle the values of a feature in the data (= "destroying" the relationship between the feature and the outcome).
 The greater the increase in error, the more important the feature is.
@@
 \textbf{Counterfactual Explanation} \\
-A counterfactual explanation of a prediction describes the smallest changes to the feature values (inputs) that change the prediction to a predefined output. \\
+A counterfactual explanation describes the smallest changes to a data points feature values (inputs)  that change the prediction to a predefined output. \\
 \textit{How do the inputs for the 989 Euro apartment have to change so that the predicted rent is over 1100 Euros?
@@
 From a game theory point of view, the feature values are players in a cooperative game who receive the value of the prediction as a payout.
-The Shapley Values method \cite{vstrumbelj2014explaining} splits the difference between the prediction and the average prediction fairly among the feature values.
+The Shapley Value method \cite{vstrumbelj2014explaining} splits the difference between the prediction and the average prediction fairly among the feature values.
 
@@
 \textit{The predicted value of the apartment is 989Euros, the average prediction of all apartments is \Sexpr{avg.pred}.
-The negative difference is explained by its small size and the ban on cats.
+The negative difference is explained by the apartments small size and the ban on cats.
 The good location has a positive effect on the predicted rent.}
@@
 \textit{
-IF $90m^2\leq \text{size} < 110m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR}
+IF $90m^2\leq \text{size} < 110m^2$  AND location $=$ "good" THEN rent is between 1540 and 1890 EUR}
 
@@
 Data points that are close to the data point of interest get a high weight.
-
+\vspace{0.5cm}
 <<lime, echo=FALSE>>=

### 2018-12-13 poster/poster_2018.Rnw
create talk for causality workshop;

@@
 <<ml>>=
-knitr::include_graphics("figure/programing-ml.png")
+knitr::include_graphics("../images/programing-ml.png")
 @
@@
 <<iml>>=
-knitr::include_graphics("figure/iml.png")
+knitr::include_graphics("../images/iml.png")
 @
@@
 <<explain-to>>=
-knitr::include_graphics("figure/explain.png")
+knitr::include_graphics("../images/explain.png")
 @
@@
 <<rent>>=
-knitr::include_graphics("figure/rent-ml.png")
+knitr::include_graphics("../images/rent-ml.png")
 @
@@
 <<rent2, out.width="0.2\\textwidth">>=
-knitr::include_graphics("figure/rent.png")
+knitr::include_graphics("../images/rent.png")
 @
@@
 <<lm>>=
-knitr::include_graphics("figure/lm.png")
+knitr::include_graphics("../images/lm.png")
 @
@@
 #plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
-knitr::include_graphics("figure/tree.png")
+knitr::include_graphics("../images/tree.png")
 @
@@
 <<surrogate>>=
-knitr::include_graphics("figure/global-surrogate.png")
+knitr::include_graphics("../images/global-surrogate.png")
 @
@@
 # plot(locmod)
-knitr::include_graphics("figure/local-surrogate.png")
+knitr::include_graphics("../images/local-surrogate.png")
 @

### 2019-01-08 2018-appliedr/talk.Rmd
clean up repo

--- 2018-03-27 talk.Rmd
@@
   ioslides_presentation:
-    css: styles.css
+    css: ../styles.css
 ---
@@
 
-## Step 1: Find some data
-Kaggle conducted an industry-wide survey of data scientists. 
-https://www.kaggle.com/kaggle/kaggle-survey-2017
-
-Information asked:
-
-- Compensation
-- Demographics
-- Job title
-- Experience
-- ...
-
-<font size="2">Contains information from Kaggle ML and Data Science Survey, 2017, which is made available here under the Open Database License (ODbL).</font>
-
-```{r load-data, warning=FALSE, include=FALSE}
-library('mlr')
-library('ggplot2')
-library('tidyr')
-library('lime')
-source('code/prepare-kaggle-data.R')
-```
-
-## Step 2: Throw ML on your data
-```{r learn, warning=FALSE}
-library('mlr')
-set.seed(42)
-task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
-lrn = makeLearner('regr.gbm', interaction.depth = 4, n.trees = 500)
-mod = train(lrn, task)
-```
-## Step 3: Profit. {.center}
-
-```{r, echo=FALSE, out.width='70%', fig.align='center'}
-knitr::include_graphics("images/done-here.gif")
-```
-
-##  {.emphasizedabit}
-
-"There is a problem with the model!"
-
-## What problem?
-
-```{r, echo=FALSE, out.width='100%', fig.align='center'}
-knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
-```
-
-## "The older the applicants, the higher the predicted salary, regardless of skills."
-```{r, echo=FALSE, out.width='100%', fig.align='center'}
-knitr::include_graphics("images/age.jpeg")
-```
-
-# Individual Conditional Expectation
-
-##
-
-```{r}
-library("iml")
-X = survey.dat[-which(names(survey.dat) == "CompensationAmount")]
-predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
-ice = Ice$new(predictor, feature ='Age')
-ice$plot()
-```
-
-<font size="2">Goldstein, A., Kapelner, A., Bleich, J., & Pitkin, E.
-(2013).
-Peeking Inside the Black Box: Visualizing Statistical Learning with Plots of Individual Conditional Expectation, 1–22.
-https://doi.org/10.1080/10618600.2014.907095 </font>
-
-##
-
-```{r}
-ice$center(20)$plot()
-```
-
-# Partial dependence plots
-
-##
-
-```{r}
-pdp = PartialDependence$new(predictor, feature = "Age")
-pdp$plot()
-```
-
-<font size="2">Friedman, J. H. (1999). Greedy Function Approximation : A Gradient Boosting Machine. North, 1(3), 1–10. https://doi.org/10.2307/2699986 </font>
-
-##  {.emphasizedabit}
-
-"We want to understand the model better!"
-
-# Permutation feature importance
-
-##
-
-```{r, warning=FALSE}
-feat.imp = FeatureImp$new(predictor, loss = 'mae')
-feat.imp$plot()
-```
-
-<font size="2">Breiman, Leo. "Random forests." Machine learning 45.1 (2001): 5-32. </font>
-
-## Gender?!
-```{r, echo=FALSE, fig.align='center', out.width='80%'}
-knitr::include_graphics("images/big-mistake.png")
-```
-
-##
-
-```{r}
-pdp = PartialDependence$new(predictor, feature="Gender")
-pdp$plot() 
-```
-
-##
-
-```{r, echo=FALSE, fig.align='center', out.width='100%'}
-knitr::include_graphics("images/facepalm.gif")
-```
-
-##
-
-```{r, echo=FALSE, fig.align='center'}
-knitr::include_graphics("images/hidden-pain-bias.jpg")
-```
-
-##
-
-```{r, echo=FALSE, fig.align='center', out.width='40%'}
-knitr::include_graphics("images/angry.gif")
-```
-
-# LIME
-
-## 
-```{r, echo=FALSE, fig.align='center'}
-set.seed(44)
-```
-
-```{r, echo=TRUE, fig.align='center', message = FALSE}
-explanation = LocalModel$new(predictor)
-explanation$explain(X[5, ])$plot()
-```
-
-<font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
-
-# Shapley
-
-## 
-```{r, echo=FALSE, fig.align='center'}
-set.seed(44)
-```
-
-```{r, echo=TRUE, fig.align='center'}
-explanation = Shapley$new(predictor)
-explanation$explain(X[5, ])$plot()
+```{r child = "../chunks/story-salaries.Rmd"}
 ```
@@
 ```
+
+```{r child = "../chunks/book.Rmd"}
+```

### 2019-01-08 2018-datageeks/talk-long.Rmd
clean up repo

--- 2018-08-20 talk-40min.Rmd
@@
     smaller: false
-css: styles.css
+css: ../styles.css
 ---
@@
 
-##  {.emphasizedabit}
-A client wants you to predict data scientist salaries with machine learning.
-
-## {.center data-background=images/wow.jpg data-background-size=contain}
-
-# Let's predict data scientist salaries
-
-# What is Machine Learning?
-
-<div class="notes">
-Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
-</div>
-
-## {data-background=images/magic.jpg data-background-size=contain}
-
-# Step 1: Find some data
-
-## Step 1: Find some data
-
-Kaggle conducted an industry-wide survey of data scientists. 
-https://www.kaggle.com/kaggle/kaggle-survey-2017
-
-Information asked:
-
-- Compensation
-- Demographics
-- Job title
-- Experience
-- ...
-
-<font size="2">Contains information from Kaggle ML and Data Science Survey, 2017, which is made available here under the Open Database License (ODbL).</font>
-
-```{r load-data,include=FALSE}
-library('mlr')
-library('ggplot2')
-library('tidyr')
-library('lime')
-source('code/prepare-kaggle-data.R')
-```
-
-# Step 2: Throw ML on your data
-
-```{r learn}
-library('mlr')
-set.seed(42)
-task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
-lrn = makeLearner('regr.randomForest')
-mod = train(lrn, task)
-```
-
-## {.center data-background=images/comp-dog.gif data-background-size=contain}
-
-<div class="notes">
-- Random Forest
-- Target: Compensation
-- All features that you found in the data
-- Train, test, ship
-</div>
-
-# Step 3: Profit {.center}
-
-## {data-background=images/done-here.gif data-background-size=contain}
-```{r, echo=FALSE, out.width='80%', include = FALSE}
-knitr::include_graphics("images/done-here.gif" )
-```
-
-##  {.emphasizedabit}
-
-Client: "There is a problem with the model!"
-
-##  {.emphasizedabit .center data-background=images/Hide-the-pain-harold-phone.jpg data-background-size=cover}
-
-"What problem?"
-
-```{r, echo=FALSE, out.height='50%', include=FALSE}
-knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
-```
-
-## {.emphasizedabit}
-
-Client: "Model predicts high salaries for old yet unskilled people."
-
-```{r, echo=FALSE, fig.width=16, include=FALSE}
-knitr::include_graphics("images/age.jpeg")
-```
-
-##  {.emphasizedabit .center data-background=images/black-box.gif data-background-size=cover}
-
-<div class="white">
-Looking inside the black box
-</div>
-
-# How do the features influence my predictions?
-
-## Partial Dependence Plot
-
-```{r out.width="80%"}
-library("iml")
-X = survey.dat[-which(names(survey.dat) == "CompensationAmount")]
-predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
-pd = Partial$new(predictor, feature ='Age')
-pd$plot() 
-```
-
-<font size="2">Goldstein, A., Kapelner, A., Bleich, J., & Pitkin, E.
-(2013).
-Peeking Inside the Black Box: Visualizing Statistical Learning with Plots of Individual Conditional Expectation, 1–22.
-https://doi.org/10.1080/10618600.2014.907095 </font>
-
-<font size="2">Friedman, J. H. (1999). Greedy Function Approximation : A Gradient Boosting Machine. North, 1(3), 1–10. https://doi.org/10.2307/2699986 </font>
-
-<div class="notes">
-- PDP answers: How does the prediction change when a feature input changes?
-- Average of ICE curves
-</div>
-
-##  {.emphasizedabit}
-
-Client: "We want to understand the model better!"
-
-# What are the most important features?
-
-## Permutation feature importance
-
-```{r, warning=FALSE}
-feat.imp = FeatureImp$new(predictor, loss = 'mae')
-feat.imp$plot()
-```
-
-<font size="2">Breiman, Leo. "Random forests." Machine learning 45.1 (2001): 5-32. </font>
-
-## Gender?! {.emphasizedabit .center data-background=images/big-mistake.png data-background-size=contain}
-
-```{r, echo=FALSE, out.width='80%', include=FALSE}
-knitr::include_graphics("images/big-mistake.png")
-```
-
-# What's the influence of gender on the prediction?
-
-##
-
-```{r}
-pdp = Partial$new(predictor, feature="Gender")
-pdp$plot() 
-```
-
-## {.center data-background=images/hidden-pain-bias.jpg data-background-size=contain}
-
-```{r, echo=FALSE, include = FALSE}
-knitr::include_graphics("images/hidden-pain-bias.jpg")
-```
-
-## {.center data-background=images/angry.gif data-background-size=contain}
-
-```{r, echo=FALSE, out.width='40%', include = FALSE}
-knitr::include_graphics("images/angry.gif")
-```
-
-# Explaining individual predictions
-
-# Local Models (LIME)
-
-## {.center data-background=images/lime-fitting-1.png data-background-size=contain}
-
-## 
-```{r, echo=FALSE}
-set.seed(44)
-```
-
-```{r}
-explanation = LocalModel$new(predictor)
-explanation$explain(X[15, ])
-explanation$plot()
-```
-
-<font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
-
-# Shapley Value
-
-## {.center data-background=images/team.jpg data-background-size=contain}
-
-## 
-```{r}
-set.seed(44)
-```
-
-```{r}
-explanation = Shapley$new(predictor)
-explanation$explain(X[15, ])
-explanation$plot()
+```{r, child="../chunks/story-salaries.Rmd"}
 ```
@@
 
-# What tools do we have?
-
-<div class="notes">
-- Intrinsically interpretable models
-- Tools for analyzing specific black box models
-- Model-agnostic tools: analysing any black box model
-- Interpretability through making data explainable
-</div>
-
-## Interpretable Models
-
-```{r, echo = FALSE}
-knitr::include_graphics("images/black-box.png")
+```{r child = "../chunks/tools.Rmd"}
 ```
 
-## Interpretable Models
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/white-box.png")
-```
-
-<div class="notes">
-- Decision trees
-- Decision rules
-- Generalized regression models
-- ...
-</div>
-
-## Model-specific methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Model-specific methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/specific-black-box.png")
-```
-
-<div class="notes">
-- Visualizing activations of neural networks
-- Random Forest specific feature importance
-- xgboost explainer
-- problem: new models, not comparable between models, missing flexibility
-</div>
-
-## Model-agnostic methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Model-agnostic methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/agnostic-black-box.png")
-```
-
-<div class="notes">
-- Partial dependence plots
-- LIME
-- ...
-</div>
-
-## Example-focused methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Example-focused methods
-```{r, echo=FALSE}
-knitr::include_graphics("images/data-box.png")
-```
-
-<div class="notes">
-- Prototypes: What is the prototype of a certain class
-- Criticisms: Which data points are unusual for a certain class. 
-- Counterfactuals: What do I need to change for an instance to change the prediction?
-- Anchors: What features do I have to anchor to keep prediction the same?
-</div>
-
-## Interested in learning more?
-
-Read my book about "Interpretable Machine Learning"
-https://christophm.github.io/interpretable-ml-book/
-
-```{r, echo=FALSE, fig.align='center', out.width='30%'}
-knitr::include_graphics("images/book.jpg")
+```{r child = "../chunks/book.Rmd"}
 ```
@@
 </div>
+
+# Shapley Value
+
+## {.center data-background=images/team.jpg data-background-size=contain}
+
+## 
+```{r}
+set.seed(44)
+```
+
+```{r}
+explanation = Shapley$new(predictor)
+explanation$explain(X[15, ])
+explanation$plot()
+```

### 2019-01-08 chunks/story-salaries.Rmd
clean up repo

--- 2018-08-17 talk-40min.Rmd
@@
----
-title: "Interpretable Machine Learning"
-author: "Christoph Molnar"
-date: "June 7, 2018"
-output: 
-  ioslides_presentation:
-    widescreen: false
-    smaller: false
-css: styles.css
----
-
-```{r, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, 
-  warning = FALSE,
-  message = FALSE, 
-  fig.align='center', 
-  out.width = '100%')
-```
-
 ##  {.emphasizedabit}
@@
 
-## {.center data-background=images/wow.jpg data-background-size=contain}
+## {.center data-background=../images/wow.jpg data-background-size=contain}
 
@@
 
-## What is Machine Learning
+# What is Machine Learning?
 
+<div class="notes">
 Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
+</div>
 
-## {data-background=images/magic.jpg data-background-size=contain}
+## {data-background=../../images/magic.jpg data-background-size=contain}
 
@@
 library('lime')
-source('code/prepare-kaggle-data.R')
+source('../code/prepare-kaggle-data.R')
 ```
@@
 
-## {.center data-background=images/comp-dog.gif data-background-size=contain}
+## {.center data-background=../images/comp-dog.gif data-background-size=contain}
 
@@
 
-# Step 3: Profit. {.center}
+# Step 3: Profit {.center}
 
-## {data-background=images/done-here.gif data-background-size=contain}
+## {data-background=../images/done-here.gif data-background-size=contain}
 ```{r, echo=FALSE, out.width='80%', include = FALSE}
-knitr::include_graphics("images/done-here.gif" )
+knitr::include_graphics("../images/done-here.gif" )
 ```
@@
 
-##  {.emphasizedabit .center data-background=images/Hide-the-pain-harold-phone.jpg data-background-size=cover}
+##  {.emphasizedabit .center data-background=../images/Hide-the-pain-harold-phone.jpg data-background-size=cover}
 
@@
 ```{r, echo=FALSE, out.height='50%', include=FALSE}
-knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
+knitr::include_graphics("../images/Hide-the-pain-harold-phone.jpg")
 ```
@@
 
-Client: "Model predicts high salaries for old yet unskilled people."
+Client: "The older the candidates, the higher the predicted salaries."
 
 ```{r, echo=FALSE, fig.width=16, include=FALSE}
-knitr::include_graphics("images/age.jpeg")
+knitr::include_graphics("../images/age.jpeg")
 ```
 
-##  {.emphasizedabit .center data-background=images/black-box.gif data-background-size=cover}
+##  {.emphasizedabit .center data-background=../images/black-box.gif data-background-size=cover}
 
@@
 
-## Gender?! {.emphasizedabit .center data-background=images/big-mistake.png data-background-size=contain}
+## Gender?! {.emphasizedabit .center data-background=../images/big-mistake.png data-background-size=contain}
 
 ```{r, echo=FALSE, out.width='80%', include=FALSE}
-knitr::include_graphics("images/big-mistake.png")
+knitr::include_graphics("../images/big-mistake.png")
 ```
 
-# What's the influence of gender on the prediction?
-
-##
+## What's the influence of gender on the prediction?
 
 ```{r}
-pdp = Partial$new(predictor, feature="Gender")
-pdp$plot() 
+pdp = Partial$new(predictor, feature="Gender", ice = FALSE)
+knitr::kable(pdp$results[c("Gender", ".y.hat")])
 ```
 
-## {.center data-background=images/hidden-pain-bias.jpg data-background-size=contain}
+## {.center data-background=../images/hidden-pain-bias.jpg data-background-size=contain}
 
 ```{r, echo=FALSE, include = FALSE}
-knitr::include_graphics("images/hidden-pain-bias.jpg")
+knitr::include_graphics("../images/hidden-pain-bias.jpg")
 ```
 
-## {.center data-background=images/angry.gif data-background-size=contain}
+## {.center data-background=../images/angry.gif data-background-size=contain}
 
 ```{r, echo=FALSE, out.width='40%', include = FALSE}
-knitr::include_graphics("images/angry.gif")
+knitr::include_graphics("../images/angry.gif")
 ```
@@
 
-# LIME
+# Local Models (LIME)
 
-## {.center data-background=images/lime-fitting-1.png data-background-size=contain}
+## {.center data-background=../images/lime-fitting-1.png data-background-size=contain}
 
@@
 <font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
-
-# Shapley Value
-
-##
-
-TODO: find image for explaining shapley value
-
-## 
-```{r}
-set.seed(44)
-```
-
-```{r}
-explanation = Shapley$new(predictor)
-explanation$explain(X[15, ])
-explanation$plot()
-```
-
-# Some Theory
-
-## The Problem
-
-```{r, echo=FALSE, fig.align='center'}
-knitr::include_graphics("images/black-box.png")
-```
-<div class="notes">
-- Opaque decision making by machine learning algorithms
-- Bias and mistakes in the data
-- Trust as a user: Should I start a therapy with sever side effects, because the machine said so?
-- Debugging as a practitioner: Why did the algorithm miss-classify sample X? Did it learn generalizable features? 
-</div>
-
-## What is interpretability?
-
-<div class="notes">
-Interpretability is the degree to which a human can understand the cause of a decision. 
-But we don't really have a good way to measure that really. Not as easy as benchmarking ML algorithms.
-
-Interpretability is also a mean to look at the data and possible issues with them.
-
-<font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>
-</div>
-
-## When do we need interpretability?
-
-<div class="notes">
-We need interpretability when the loss function does not cover all constraints.
-
-- Decisions about humans.
-- Critical applications that decide about life and death.
-- Newly developed systems with unknown consequences.
-- Models using proxies instead of causal inputs.
-- Debugging the models.
-- Increasing trust.
-</div>
-
-## Examples where we need interpretability
-
-TODO: Insert images
-- Credit scores
-- Autonomous vehicles
-- Medical recommendations
-- Automated diagnoses
-
-## When don't we need interpretability?
-
-When we can capture everything in the loss function and the data collection.
-only causal relationships.
-perfect operationalization of features.
-
-Things that work well. 
-Well defined problems
-
-## Example: Digit recognition {.center data-background=images/mnist.jpeg data-background-size=cover}
-
-## What tools do we have?
-
-TODO: replace with image
-
-- Intrinsically interpretable models
-- Tools for analyzing specific black box models
-- Model-agnostic tools: analysing any black box model
-- Interpretability through making data explainable
-
-## Interpretable Models
-
-```{r, echo = FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Interpretable Models
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/white-box.png")
-```
-
-<div class="notes">
-- Decision trees
-- Decision rules
-- Generalized regression models
-- ...
-</div>
-
-## Model-specific methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Model-specific methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/specific-black-box.png")
-```
-
-<div class="notes">
-- Visualizing activations of neural networks
-- Random Forest specific feature importance
-- xgboost explainer
-- problem: new models, not comparable between models, missing flexibility
-</div>
-
-## Model-agnostic methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Model-agnostic methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/agnostic-black-box.png")
-```
-
-<div class="notes">
-- Partial dependence plots
-- LIME
-- ...
-</div>
-
-## Example-focused methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
-```
-
-## Example-focused methods
-```{r, echo=FALSE}
-knitr::include_graphics("images/data-box.png")
-```
-
-<div class="notes">
-- Prototypes: What is the prototype of a certain class
-- Criticisms: Which data points are unusual for a certain class. 
-- Counterfactuals: What do I need to change for an instance to change the prediction?
-- Anchors: What features do I have to anchor to keep prediction the same?
-</div>
-
-## Interested in learning more?
-
-Read my book about "Interpretable Machine Learning"
-https://christophm.github.io/interpretable-ml-book/
-
-```{r, echo=FALSE, fig.align='center', out.width='30%'}
-knitr::include_graphics("images/book.jpg")
-```

### 2019-01-08 chunks/tools.Rmd
clean up repo

--- 2018-06-04 talk-40min.Rmd
@@
----
-title: "Interpretable Machine Learning"
-author: "Christoph Molnar"
-date: "June 7, 2018"
-output: 
-  ioslides_presentation:
-    widescreen: false
-    smaller: false
-css: styles.css
----
+# What tools do we have?
 
-```{r, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, 
-  warning = FALSE,
-  message = FALSE, 
-  fig.align='center', 
-  out.width = '100%')
-```
-
-##  {.emphasizedabit}
-A client wants you to predict data scientist salaries with machine learning.
-
-##
-
-```{r, echo=FALSE, out.height = '100%', out.width=NULL}
-knitr::include_graphics("images/wow.jpg")
-```
-
-# Let's predict data scientist salaries
-
-## What is Machine Learning
-
-Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
-
-```{r, echo=FALSE, out.height = '100%', out.width=NULL}
-knitr::include_graphics("images/magic.jpg")
-```
-
-## Step 1: Find some data
-Kaggle conducted an industry-wide survey of data scientists. 
-https://www.kaggle.com/kaggle/kaggle-survey-2017
-
-Information asked:
-
-- Compensation
-- Demographics
-- Job title
-- Experience
-- ...
-
-<font size="2">Contains information from Kaggle ML and Data Science Survey, 2017, which is made available here under the Open Database License (ODbL).</font>
-
-```{r load-data,include=FALSE}
-library('mlr')
-library('ggplot2')
-library('tidyr')
-library('lime')
-source('code/prepare-kaggle-data.R')
-```
-
-## Step 2: Throw ML on your data
-```{r learn}
-library('mlr')
-set.seed(42)
-task = makeRegrTask(data = survey.dat, target = 'CompensationAmount')
-lrn = makeLearner('regr.randomForest')
-mod = train(lrn, task)
-```
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/comp-dog.gif")
-```
-
-## Step 3: Profit. {.center}
-
-```{r, echo=FALSE, out.height=NULL, out.width='100%'}
-knitr::include_graphics("images/done-here.gif")
-```
-
-##  {.emphasizedabit}
-
-Client: "There is a problem with the model!"
-
-## {.emphasizedabit}
-
-"What problem?""
-
-```{r, echo=FALSE, out.height='50%'}
-knitr::include_graphics("images/Hide-the-pain-harold-phone.jpg")
-```
-
-## {.emphasizedabit}
-
- Client: "Model predicts high salaries for old yet unskilled people."
-
-```{r, echo=FALSE, out.height='30%'}
-knitr::include_graphics("images/age.jpeg")
-```
-
-# Partial Dependence Plots
-
-##
-
-```{r}
-library("iml")
-X = survey.dat[-which(names(survey.dat) == "CompensationAmount")]
-predictor = Predictor$new(mod, data = X, y = getTaskTargets(task))
-pd = Partial$new(predictor, feature ='Age')
-pd$plot() 
-```
-
-<font size="2">Goldstein, A., Kapelner, A., Bleich, J., & Pitkin, E.
-(2013).
-Peeking Inside the Black Box: Visualizing Statistical Learning with Plots of Individual Conditional Expectation, 1–22.
-https://doi.org/10.1080/10618600.2014.907095 </font>
-
-<font size="2">Friedman, J. H. (1999). Greedy Function Approximation : A Gradient Boosting Machine. North, 1(3), 1–10. https://doi.org/10.2307/2699986 </font>
-
-##  {.emphasizedabit}
-
-Client: "We want to understand the model better!"
-
-# Permutation feature importance
-
-##
-
-```{r, warning=FALSE}
-feat.imp = FeatureImp$new(predictor, loss = 'mae')
-feat.imp$plot()
-```
-
-<font size="2">Breiman, Leo. "Random forests." Machine learning 45.1 (2001): 5-32. </font>
-
-## Gender?!
-```{r, echo=FALSE, out.width='80%'}
-knitr::include_graphics("images/big-mistake.png")
-```
-
-##
-
-```{r}
-pdp = Partial$new(predictor, feature="Gender")
-pdp$plot() 
-```
-
-##
-
-```{r, echo=FALSE}
-knitr::include_graphics("images/hidden-pain-bias.jpg")
-```
-
-##
-
-```{r, echo=FALSE, out.width='40%'}
-knitr::include_graphics("images/angry.gif")
-```
-
-# LIME
-
-## 
-```{r, echo=FALSE}
-set.seed(44)
-```
-
-```{r}
-explanation = LocalModel$new(predictor)
-explanation$explain(X[15, ])
-explanation$plot()
-```
-
-<font size="2">Ribeiro, M. T., Singh, S., & Guestrin, C. (2016). “Why Should I Trust You?”: Explaining the Predictions of Any Classifier. Retrieved from http://arxiv.org/abs/1602.04938</font>
-
-# Shapley
-
-## 
-```{r}
-set.seed(44)
-```
-
-```{r}
-explanation = Shapley$new(predictor)
-explanation$explain(X[15, ])
-explanation$plot()
-```
-
-# Some Theory
-
-## The Problem
-
-```{r, echo=FALSE, fig.align='center'}
-knitr::include_graphics("images/black-box.png")
-```
-
-## The Problem
-
-- Opaque decision making by machine learning algorithms
-- Bias and mistakes in the data
-- Trust as a user: Should I start a therapy with sever side effects, because the machine said so?
-- Debugging as a practitioner: Why did the algorithm miss-classify sample X? Did it learn generalizable features?
-
-## What is interpretability?
-
-Interpretability is the degree to which a human can understand the cause of a decision. 
-But we don't really have a good way to measure that really. Not as easy as benchmarking ML algorithms.
-
-Interpretability is also a mean to look at the data and possible issues with them.
-
-<font size="2">Miller, Tim. 2017. “Explanation in Artificial Intelligence: Insights from the Social Sciences.” arXiv Preprint arXiv:1706.07269.</font>
-
-## When do we need interpretability?
-
-We need interpretability when the loss function does not cover all constraints.
-
-- Decisions about humans.
-- Critical applications that decide about life and death.
-- Newly developed systems with unknown consequences.
-- Models using proxies instead of causal inputs.
-- Debugging the models.
-- Increasing trust.
-
-### Examples where we need interpretability
-
-- Credit scores
-- Autonomous vehicles
-- Medical recommendations
-- Automated diagnoses
-
-## When don't we need interpretability?
-
-When we can capture everything in the loss function and the data collection.
-only causal relationships.
-perfect operationalization of features.
-
-Things that work well. 
-Well defined problems
-
-## Example: Digit recognition
-
-![](images/mnist.jpeg)
-
-## What tools do we have?
-
+<div class="notes">
 - Intrinsically interpretable models
@@
 ```{r, echo = FALSE}
-knitr::include_graphics("images/black-box.png")
+knitr::include_graphics("../images/black-box.png")
 ```
@@
 ```{r, echo=FALSE}
-knitr::include_graphics("images/white-box.png")
+knitr::include_graphics("../images/white-box.png")
 ```
 
-## Interpretable Models
-
+<div class="notes">
 - Decision trees
@@
 - ...
+</div>
+
+## Intepretable Model: Linear Regression {.center data-background=../images/lm.png data-background-size=contain}
+
+## Intepretable Model: Decision Tree {.center data-background=../images/tree.png data-background-size=contain}
+
+## Interpretable Model: Decision Rules {.emphasizedabit}
+
+IF $90m^2\
 
@@
 ```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
+knitr::include_graphics("../images/black-box.png")
 ```
@@
 ```{r, echo=FALSE}
-knitr::include_graphics("images/specific-black-box.png")
+knitr::include_graphics("../images/specific-black-box.png")
 ```
 
-## Model-specific methods
-
-- RandomForestExplainer
+<div class="notes">
 - Visualizing activations of neural networks
--
-
-TODO: add images from papers
+- Random Forest specific feature importance
+- xgboost explainer
+- problem: new models, not comparable between models, missing flexibility
+</div>
 
@@
 ```{r, echo=FALSE}
-knitr::include_graphics("images/agnostic-black-box.png")
+knitr::include_graphics("../images/agnostic-black-box.png")
 ```
 
-## Model-agnostic methods
-
+<div class="notes">
 - Partial dependence plots
@@
 - ...
+</div>
 
-## Example-focused methods
+## Example-based methods
 
 ```{r, echo=FALSE}
-knitr::include_graphics("images/black-box.png")
+knitr::include_graphics("../images/black-box.png")
 ```
@@
 ```{r, echo=FALSE}
-knitr::include_graphics("images/data-box.png")
+knitr::include_graphics("../images/data-box.png")
 ```
 
-## Example-focused methods
-
+<div class="notes">
 - Prototypes: What is the prototype of a certain class
@@
 - Anchors: What features do I have to anchor to keep prediction the same?
-
-TODO: Cite
-
-## Outcomes of interpretability method:
-
-- Model
-- Data point
-- Interpretable unit (like a weight)
-- Summary statistic for a feature
-
-## Interested in learning more?
-
-Read my book about "Interpretable Machine Learning"
-https://christophm.github.io/interpretable-ml-book/
-
-```{r, echo=FALSE, fig.align='center', out.width='30%'}
-knitr::include_graphics("images/book.jpg")
-```
+</div>

### 2019-01-08 code/prepare-kaggle-data.R

@@ -1,7 +1,8 @@
 ## Prepare survey data
 library('dplyr')
 library('mlr')
-data.dir = 'data/'
+
+data.dir = '../data/'
 
 survey.raw = read.csv(sprintf('%smultipleChoiceResponses.csv', data.dir), na.strings = c('NA', '-1', '-99', ''))
 exchange = read.csv(sprintf('%sconversionRates.csv', data.dir), row.names = 1)

### 2019-01-08 chunks/story-wines.Rmd
starts contructing wine analysis

TODO: Add wine csv and R script

##  {.emphasizedabit}
A startup wants you to predict wine quality from its chemical composition.

# Let's predict wine quality

# What is Machine Learning?

<div class="notes">
Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
</div>

## {data-background=../images/magic.jpg data-background-size=contain}

## Step 1: Find data

Red and white variants of the Portuguese "Vinho Verde" wine.

Target: Quality from 1 to 10
Features: Acidity, alcohol ...

<font size="2">P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>

## Step 1: Find data

TODO: Draw image with features.

## Step 1: Find data

TODO: Show distribution of target

```{r load-data,include=FALSE}
library('mlr')
library('ggplot2')
library('tidyr')
library('lime')
source('../code/prepare-wine-data.R')
```

# Step 2: Throw ML on your data

TODO: Benchmark some models, report error

```{r learn}
library("mlr")
lrn.ranger = makeLearner("regr.ranger")
lrn.lm = makeLearner("regr.lm")
lrn.rpart = makeLearner("regr.rpart")

rdesc = makeResampleDesc("Holdout")

lrns = list(lrn.ranger, lrn.lm, lrn.rpart)

tsk = makeRegrTask(data = wine, target = "quality")

bmr = benchmark(lrns, tsk, rdesc)

mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
preds = predict(mod, task = tsk, subset = seq(2, n, 2))
performance(preds, measures = list(mae, mse))
pred  = Predictor$new(mod, data = wdat, y = "quality")

n_segs(pred)
ale_fanova(pred)

plot_all_ale = function(pred) {
  pls = lapply(setdiff(colnames(wdat), c("quality")), function(fname) {
    fe = FeatureEffect$new(pred, fname, method = "ale")
    res = fe$results
    colnames(res)[colnames(res) == fname] = ".feature"
    res$.feature.name = fname
    res$.feature = as.numeric(res$.feature)
    res$.feature.type = fe$feature.type
    res
  })

  res  = rbindlist(pls, use.names = TRUE)
  res$.feature = as.numeric(res$.feature)

  ggplot(res, aes(x = as.numeric(.feature), y= .ale)) +
    geom_line(aes(alpha = .feature.type)) +
    geom_point() +
    facet_wrap(".feature.name", scales = "free_x") +
    scale_alpha_discrete(guide = "none")
}

plot_all_ale(pred)

fi = FeatureImp$new(pred, loss = "mae", n.repetitions = 10)
fi$plot()

x.interest = wdat[20,]
sh = Shapley$new(pred, x.interest = x.interest)
plot(sh)

## Simpler model

lrn = makeLearner("regr.rpart", maxdepth = 2)
tsk = makeRegrTask(data = wdat, target = "quality")
mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
preds = predict(mod, task = tsk, subset = seq(2, n, 2))
performance(preds, measures = list(mae, mse))
pred.tree  = Predictor$new(mod, data = wdat, y = "quality")

n_segs(pred.tree)
ale_fanova(pred.tree)

plot_all_ale(pred.tree)

mod$learner.model

fi = FeatureImp$new(pred, loss = "mae", n.repetitions = 10)
plot(fi)

plot_segs_feature(pred, feature.name = "alcohol", cat.mode = FALSE)
plot_segs_feature(pred, feature.name = "volatile.acidity", cat.mode = FALSE)
n_segs_feature(pred, "alcohol", cat.mode = TRUE)
n_segs_feature(pred, "sulphates", cat.mode = FALSE)
sh = Shapley$new(pred, x.interest = x.interest)
plot(sh)
li = LocalModel$new(pred, x.interest = x.interest)
plot(li)

lrn = makeLearner("regr.lm")
tsk = makeRegrTask(data = wdat, target = "quality")
mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
preds = predict(mod, task = tsk, subset = seq(2, n, 2))
performance(preds, measures = list(mae, mse))
pred  = Predictor$new(mod, data = wdat, y = "quality")

library('mlr')
set.seed(42)
task = makeRegrTask(data = wine, target = 'quality')
lrn = makeLearner('regr.randomForest')
mod = train(lrn, task)
```

## {.center data-background=../images/comp-dog.gif data-background-size=contain}

<div class="notes">
- Random Forest
- Target: Compensation
- All features that you found in the data
- Train, test, ship
</div>

## Step 2: Throw ML on your data

TODO: Benchmark some models, report error

Test: linear model, BRL, xgboost
Show table

## Step 2: Throw ML on your data

TODO: Benchmark some models, report error

Test: linear model, BRL, xgboost
Show table

## Step 2: Throw ML on your data

TODO: Show correlation between  predictions and true for test data

## {data-background=../images/delivery.gif data-background-size=contain}

# Step 3: Profit {.center}

## {data-background=../images/done-here.gif data-background-size=contain}
```{r, echo=FALSE, out.width='80%', include = FALSE}
knitr::include_graphics("../images/done-here.gif" )
```

##  {.emphasizedabit}

Client: "We would love to learn some insights."

##  {.emphasizedabit .center data-background=../images/black-box.gif data-background-size=cover}

<div class="white">
Looking inside the black box
</div>

## What are the most important features?

TODO: Feature importances

## How do features affect predictions?

TODO: All effects

## How do features affect predictions?

TODO: Effect of alcohol

## Customers want to have individual explanations

TODO: Image where customers checks wine?

## Shapley

TODO: Explain with Shapley

## LIME

TODO: Explain with LIME

### 2019-01-08 chunks/tools.Rmd
starts contructing wine analysis

@@
 
-IF $90m^2\
+IF $90m^2\leq \text{size} < 110m^2$  AND location $=$ "good" THEN rent is between 1540 and 1890 EUR
 
@@
 ```
+
+## Model-specific methods
+
+TODO: Example for CNNs
+
+## Model-specific methods
+
+TODO: Example for text (RNNs and attention?)
 
@@
 
-## Example-based methods
+## Model-agnostic Methods
+
+TODO: PDP gif
+
+## Model-agnostic methods
+
+TODO: Feature importance figure
+
+## Model-agnostic methods: Global Surrogate {.center data-background=../images/global-surrogate.png data-background-size=contain}
+
+## Model-agnostic methods: Local Surrogate {.center data-background=../images/local-surrogate.png data-background-size=contain}
+
+## Example-based Methods
 
@@
 
-## Example-focused methods
+## Example-focused Methods
 ```{r, echo=FALSE}
@@
 </div>
+
+## Example-focused Methods
+
+TODO: Graphic for counterfactuals
+
+## Example-focused Methods
+
+TODO: Graphic for prototypes

### 2019-01-08 code/prepare-wine-data.R

@@ -1,0 +1,7 @@
+wine = read.csv(file.path("../data", "wine.csv"))
+# Kicking out 36 wines with missing values
+wine = na.omit(wine)
+# Not sure if there is some order to it
+set.seed(42)
+wine = wine[sample(1:nrow(wine)),]
+n = nrow(wine)

### 2019-01-09 chunks/story-wines.Rmd
continues analysis

@@
 
+1 - fixed acidity
+2 - volatile acidity
+3 - citric acid
+4 - residual sugar
+5 - chlorides
+6 - free sulfur dioxide
+7 - total sulfur dioxide
+8 - density
+9 - pH
+10 - sulphates
+11 - alcohol
+Output variable (based on sensory data):
+12 - quality (score between 0 and 10)
+
+Draw:
+Volatile acidity as polish remover, pH as pH paper, alchohol as liqour bottle, chlorides as salt prinkler, citric acid as lemon, 
+sulfur as mushroom?,
+
 ## Step 1: Find data
-
-TODO: Show distribution of target
 
@@
 library('lime')
+devtools::install_github("christophM/iml", ref = "feature-effects")
+library('iml')
 source('../code/prepare-wine-data.R')
+```
+
+```{r show-dist}
+ggplot(wine) + 
+  geom_bar(aes(x = quality)) + 
+  scale_x_continuous("Wine Quality", labels = 1:10, breaks = 1:10)
 ```
@@
 
-TODO: Benchmark some models, report error
+##
 
-```{r learn}
+10x CV
+Test linear regression model, decision tree, random forest
+
+## {.center data-background=../images/comp-dog.gif data-background-size=contain}
+
+## Step 2: Throw ML on your data
+
+```{r benchmark}
 library("mlr")
@@
 
-rdesc = makeResampleDesc("Holdout")
+rdesc = cv5
 
@@
 
-bmr = benchmark(lrns, tsk, rdesc)
-
-mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
-preds = predict(mod, task = tsk, subset = seq(2, n, 2))
-performance(preds, measures = list(mae, mse))
-pred  = Predictor$new(mod, data = wdat, y = "quality")
-
-n_segs(pred)
-ale_fanova(pred)
-
-plot_all_ale = function(pred) {
-  pls = lapply(setdiff(colnames(wdat), c("quality")), function(fname) {
-    fe = FeatureEffect$new(pred, fname, method = "ale")
-    res = fe$results
-    colnames(res)[colnames(res) == fname] = ".feature"
-    res$.feature.name = fname
-    res$.feature = as.numeric(res$.feature)
-    res$.feature.type = fe$feature.type
-    res
-  })
-
-  res  = rbindlist(pls, use.names = TRUE)
-  res$.feature = as.numeric(res$.feature)
-
-  ggplot(res, aes(x = as.numeric(.feature), y= .ale)) +
-    geom_line(aes(alpha = .feature.type)) +
-    geom_point() +
-    facet_wrap(".feature.name", scales = "free_x") +
-    scale_alpha_discrete(guide = "none")
-}
-
-plot_all_ale(pred)
-
-fi = FeatureImp$new(pred, loss = "mae", n.repetitions = 10)
-fi$plot()
-
-x.interest = wdat[20,]
-sh = Shapley$new(pred, x.interest = x.interest)
-plot(sh)
-
-## Simpler model
-
-lrn = makeLearner("regr.rpart", maxdepth = 2)
-tsk = makeRegrTask(data = wdat, target = "quality")
-mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
-preds = predict(mod, task = tsk, subset = seq(2, n, 2))
-performance(preds, measures = list(mae, mse))
-pred.tree  = Predictor$new(mod, data = wdat, y = "quality")
-
-n_segs(pred.tree)
-ale_fanova(pred.tree)
-
-plot_all_ale(pred.tree)
-
-mod$learner.model
-
-fi = FeatureImp$new(pred, loss = "mae", n.repetitions = 10)
-plot(fi)
-
-plot_segs_feature(pred, feature.name = "alcohol", cat.mode = FALSE)
-plot_segs_feature(pred, feature.name = "volatile.acidity", cat.mode = FALSE)
-n_segs_feature(pred, "alcohol", cat.mode = TRUE)
-n_segs_feature(pred, "sulphates", cat.mode = FALSE)
-sh = Shapley$new(pred, x.interest = x.interest)
-plot(sh)
-li = LocalModel$new(pred, x.interest = x.interest)
-plot(li)
-
-lrn = makeLearner("regr.lm")
-tsk = makeRegrTask(data = wdat, target = "quality")
-mod = mlr::train(lrn, tsk,  subset = seq(1, n, 2))
-preds = predict(mod, task = tsk, subset = seq(2, n, 2))
-performance(preds, measures = list(mae, mse))
-pred  = Predictor$new(mod, data = wdat, y = "quality")
-
-library('mlr')
-set.seed(42)
-task = makeRegrTask(data = wine, target = 'quality')
-lrn = makeLearner('regr.randomForest')
-mod = train(lrn, task)
+bmr = benchmark(lrns, tsk, rdesc, measures = list(mae))
+bmr_tab = getBMRAggrPerformances(bmr, as.df = TRUE)
+knitr::kable(bmr_tab[-1])
 ```
 
-## {.center data-background=../images/comp-dog.gif data-background-size=contain}
+=> The random forest (ranger) is the best model.
 
-<div class="notes">
-- Random Forest
-- Target: Compensation
-- All features that you found in the data
-- Train, test, ship
-</div>
+```{r final-model}
+mod = train(lrn.ranger, tsk)
+
+set.seed(42)
+sample_size = 500
+wine_subset = wine[sample(1:nrow(wine), size = sample_size),]
+
+pred = Predictor$new(mod, data = wine_subset, y = "quality")
+```
 
@@
 
-TODO: Benchmark some models, report error
+```{r check-model}
 
-Test: linear model, BRL, xgboost
-Show table
-
-## Step 2: Throw ML on your data
-
-TODO: Benchmark some models, report error
-
-Test: linear model, BRL, xgboost
-Show table
-
-## Step 2: Throw ML on your data
-
-TODO: Show correlation between  predictions and true for test data
+preds = pred$predict(wine)
+preds$actual = wine$quality
+ggplot(preds, aes(x = actual, y = .prediction, group = actual)) + 
+  geom_violin(aes(fill = ..n..)) +
+  scale_x_continuous("Actual quality (jittered)", 
+    labels = 1:10, breaks = 1:10) + 
+  scale_y_continuous("Predicted quality", labels = 1:10, breaks = 1:10) + 
+  scale_fill_gradient(low = "white", high = "darkblue", guide = "none")
+```
 
@@
 
-TODO: Feature importances
+```{r feature-importance}
+imp = FeatureImp$new(pred, loss = "mae")
+plot(imp)
+```
 
@@
 
-TODO: All effects
+```{r feature-effects}
+effs = FeatureEffects$new(pred)
+plot(effs, ncols = 4)
+```
 
@@
 
-TODO: Effect of alcohol
+```{r feature-effect-alcohol}
+eff = FeatureEffect$new(pred, "alcohol")
+plot(eff)
+```
+
+## Rule of thumb what makes wine good?
+
+```{r surrogate}
+sbrl()
+tree = TreeSurrogate$new(pred, maxdepth  = 2)
+plot(tree)
+```
 
@@
 
-## Shapley
+## Exceptionally bad wine
 
-TODO: Explain with Shapley
+```{r bad-wine}
+# find bad wine in data
+predictions = pred$predict(wine)
+
+min_pred = min(predictions)
+worst = wine[which(predictions == min_pred),]
+
+# explain with shapley
+knitr::kable(t(worst))
+# explain with cf?
+
+min_pred
+
+```
+
+## Shapley Value
+
+TODO: Draw features as players and within coalitions
+
+## Shapley Value
+
+```{r shapley}
+shap = Shapley$new(pred, x.interest = worst)
+plot(shap)
+```
+
+## Volatile Acidity
+
+```{r volatile-acidity}
+eff = FeatureEffect$new(pred, "volatile.acidity", method = "pdp+ice")
+plot(eff)
+```
+## Volatile Acidity
+
+‘Although the presence of high amounts of VA is considered undesirable, in some cases a touch of volatility is no bad thing,’ said Natasha Hughes MW in her guide to wine ‘flaws’.
+
+```{r volatile-acidity}
+
+# Get wines with low volatile acidity
+predictions = pred$predict(wine_subset)
+
+wine2 = wine_subset
+wine2$volatile.acidity = wine2$volatile.acidity + 0.2
+predictions2 = pred$predict(wine2)
+
+plot(hist(predictions2$.prediction - predictions$.prediction))
+
+wine_improve = which(predictions2$.prediction - predictions$.prediction > 0.3)
+
+eff = FeatureEffect$new(pred, "volatile.acidity", method = "ice")
+
+res = eff$results
+res$improves = res$.id %in% wine_improve
+
+ggplot(res) + 
+  geom_line(aes(x = volatile.acidity, y = .y.hat, group = .id), alpha = 0.1) + 
+  facet_wrap("improves")
+```
 
@@
 TODO: Explain with LIME
+
+## Sulfur
+
+At low levels reduction causes truffle, mushroom, radish, and green olive-like smells to wine.
+These have been accepted as positive traits and increase complexity in the wine.
+At higher levels, reduction is turns into a fault, giving wines a pungent smell of rotten egg, onion, garlic, cooked cabbage, canned corn, or even burnt rubber.
+
+```{r sulfur}
+```

### 2019-01-10 chunks/story-wines.Rmd
continues talk

@@
-TODO: Add wine csv and R script
-
 ##  {.emphasizedabit}
@@
 
-# What is Machine Learning?
-
-<div class="notes">
-Machine learning is a method for teaching computers to make and improve predictions or behaviours based on data.
-</div>
-
-## {data-background=../images/magic.jpg data-background-size=contain}
-
 ## Step 1: Find data
 
-Red and white variants of the Portuguese "Vinho Verde" wine.
+Free online dataset about red and white variants of the Portuguese "Vinho Verde" wine.
 
@@
 
-TODO: Draw image with features.
+- fixed acidity
+- volatile acidity
+- citric acid
+- residual sugar
+- chlorides
+- free sulfur dioxide
+- total sulfur dioxide
+- density
+- pH
+- sulphates
+- alcohol
+- quality based on sensory data (0-10)
 
-1 - fixed acidity
-2 - volatile acidity
-3 - citric acid
-4 - residual sugar
-5 - chlorides
-6 - free sulfur dioxide
-7 - total sulfur dioxide
-8 - density
-9 - pH
-10 - sulphates
-11 - alcohol
-Output variable (based on sensory data):
-12 - quality (score between 0 and 10)
-
-Draw:
-Volatile acidity as polish remover, pH as pH paper, alchohol as liqour bottle, chlorides as salt prinkler, citric acid as lemon, 
-sulfur as mushroom?,
+<!-- Draw: -->
+<!-- Volatile acidity as polish remover, pH as pH paper, alchohol as liqour bottle, chlorides as salt prinkler, citric acid as lemon,  -->
+<!-- sulfur as mushroom?,  -->
 
@@
 
-```{r feature-effects}
+```{r}
 effs = FeatureEffects$new(pred)
@@
 ## How do features affect predictions?
-
-```{r feature-effect-alcohol}
-eff = FeatureEffect$new(pred, "alcohol")
+Method: Partial dependence plot, individual conditional expectation curves
+```{r}
+eff = FeatureEffect$new(pred, "alcohol", method = 'pdp+ice')
 plot(eff)
@@
 
-## Rule of thumb what makes wine good?
+## Interactions with type of wine?
+Method: H-statistic for interactions
+
+```{r}
+inter = Interaction$new(pred)
+plot(inter)
+```
+
+## Interactions of alcohol?
+Method: H-statistic for interactions
+
+```{r}
+inter.a = Interaction$new(pred, "alcohol")
+plot(inter.a)
+```
+
+## Rule of thumb for wine quality?
+Method: Surrogate Model
 
 ```{r surrogate}
-sbrl()
 tree = TreeSurrogate$new(pred, maxdepth  = 2)
-plot(tree)
+plot(tree$tree)
 ```
 
-## Customers want to have individual explanations
+## Exceptionally bad wine
 
-TODO: Image where customers checks wine?
-
-## Exceptionally bad wine
+A customer gets a really bad prediction. What was the reason?
 
@@
 
-TODO: Draw features as players and within coalitions
-
-## Shapley Value
-
-```{r shapley}
+```{r}
 shap = Shapley$new(pred, x.interest = worst)
@@
 
-## Volatile Acidity
+## The best wine
 
-```{r volatile-acidity}
-eff = FeatureEffect$new(pred, "volatile.acidity", method = "pdp+ice")
-plot(eff)
-```
-## Volatile Acidity
+```{r best-wine}
+# find bad wine in data
+max_pred = max(predictions)
+best = wine[which(predictions == max_pred),]
 
-‘Although the presence of high amounts of VA is considered undesirable, in some cases a touch of volatility is no bad thing,’ said Natasha Hughes MW in her guide to wine ‘flaws’.
+# explain with shapley
+knitr::kable(t(best))
+# explain with cf?
 
-```{r volatile-acidity}
+max_pred
 
-# Get wines with low volatile acidity
-predictions = pred$predict(wine_subset)
-
-wine2 = wine_subset
-wine2$volatile.acidity = wine2$volatile.acidity + 0.2
-predictions2 = pred$predict(wine2)
-
-plot(hist(predictions2$.prediction - predictions$.prediction))
-
-wine_improve = which(predictions2$.prediction - predictions$.prediction > 0.3)
-
-eff = FeatureEffect$new(pred, "volatile.acidity", method = "ice")
-
-res = eff$results
-res$improves = res$.id %in% wine_improve
-
-ggplot(res) + 
-  geom_line(aes(x = volatile.acidity, y = .y.hat, group = .id), alpha = 0.1) + 
-  facet_wrap("improves")
 ```
 
-## LIME
+## The best wine
+Technique: Shapley Values
 
-TODO: Explain with LIME
-
-## Sulfur
-
-At low levels reduction causes truffle, mushroom, radish, and green olive-like smells to wine.
-These have been accepted as positive traits and increase complexity in the wine.
-At higher levels, reduction is turns into a fault, giving wines a pungent smell of rotten egg, onion, garlic, cooked cabbage, canned corn, or even burnt rubber.
-
-```{r sulfur}
+```{r}
+shap = Shapley$new(pred, x.interest = best)
+plot(shap)
 ```

### 2019-01-14 chunks/story-wines.Rmd
continues slides

@@
 ##  {.emphasizedabit}
-A startup wants you to predict wine quality from its chemical composition.
+You want to predict wine quality from its physicochemical properties.
 
-# Let's predict wine quality
+## {.center data-background=../images/wine-features.jpg data-background-size=contain}
 
 ## Step 1: Find data
+Free dataset of red and white "Vinho Verde" wine.
 
-Free online dataset about red and white variants of the Portuguese "Vinho Verde" wine.
+```{r}
+knitr::include_graphics("../images/minho.png")
+```
 
-Target: Quality from 1 to 10
-Features: Acidity, alcohol ...
+Three sensory assessors (using blind tastes), which graded the wine in a scale that ranges from
+0 (very bad) to 10 (excellent)
 
@@
 Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>
-
-## Step 1: Find data
-
-- fixed acidity
-- volatile acidity
-- citric acid
-- residual sugar
-- chlorides
-- free sulfur dioxide
-- total sulfur dioxide
-- density
-- pH
-- sulphates
-- alcohol
-- quality based on sensory data (0-10)
 
@@
 
-10x CV
-Test linear regression model, decision tree, random forest
+Compare different models with 10x CV
+- Linear regression model
+- Decision tree
+- Random forest
 
@@
   geom_violin(aes(fill = ..n..)) +
-  scale_x_continuous("Actual quality (jittered)", 
+  scale_x_continuous("Actual quality", 
     labels = 1:10, breaks = 1:10) + 
@@
 
+TODO: Slide explaining permutation feature importance
+
+## What are the most important features?
+
 ```{r feature-importance}
@@
 plot(imp)
+```
+
+## How do features affect predictions?
+
+TODO: Slide explaining ALE plots
+
+## How do features affect predictions?
+Method: Accumulated Local Effects
+```{r}
+eff = FeatureEffect$new(pred, "alcohol", method = 'ale')
+plot(eff)
+```
+
+## How do features affect predictions?
+Method: Accumulated Local Effects
+```{r}
+eff = FeatureEffect$new(pred, "volatile.acidity", method = 'ale')
+plot(eff)
 ```
@@
 
-## How do features affect predictions?
-Method: Partial dependence plot, individual conditional expectation curves
-```{r}
-eff = FeatureEffect$new(pred, "alcohol", method = 'pdp+ice')
-plot(eff)
-```
-
-## Interactions with type of wine?
-Method: H-statistic for interactions
+## Interactions between alcohol and volatile acidity?
 
 ```{r}
-inter = Interaction$new(pred)
+inter = FeatureEffect$new(pred, feature = c("alcohol", "volatile.acidity"))
 plot(inter)
@@
 
-## Interactions of alcohol?
-Method: H-statistic for interactions
+## Rule of thumb for wine quality?
 
-```{r}
-inter.a = Interaction$new(pred, "alcohol")
-plot(inter.a)
+```{r surrogate}
+library(partykit)
+tree = TreeSurrogate$new(pred, maxdepth  = 2)
+node_inner2 = node_inner(tree$tree, pval = FALSE, id=FALSE)
+node_terminal2 = node_boxplot(tree$tree, id = FALSE)
+plot(tree$tree, inner_panel = node_inner2, terminal_panel = node_terminal2)
+
+rsquared = tree$r.squared
 ```
 
-## Rule of thumb for wine quality?
-Method: Surrogate Model
-
-```{r surrogate}
-tree = TreeSurrogate$new(pred, maxdepth  = 2)
-plot(tree$tree)
-```
+The tree explains `r sprintf("%.2f%s", 100 *  rsquared, "%")` of the black box prediction variance.
 
 ## Exceptionally bad wine
-
-A customer gets a really bad prediction. What was the reason?
 
@@
 worst = wine[which(predictions == min_pred),]
+```
 
-# explain with shapley
-knitr::kable(t(worst))
-# explain with cf?
+TODO: Image of really bad wine
 
-min_pred
+Predicted quality: `r predictions[which(predictions == min_pred),]`
 
-```
+## Shapley Value
+
+TODO: Slide to explain Shapley value
 
@@
 
-## The best wine
+## What needs to change?
 
-```{r best-wine}
-# find bad wine in data
-max_pred = max(predictions)
-best = wine[which(predictions == max_pred),]
+TODO: Slide to explain counterfactuals
 
-# explain with shapley
-knitr::kable(t(best))
-# explain with cf?
+## Counterfactual explanation
 
-max_pred
+How do we get the wine above predicted quality of 5?
 
+```{r}
+worst
+
+worst2 = function(){
+  w = worst
+  w$volatile.acidity = 0.2
+  w
+}
+
+pred$predict(worst2())[[1]]
+
+worst3 = function(){
+  w = worst
+  w$alcohol = 13
+  w$volatile.acidity = 1
+  w
+}
+
+pred$predict(worst3())[[1]]
 ```
 
-## The best wine
-Technique: Shapley Values
+TODO: Image
 
-```{r}
-shap = Shapley$new(pred, x.interest = best)
-plot(shap)
-```
+- Decreasing volatile acidity to 0.2 yields predicted quality of `r sprintf('%.2f', pred$predict(worst2())[[1]])`
+- Decreasing volatile acidity to 1.0 and increasing alcohol to 13% yields predicted quality of `r sprintf('%.2f', pred$predict(worst3())[[1]])`

### 2019-01-14 chunks/tools.Rmd
continues slides

@@
 </div>
+
+##
+
+TODO: Overview (drawing)
+
+- global vs. local
+- interpretable model vs. post-hoc
+- model-specific vs. model-agnostic
+- data type image vs. text vs. tabular
 
@@
 
-TODO: PDP gif
+TODO: Drawing of feature effect
 
@@
 
-TODO: Feature importance figure
+TODO: Drawing of importance
 
@@
 ## Model-agnostic methods: Local Surrogate {.center data-background=../images/local-surrogate.png data-background-size=contain}
-
-## Example-based Methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("../images/black-box.png")
-```
-
-## Example-focused Methods
-```{r, echo=FALSE}
-knitr::include_graphics("../images/data-box.png")
-```
-
-<div class="notes">
-- Prototypes: What is the prototype of a certain class
-- Criticisms: Which data points are unusual for a certain class. 
-- Counterfactuals: What do I need to change for an instance to change the prediction?
-- Anchors: What features do I have to anchor to keep prediction the same?
-</div>
 

### 2019-01-14 chunks/story-wines.Rmd
continues talk

@@
 ##  {.emphasizedabit}
-You want to predict wine quality from its physicochemical properties.
+Predict wine quality from its physicochemical properties.
 
@@
 
-## Step 1: Find data
-Free dataset of red and white "Vinho Verde" wine.
+# Step 1: Find data
 
-```{r}
-knitr::include_graphics("../images/minho.png")
-```
+## {.center data-background=../images/minho.png data-background-size=contain}
 
-Three sensory assessors (using blind tastes), which graded the wine in a scale that ranges from
-0 (very bad) to 10 (excellent)
+## Wine Dataset
+
+- ~6500 red and white Portuguese "Vinho Verde" wines
+- Features: physiochemical properties
+- Quality assessed by blind tasting, from 0 (very bad) to 10 (excellent)
 
@@
 
-<!-- Draw: -->
-<!-- Volatile acidity as polish remover, pH as pH paper, alchohol as liqour bottle, chlorides as salt prinkler, citric acid as lemon,  -->
-<!-- sulfur as mushroom?,  -->
-
-## Step 1: Find data
+## Quality Distribution
 
@@
 
-# Step 2: Throw ML on your data
+# Step 2: Apply Machine Learning
 
-##
+## Compare Models with 10x CV {.emphasizedabit}
 
-Compare different models with 10x CV
-- Linear regression model
+- Linear regression
 - Decision tree
@@
 ## {.center data-background=../images/comp-dog.gif data-background-size=contain}
-
-## Step 2: Throw ML on your data
 
@@
 bmr_tab = getBMRAggrPerformances(bmr, as.df = TRUE)
-knitr::kable(bmr_tab[-1])
+#knitr::kable(bmr_tab[-1])
 ```
-
-=> The random forest (ranger) is the best model.
 
@@
 
-## Step 2: Throw ML on your data
+## Compare Models with 10x CV {.emphasizedabit}
+
+- Linear regression (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.lm"])`)
+- Decision tree (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.rpart"])`)
+- Random forest (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`)
+
+=> The random forest is the best model.
+
+## Prediction vs. Actual Quality
 
@@
 
-## {data-background=../images/delivery.gif data-background-size=contain}
-
 # Step 3: Profit {.center}
@@
 
-##  {.emphasizedabit}
-
-Client: "We would love to learn some insights."
+##  {.emphasizedabit .center data-background=../images/no-idea.gif data-background-size=cover}
 
@@
 
-TODO: Slide explaining permutation feature importance
+## What are the most important features?
+
+```{r}
+knitr::include_graphics("../images/permimp-shuffle.png")
+```
 
@@
 
-TODO: Slide explaining ALE plots
+## How do features affect predictions?
+
+```{r}
+knitr::include_graphics("../images/pdp-rent.png")
+```s
 
 ## How do features affect predictions?
-Method: Accumulated Local Effects
 ```{r}
@@
 ## How do features affect predictions?
-Method: Accumulated Local Effects
 ```{r}
@@
 
-## Interactions between alcohol and volatile acidity?
+## Interaction: Alcohol + volatile acidity
 
@@
 
+## Rule of thumb for wine quality? {.center data-background=../images/global-surrogate.png data-background-size=contain}
+
 ## Rule of thumb for wine quality? 
-
 ```{r surrogate}
@@
 
-The tree explains `r sprintf("%.2f%s", 100 *  rsquared, "%")` of the black box prediction variance.
+Tree explains `r sprintf("%.2f%s", 100 *  rsquared, "%")` of black box prediction variance.
 
-## Exceptionally bad wine
+## Explain individual prediction (e.g. worst wine)
 
@@
 
-TODO: Image of really bad wine
-
-Predicted quality: `r predictions[which(predictions == min_pred),]`
+##  {.center data-background=../images/bad-wine.gif data-background-size=contain}
 
@@
 
-pred$predict(worst2())[[1]]
-
 worst3 = function(){
@@
 }
-
-pred$predict(worst3())[[1]]
 ```

### 2019-01-14 chunks/tools.Rmd
continues talk

@@
 
-## Model-specific methods
+## Model-specific Methods
 
@@
 
-## Model-agnostic methods
+## Model-agnostic Methods
 
@@
 
-<div class="notes">
-- Partial dependence plots
-- LIME
-- ...
-</div>
-
 ## Model-agnostic Methods
@@
 
-## Model-agnostic methods: Global Surrogate {.center data-background=../images/global-surrogate.png data-background-size=contain}
+## Model-agnostic Methods: Global Surrogate {.center data-background=../images/global-surrogate.png data-background-size=contain}
 
-## Model-agnostic methods: Local Surrogate {.center data-background=../images/local-surrogate.png data-background-size=contain}
+## Model-agnostic Methods: Local Surrogate {.center data-background=../images/local-surrogate.png data-background-size=contain}
 

### 2019-01-15 chunks/story-wines.Rmd
continues slides

@@
 Predict wine quality from its physicochemical properties.
-
-## {.center data-background=../images/wine-features.jpg data-background-size=contain}
 
@@
 - ~6500 red and white Portuguese "Vinho Verde" wines
-- Features: physiochemical properties
+- Features: Physiochemical properties
 - Quality assessed by blind tasting, from 0 (very bad) to 10 (excellent)
@@
 Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>
+
+## {.center data-background=../images/wine-features.jpg data-background-size=contain}
 
@@
 
-=> The random forest is the best model.
+=> Random forest has lowest Mean Absolute Error
 
@@
 
-## {data-background=../images/done-here.gif data-background-size=contain}
-```{r, echo=FALSE, out.width='80%', include = FALSE}
-knitr::include_graphics("../images/done-here.gif" )
-```
+## {data-background=../images/dog-car.gif data-background-size=contain}
 
@@
 
-## What are the most important features?
+## Which features are important?
 
-## What are the most important features?
+## Which features are important?
 
@@
 
-## What are the most important features?
+## Which features are important?
 
@@
 ## How do features affect predictions?
-
 ```{r}
-knitr::include_graphics("../images/pdp-rent.png")
-```s
+knitr::include_graphics("../images/ale-intuition-1.jpg")
+```
 
 ## How do features affect predictions?
+```{r}
+knitr::include_graphics("../images/ale-intuition-2.jpg")
+```
+
+## How do features affect predictions?
+```{r}
+knitr::include_graphics("../images/ale-intuition-3.jpg")
+```
+
+## How do features affect predictions?
+```{r}
+knitr::include_graphics("../images/ale-intuition-4.jpg")
+```
+
+## How do features affect predictions?
+```{r}
+knitr::include_graphics("../images/ale-intuition-5.jpg")
+```
+
+## Effect of Alchohol
 ```{r}
@@
 
-## How do features affect predictions?
+## Effect of Volatile Aciditys
 ```{r}
@@
 inter = FeatureEffect$new(pred, feature = c("alcohol", "volatile.acidity"))
-plot(inter)
+plot(inter, show.data = TRUE)
 ```
+
+## Rule of thumb for wine quality?
 
@@
 shap = Shapley$new(pred, x.interest = worst)
-plot(shap)
+plot(shap) + 
+  scale_y_discrete("") + 
+  scale_x_continuous("Feature contribution")
 ```
@@
 ```{r}
-worst
-
 worst2 = function(){

### 2019-01-15 chunks/tools.Rmd
continues slides

@@
 # What tools do we have?
-
-<div class="notes">
-- Intrinsically interpretable models
-- Tools for analyzing specific black box models
-- Model-agnostic tools: analysing any black box model
-- Interpretability through making data explainable
-</div>
-
-##
-
-TODO: Overview (drawing)
-
-- global vs. local
-- interpretable model vs. post-hoc
-- model-specific vs. model-agnostic
-- data type image vs. text vs. tabular
 
@@
 ## Model-specific Methods
+Layerwise Relevance Propagation (LRP)
 
-TODO: Example for CNNs
+```{r, echo=FALSE}
+knitr::include_graphics("../images/lrp.png")
+```
+
+<font size="2">
+Bach, Sebastian, et al. "On pixel-wise explanations for non-linear classifier decisions by layer-wise relevance propagation." PloS one 10.7 (2015): e0130140.
+</font>
 
@@
 
-TODO: Example for text (RNNs and attention?)
+```{r, echo=FALSE}
+knitr::include_graphics("../images/cnn-feature-viz.png")
+```
 
-<div class="notes">
-- Visualizing activations of neural networks
-- Random Forest specific feature importance
-- xgboost explainer
-- problem: new models, not comparable between models, missing flexibility
-</div>
+<font size="2">
+https://distill.pub/2017/feature-visualization/
+</font>
 
@@
 
-TODO: Drawing of feature effect
-
-## Model-agnostic Methods
-
-TODO: Drawing of importance
+```{r}
+knitr::include_graphics("../images/pdp-rent.png")
+```
 
@@
 TODO: Graphic for counterfactuals
-
-## Example-focused Methods
-
-TODO: Graphic for prototypes

### 2019-01-15 chunks/story-wines.Rmd
fixes plot
adds intuition for counterfactual explanations

@@
 plot(shap) + 
-  scale_y_discrete("") + 
-  scale_x_continuous("Feature contribution")
+  scale_x_discrete("") + 
+  scale_y_continuous("Feature contribution")
 ```
 
-## What needs to change?
+## Counterfactual Explanations  {.center data-background=../images/counterfactual-intuition-1.png data-background-size=contain}
 
-TODO: Slide to explain counterfactuals
+## Counterfactual Explanations  {.center data-background=../images/counterfactual-intuition-2.png data-background-size=contain}
+
+## Counterfactual Explanations  {.center data-background=../images/counterfactual-intuition-3.png data-background-size=contain}
 

### 2019-01-16 chunks/story-wines.Rmd
adds intuition shapley
polish slides

@@
 Predict wine quality from its physicochemical properties.
-
-# Step 1: Find data
 
@@
 
-- ~6500 red and white Portuguese "Vinho Verde" wines
-- Features: Physiochemical properties
+- 6500 red and white Portuguese "Vinho Verde" wines
+- Features: Physicochemical properties
 - Quality assessed by blind tasting, from 0 (very bad) to 10 (excellent)
 
-<font size="2">P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
+```{r, out.width= "30%"}
+knitr::include_graphics("../images/judges.png")
+```
+
+<font size="2">
+P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
 Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.</font>
@@
 
+## How can we solve this problem? {.emphasizedabit}
+
+- Hand-written rules
+- Statistical Model
+- Machine Learning
+
+```{r, child="../chunks/what-is-ml.Rmd"}
+```
+
+# Step 1: Find data
+
+## 
+```{r}
+knitr::include_graphics("../images/checkmark.png")
+```
+
 # Step 2: Apply Machine Learning
 
-## Compare Models with 10x CV {.emphasizedabit}
+## Random Forest
 
-- Linear regression
-- Decision tree
-- Random forest
+![](../images/random-forest.jpg)
+
+<font size="2">
+Image: 
+http://www.hallwaymathlete.com/2016/05/introduction-to-machine-learning-with.html
+</font>
 
@@
 
-## Compare Models with 10x CV {.emphasizedabit}
+## Train Random Forest to Predict Quality {.emphasizedabit}
 
-- Linear regression (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.lm"])`)
-- Decision tree (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.rpart"])`)
-- Random forest (`r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`)
-
-=> Random forest has lowest Mean Absolute Error
+Mean absolute error on test data (cross-validated): `r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`
 
@@
 ```{r check-model}
-
 preds = pred$predict(wine)
@@
 ggplot(preds, aes(x = actual, y = .prediction, group = actual)) + 
-  geom_violin(aes(fill = ..n..)) +
+  geom_abline(intercept = 0, slope = 1, lty = 2) +
+  geom_violin() +
   scale_x_continuous("Actual quality", 
-    labels = 1:10, breaks = 1:10) + 
-  scale_y_continuous("Predicted quality", labels = 1:10, breaks = 1:10) + 
-  scale_fill_gradient(low = "white", high = "darkblue", guide = "none")
+    labels = 1:10, breaks = 1:10, limit = c(2.5,9.5)) + 
+  scale_y_continuous("Predicted quality", labels = 1:10, breaks = 1:10, limits = c(2.5,9.5)) 
+#scale_fill_gradient(low = "white", high = "darkblue") + 
 ```
@@
 ##  {.emphasizedabit .center data-background=../images/no-idea.gif data-background-size=cover}
+
+## We want to know:
+
+- Which wine properties are the most predictive for quality?
+- How does a property affect the predicted wine quality?
+- Can we extract a "Rule of Thumb" from the black box?
+- Why did a wine get a certain prediction?
+- How do we have to change a wine to achieve a different prediction?
 
@@
 
-## Which features are important?
+## Permutation Feature Importance
 
@@
 
-## How do features affect predictions?
-```{r}
+## Accumulated Local Effects
+
+```{r,  out.width = '78%'}
 knitr::include_graphics("../images/ale-intuition-1.jpg")
@@
 
-## How do features affect predictions?
-```{r}
+## Accumulated Local Effects
+
+```{r,  out.width = '78%'}
 knitr::include_graphics("../images/ale-intuition-2.jpg")
@@
 
-## How do features affect predictions?
-```{r}
+## Accumulated Local Effects
+
+```{r,  out.width = '78%'}
 knitr::include_graphics("../images/ale-intuition-3.jpg")
@@
 
-## How do features affect predictions?
-```{r}
+## Accumulated Local Effects
+```{r,  out.width = '78%'}
 knitr::include_graphics("../images/ale-intuition-4.jpg")
@@
 
-## How do features affect predictions?
-```{r}
+## Accumulated Local Effects
+```{r,  out.width = '78%'}
 knitr::include_graphics("../images/ale-intuition-5.jpg")
@@
 
-## Effect of Alchohol
+## Effect of Alcohol
 ```{r}
@@
 
-## Effect of Volatile Aciditys
+## Effect of Volatile Acidity
 ```{r}
@@
 
-## Interaction: Alcohol + volatile acidity
-
-```{r}
-inter = FeatureEffect$new(pred, feature = c("alcohol", "volatile.acidity"))
-plot(inter, show.data = TRUE)
-```
-
 ## Rule of thumb for wine quality?
 
-## Rule of thumb for wine quality? {.center data-background=../images/global-surrogate.png data-background-size=contain}
+## Surrogate Model {.center data-background=../images/global-surrogate.png data-background-size=contain}
 
-## Rule of thumb for wine quality? 
+## Surrogate Model
 ```{r surrogate}
@@
 
-## Explain individual prediction (e.g. worst wine)
+## Explain individual predictions
 
@@
 min_pred = min(predictions)
+max_pred = max(predictions)
 worst = wine[which(predictions == min_pred),]
+best = wine[which(predictions == max_pred),]
+```
+
+## Shapley Value
+
+```{r}
+knitr::include_graphics("../images/shapley-intuition-1.png")
+```
+
+## Explain best wine
+
+```{r}
+shap = Shapley$new(pred, x.interest = best)
+plot(shap) + 
+  scale_x_discrete("") + 
+  scale_y_continuous("Feature contribution")
 ```
@@
 
-## Shapley Value
-
-TODO: Slide to explain Shapley value
-
-## Shapley Value
+## Explain worst wine
 
@@
 
+## Improve worst wine?
+
 ## Counterfactual Explanations  {.center data-background=../images/counterfactual-intuition-1.png data-background-size=contain}
@@
 
-## Counterfactual explanation
+## Improve worst wine?
 
@@
 
-TODO: Image
-
 - Decreasing volatile acidity to 0.2 yields predicted quality of `r sprintf('%.2f', pred$predict(worst2())[[1]])`

### 2019-01-16 chunks/what-is-ml.Rmd
polish slides

## Programing vs. Machine Learning  {.center data-background=../images/programing-ml.png data-background-size=contain}

## Statistics vs. Machine Learning {.emphasizedabit}

linear $\Rightarrow$ non-linear

additive $\Rightarrow$ interactions

theory-driven $\Rightarrow$ optimization-driven

## Machine Learning (supervised)

```{r, echo=FALSE}
knitr::include_graphics("../images/ml-box.png")
```

## Machine Learning (supervised)

```{r, echo=FALSE}
knitr::include_graphics("../images/magic-box.png")
```

## Machine Learning (supervised)

```{r, echo=FALSE}
knitr::include_graphics("../images/black-box.png")
```

### 2019-03-11 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
starts Dagstat slides

---
title: "Measuring and optimizing machine learning interpretability"
author: "Christoph Molnar"
date: "2019-03-21"
output: 
  ioslides_presentation:
    widescreen: false
    smaller: false
css: ../styles.css
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
  warning = FALSE,
  message = FALSE, 
  fig.align='center', 
  out.width = '90%',
  dev = "CairoPNG")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 15))
```

## Black Box problem

```{r, echo = FALSE}
knitr::include_graphics("../images/black-box.png")
```

## Solution 1: Interpretable Model

```{r, echo=FALSE}
knitr::include_graphics("../images/white-box.png")
```

## But: Peformance low and how to compare interpretability?

```{r}
knitr::include_graphics("../images/simple-models.png")
```

## Solution 2: Post-Hoc Analysis

```{r, echo=FALSE}
knitr::include_graphics("../images/agnostic-black-box.png")
```

## But: Verbose and unreliable {.center data-background=../images/pdp-multiple.png}

```{r, eval=FALSE}
library('iml')
dat = data.frame(matrix(rnorm(1800), ncol = 180))
pred.fun = function(newdata, object){
	data.frame(y = sin(rowSums(newdata)) + rnorm(nrow(newdata), sd = 0.1))
}
pred = Predictor$new(data = dat, predict.fun = pred.fun)
fes = FeatureEffects$new(pred)
fes$plot(ncol = 20)
```
```{r eval=FALSE}
knitr::include_graphics("../images/pdp-multiple.png")
```

## Solution 3: Compare models with interpretability / complexity measures

TODO: Image with 3 black boxes and numbers attached to it

## Desiderata

- Few features (Sparsity)
- Few interactions (Additivity)
- Simple relationships (Simplicity)

## Measure: Number of features heuristic

TODO: Visualisation of algorithm

## Functional Decomposition

TODO: Formula with decomposition

## Accumulated Local Effects (ALE)

```{r}
knitr::include_graphics("../images/ale-intuition-3.jpg")
```

## Functional Decomposition with ALE

TODO: copy paste new decomposition

## Measure: Interaction Strength

TODO: PRL formula with R2 and SSQ and so on

## Measure: Complexity

TODO: AleNumApprox plot

## Measure: Complexity cont.

TODO: Multiple AleNumApprox plots with averaging formula

## Application: Multi-Objective Optimization

TODO: Wine plot
TODO: List of models

## Application: Multi-Objective Optimization

TODO: Show result graphics

## Application: Multi-Objective Optimization

TODO: Show Interpretability vs. Performance Tradeoff

```{r child = "../chunks/book.Rmd"}
```

# Backup

## Units in Wine dataset

- fixed acidity g(tartaric acid)/dm^3^
- volatile acidity: g(acetric acid/dm^3^)
- citric acid: g/dm^3^
- residual sugar: g/dm^3^
- chlorides: g(sodium chloride)/dm^3^
- free sulfur dioxide: mg/dm^3^
- total sulfur dioxide: mg/dm^3^
- density> g/cm^3^
- pH
- sulphates: g(postassium sulphate) / dm^3^
- alcohol vol.%

### 2019-03-18 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
pdf; continue

@@
 title: "Measuring and optimizing machine learning interpretability"
-author: "Christoph Molnar"
+author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
 date: "2019-03-21"
 output: 
-  ioslides_presentation:
-    widescreen: false
-    smaller: false
-css: ../styles.css
+    beamer_presentation
 ---
@@
 ```
+## Problem: Tradeoff interpretability and performance
+Use only interpretable models (e.g. linear model)! $\Rightarrow$ Pay with loss in performance.
 
-## Solution 1: Interpretable Model
-
-```{r, echo=FALSE}
+```{r, echo = FALSE}
 knitr::include_graphics("../images/white-box.png")
 ```
-
-## But: Peformance low and how to compare interpretability?
-
-```{r}
-knitr::include_graphics("../images/simple-models.png")
-```
-
-## Solution 2: Post-Hoc Analysis
+## Problem: Tradeoff interpretability and performance
+Just use post-hoc interpretation methods (e.g. feature importance)! $\Rightarrow$ Works better for less complex models.
 
@@
 
-## But: Verbose and unreliable {.center data-background=../images/pdp-multiple.png}
+## What's interpretability anyway?
 
-```{r, eval=FALSE}
-library('iml')
-dat = data.frame(matrix(rnorm(1800), ncol = 180))
-pred.fun = function(newdata, object){
-	data.frame(y = sin(rowSums(newdata)) + rnorm(nrow(newdata), sd = 0.1))
-}
-pred = Predictor$new(data = dat, predict.fun = pred.fun)
-fes = FeatureEffects$new(pred)
-fes$plot(ncol = 20)
-```
-```{r eval=FALSE}
-knitr::include_graphics("../images/pdp-multiple.png")
-```
+TODO of fuzzy animal which dodges touch of hand
+Or use image of cat that doges hand
 
-## Solution 3: Compare models with interpretability / complexity measures
+## We propose
 
-TODO: Image with 3 black boxes and numbers attached to it
+TODO: Replace text with image: Black box in the middle, three properties attached to it
 
-## Desiderata
+Measure model complexity in a model-agnostic way: number of features, interaction strength, main effect complexity
 
-- Few features (Sparsity)
-- Few interactions (Additivity)
-- Simple relationships (Simplicity)
+$\Rightarrow$ Allows model comparision
 
-## Measure: Number of features heuristic
+$\Rightarrow$ Allows direct optimization for interpretability
 
-TODO: Visualisation of algorithm
+$\Rightarrow$ Makes claims of interpretability more explicit ("Model A uses less features than B and has less interactions)
 
@@
 
-TODO: Formula with decomposition
+$$f(x) = \overbrace{f_0}^\text{Intercept} + \overbrace{\sum_{j=1}^p f_j(x_j)}^\text{1st order effects} + \overbrace{\sum_{j\neq k}^p f_{jk}(x_j, x_k)}^\text{2nd order effects} + \ldots + \overbrace{f_{1,\ldots,p}(x_1, \ldots, x_p)}^\text{p-th order effect}
+$$
+
+## Functional Decomposition
+
+$$f(x)  = \overbrace{f_0}^\text{Intercept} + \overbrace{\sum_{j=1}^p f_j(x_j)}^\text{1st order effects} + \overbrace{\sum_{S \subseteq \{1,\ldots,p\},|S| \geq 2} f_{S}(x_S)}^{\text{Higher order effects}}
+$$
+
+## Functional Decomposition
+
+$$f(x)  = \underbrace{f_0 + \sum_{j=1}^p \overbrace{f_j(x_j)}^\text{How complex?} + \overbrace{IA(x)}^{\text{How much interaction?}}}_{\text{How many feature used?}}
+$$
 
 ## Accumulated Local Effects (ALE)
+
+TODO: Use all plots
 
@@
 ```
+## Measure: Main effect complexity
 
-## Functional Decomposition with ALE
+TODO: Plot of single ALE plot with complexity (from paper)
 
-TODO: copy paste new decomposition
+Plus formula how to average those for all plots
 
@@
 
-TODO: PRL formula with R2 and SSQ and so on
+Measure main effect model with proportion of error explained (pre):
 
-## Measure: Complexity
-
-TODO: AleNumApprox plot
-
-## Measure: Complexity cont.
-
-TODO: Multiple AleNumApprox plots with averaging formula
+$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_{ALE}))}{\mathbb{E}(L(\hat{f}, c))}$$
 
@@
 
-TODO: Show result graphics
+TODO: Show Interpretability vs. Performance Tradeoff
 
@@
 
-TODO: Show Interpretability vs. Performance Tradeoff
-
-```{r child = "../chunks/book.Rmd"}
-```
+TODO: Show result graphics
 

### 2019-03-18 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
almost finish first draft

@@
 output: 
-    beamer_presentation
+  beamer_presentation:
+    includes:
+      in_header: "header_pagenrs.tex"
 ---
@@
 ## Functional Decomposition
-
 $$f(x)  = \overbrace{f_0}^\text{Intercept} + \overbrace{\sum_{j=1}^p f_j(x_j)}^\text{1st order effects} + \overbrace{\sum_{S \subseteq \{1,\ldots,p\},|S| \geq 2} f_{S}(x_S)}^{\text{Higher order effects}}
@@
 
-$$f(x)  = \underbrace{f_0 + \sum_{j=1}^p \overbrace{f_j(x_j)}^\text{How complex?} + \overbrace{IA(x)}^{\text{How much interaction?}}}_{\text{How many feature used?}}
-$$
+$$f(x)  = \underbrace{f_0 + \sum_{j=1}^p \overbrace{f_j(x_j)}^\text{How complex?} + \overbrace{IA(x)}^{\text{How much interaction?}}}_{\text{How many feature used?}}$$
 
@@
 
-TODO: Use all plots
+```{r}
+knitr::include_graphics("../images/ale-intuition-1.jpg")
+```
+## Accumulated Local Effects (ALE)
+
+```{r}
+knitr::include_graphics("../images/ale-intuition-2.jpg")
+```
+
+## Accumulated Local Effects (ALE)
 
@@
 ```
-## Measure: Main effect complexity
 
-TODO: Plot of single ALE plot with complexity (from paper)
+## Accumulated Local Effects (ALE)
 
-Plus formula how to average those for all plots
+```{r}
+knitr::include_graphics("../images/ale-intuition-4.jpg")
+```
+## Accumulated Local Effects (ALE)
 
-## Measure: Interaction Strength
+```{r}
+knitr::include_graphics("../images/ale-intuition-5.jpg")
+```
 
-Measure main effect model with proportion of error explained (pre):
+## Main effect complexity
 
-$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_{ALE}))}{\mathbb{E}(L(\hat{f}, c))}$$
+- Approximate ALE plot with linear segments
+- Count number of non-zero coefficients
+- Average over all features, weight with variance
+
+```{r}
+knitr::include_graphics("../images/c-demo-1.jpeg")
+```
+## Interaction Strength
+
+Measure main effect model with proportion of error explained:
+
+$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_0 + \sum_{j=1}^p f_j)}{\mathbb{E}(L(\hat{f}, c))}$$
 
@@
 
-TODO: Wine plot
-TODO: List of models
+- Predict wine quality from phsicochemical properties
+- Minimize MAE, number of features, interaction strength, main effect complexity
+- Search across different model classes and hyperparameter settings
+
+```{r, out.width = "70%"}
+knitr::include_graphics("../images/wine-features.jpg")
+```
 
@@
 
-TODO: Show Interpretability vs. Performance Tradeoff
+```{r}
+knitr::include_graphics("../images/perf-interpret-tradeoff-1.jpeg")
+```
 
@@
 TODO: Show result graphics
+
+# The End
 

### 2019-03-19 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
continues talk

@@
 author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
-date: "2019-03-21"
+date: "DagStat 2019-03-21"
+institute: "LMU Munich"
 output: 
@@
 
-## Black Box problem
+## Black Box Problem
 
@@
 ## Problem: Tradeoff interpretability and performance
-Use only interpretable models (e.g. linear model)! $\Rightarrow$ Pay with loss in performance.
+Use only interpretable models (e.g. linear model)!
+
+$\Rightarrow$ Pay with loss in performance.
 
@@
 ## Problem: Tradeoff interpretability and performance
-Just use post-hoc interpretation methods (e.g. feature importance)! $\Rightarrow$ Works better for less complex models.
+Just use post-hoc interpretation methods (e.g. feature importance)!
+
+$\Rightarrow$ Works better for less complex models.
 
@@
 
-TODO of fuzzy animal which dodges touch of hand
-Or use image of cat that doges hand
+```{r}
+knitr::include_graphics("../images/spirit-board.jpg")
+```
 
@@
 
-# The End
+## Interested in interpretable machine learning?
+More on interpretable machine learning in my book [http://christophm.github.io/interpretable-ml-book/](http://christophm.github.io/interpretable-ml-book/).
 
-# Backup
-
-## Units in Wine dataset
-
-- fixed acidity g(tartaric acid)/dm^3^
-- volatile acidity: g(acetric acid/dm^3^)
-- citric acid: g/dm^3^
-- residual sugar: g/dm^3^
-- chlorides: g(sodium chloride)/dm^3^
-- free sulfur dioxide: mg/dm^3^
-- total sulfur dioxide: mg/dm^3^
-- density> g/cm^3^
-- pH
-- sulphates: g(postassium sulphate) / dm^3^
-- alcohol vol.%
+```{r, out.width="5cm"}
+knitr::include_graphics("../images/cover-bordered.jpg")
+```

### 2019-03-19 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
finishes slides

@@
 
-## We propose
-
-TODO: Replace text with image: Black box in the middle, three properties attached to it
+## We propose measures of model complexity
 
@@
 
-$\Rightarrow$ Makes claims of interpretability more explicit ("Model A uses less features than B and has less interactions)
+$\Rightarrow$ Makes claims of interpretability more explicit ("Model A uses less features than B and has less interactions")
 
@@
 
-TODO: Show result graphics
+```{r, out.width='70%'}
+knitr::include_graphics("../images/multi-obj-selection.png")
+```
 

### 2019-03-20 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
fix some errors

@@
 ```
-## Problem: Tradeoff interpretability and performance
-Use only interpretable models (e.g. linear model)!
+## Solution: Interpretable Models?
 
-$\Rightarrow$ Pay with loss in performance.
+$\Rightarrow$ Pay with loss in performance; Unclear how to compare models
 
@@
 ```
-## Problem: Tradeoff interpretability and performance
-Just use post-hoc interpretation methods (e.g. feature importance)!
+## Solution: Post-hoc Interpretation?
 
-$\Rightarrow$ Works better for less complex models.
+$\Rightarrow$ Post-hoc interpretation (e.g. feature importance, partial dependence plots) works better for less complex models.
 
@@
 
-## What's interpretability anyway?
+## Is Interpretability Unscientific?
 
@@
 
-## We propose measures of model complexity
+## We Propose Measures of Model Complexity
 
@@
 
-$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_0 + \sum_{j=1}^p f_j)}{\mathbb{E}(L(\hat{f}, c))}$$
+$$\text{Interaction Strength} =  \frac{\mathbb{E}(L(\hat{f}, f_0 + \sum_{j=1}^p f_j))}{\mathbb{E}(L(\hat{f}, c))}$$
 
@@
 
-- Predict wine quality from phsicochemical properties
+- Predict wine quality from physicochemical properties
 - Minimize MAE, number of features, interaction strength, main effect complexity

### 2019-05-22 2019-05-sommerklausur/2019-sommerklausur-molnar.Rmd
adds slides for sommerklausur

--- 2019-03-20 2019-03-21-dagstat-imlm/dagstat-molnar.Rmd
@@
 author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
-date: "DagStat 2019-03-21"
+date: "Sommerklausur 2019-05-24"
 institute: "LMU Munich"
@@
 
-## Interested in interpretable machine learning?
-More on interpretable machine learning in my book [http://christophm.github.io/interpretable-ml-book/](http://christophm.github.io/interpretable-ml-book/).
+# The End
 
-```{r, out.width="5cm"}
-knitr::include_graphics("../images/cover-bordered.jpg")
+## Why not GAM?
+
+```{r echo = TRUE}
+library("mgcv")
+library("iml")
+set.seed(42)
+X = data.frame(x1 = rnorm(100), x2 = rnorm(100))
+f = function(model = NULL, newdata) {
+	1 *  (newdata[,1] > 0)
+}
+y = f(newdata = X)
+dat = cbind(y, X)
+mod.gam = gam(y ~ s(x1) + s(x2), data = dat)
+
 ```
+
+## Why not GAM?
+
+```{r fig.height = 5}
+plot(mod.gam)
+```
+
+## Why not GAM?
+
+```{r}
+pred = Predictor$new(predict.fun = f, data = X)
+fes = FeatureEffects$new(pred)
+plot(fes, ncol = 2)
+```

### 2019-05-24 2019-05-sommerklausur/2019-sommerklausur-molnar.Rmd
final version of slides

@@
 ```
+
+## Outline
+
+- Why measure interpretability?
+- Functional Decomposition with Accumulated Local Effects
+- Measures
+- Application
 
@@
 
-$\Rightarrow$ Allows model comparision
+$\Rightarrow$ Allows model comparison
 
@@
 
+## Summary
+
+- Model-agnostic interpretability measures: number of features, interaction strength, main effect complexity
+- Use: model comparison and optimization metric
+
+- Implementation: \url{https://github.com/compstat-lmu/paper_2019_iml_measures}
+- Paper:[https://arxiv.org/abs/1904.03867]()
+
 # The End

### 2019-06-26 2019-reisensburg/2019-reisensburg-molnar.Rmd
start reisensburg slides

--- 2019-05-22 2019-05-sommerklausur/2019-sommerklausur-molnar.Rmd
@@
 author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
-date: "Sommerklausur 2019-05-24"
+date: "Statistical Computing 2019-07-03"
 institute: "LMU Munich"
@@
 
+## Outline
+
+- Minimizing model complexity maximizes interpretability
+- Functional Decomposition
+- Measures
+- Application
+
 ## Black Box Problem
@@
 ```
-## Solution: Interpretable Models?
 
-$\Rightarrow$ Pay with loss in performance; Unclear how to compare models
-
-```{r, echo = FALSE}
-knitr::include_graphics("../images/white-box.png")
-```
 ## Solution: Post-hoc Interpretation?
 
-$\Rightarrow$ Post-hoc interpretation (e.g. feature importance, partial dependence plots) works better for less complex models.
+<!-- $\Rightarrow$ Post-hoc interpretation (e.g. feature importance, partial dependence plots) works better for less complex models. -->
 
@@
 
-## Is Interpretability Unscientific?
+## Sparsity Improves Interpretation
 
-```{r}
-knitr::include_graphics("../images/spirit-board.jpg")
-```
+TODO: Image of 3 vs 100 feature in importance
+
+## Additivity Improves Interpretation
+
+TODO: Image from paper with PDP and increasing interactions
+
+## Linearity Improves Interpretation
+
+TODO: Image with increasing MEC
 
@@
 
-$\Rightarrow$ Makes claims of interpretability more explicit ("Model A uses less features than B and has less interactions")
+$\Rightarrow$ Allows choosing models with improved post-hoc interpretation
 
@@
 
-```{r}
-knitr::include_graphics("../images/perf-interpret-tradeoff-1.jpeg")
-```
-
-## Application: Multi-Objective Optimization
-
 ```{r, out.width='70%'}
@@
 ```
+
+## Summary
+
+- Model-agnostic complexity measures: number of features, interaction strength, main effect complexity
+- Use: model comparison and optimization metric
+
+- Implementation: \url{https://github.com/compstat-lmu/paper_2019_iml_measures}
+- Paper:[https://arxiv.org/abs/1904.03867]()
 

### 2019-06-27 2019-reisensburg/2019-reisensburg-molnar.Rmd
finishes version of reisensburg talk

@@
 
-TODO: Image of 3 vs 100 feature in importance
+```{r, echo=FALSE, out.width = '\\textwidth'} 
+knitr::include_graphics("../images/pdp-multiple.png")
+```
 
@@
 
-TODO: Image from paper with PDP and increasing interactions
+```{r, echo=FALSE}
+knitr::include_graphics("../images/pdp-interactions.jpeg")
+```
 
@@
 
-TODO: Image with increasing MEC
+```{r}
+set.seed(2)
+x = seq(from = -1, to = 1, length.out = 50)
+y1 = x
+y2 = x * (x < 0.3) + 0.4 * (x > 0.3)
+y3 = x + sin(8 * x) 
+y3 = y2 + cumsum(rt(50, df = 1)) / 50 - 1
+
+dat = data.frame(x = rep(x, times = 3), y = c(y1, y2, y3), situation = rep(c(1,2,3), each = 50))
+library(ggplot2)
+ggplot(dat) + geom_line(aes(x = x, y = y)) + facet_grid(. ~ situation)
+
+```
 

### 2019-07-02 2019-reisensburg/2019-reisensburg-molnar.Rmd
finalize
fixes title

@@
 ---
-title: "Measuring and optimizing machine learning interpretability"
+title: "Quantifying Interpretability of Arbitrary Machine Learning Models Through Functional Decomposition"
 author: "Christoph Molnar, Giuseppe Casalicchio, Bernd Bischl"
@@
 
-## Additivity Improves Interpretation
-
-```{r, echo=FALSE}
-knitr::include_graphics("../images/pdp-interactions.jpeg")
-```
-
 ## Linearity Improves Interpretation
@@
 
+```
+ ## Additivity Improves Interpretation
+
+```{r, echo=FALSE}
+knitr::include_graphics("../images/pdp-interactions.jpeg")
 ```

### 2019-07-03 chunks/story-wines.Rmd
finalize talk

@@
 
+## Disclaimer: I am NOT a wine expert!
+
+```{r, out.width = "50%"}
+knitr::include_graphics("../images/wine-expert.png")
+```
+
 ## Quality Distribution
@@
 
-## How can we solve this problem? {.emphasizedabit}
-
-- Hand-written rules
-- Statistical Model
-- Machine Learning
+# How can we solve this problem?
 
@@
 
-## {data-background=../images/dog-car.gif data-background-size=contain}
+## {data-background=../images/make-it-rain.gif data-background-size=contain}
 

### 2019-07-04 chunks/story-wines.Rmd
finalizes talk\

@@
-##  {.emphasizedabit}
-Predict wine quality from its physicochemical properties.
+# Let's Predict Wine Quality
 
-## {.center data-background=../images/minho.png data-background-size=contain}
+## 
+```{r, out.width = "100%"}
+knitr::include_graphics("../images/wine-features.jpg")
+```
+
+## Disclaimer: I am NOT a wine expert!
+
+```{r, out.width = "50%"}
+knitr::include_graphics("../images/wine-expert.png")
+```
+
+# How can we develop a wine quality prediction program?
+
+```{r, child="../chunks/what-is-ml.Rmd"}
+```
+
+# Step 1: Find data
 
@@
 
-## {.center data-background=../images/wine-features.jpg data-background-size=contain}
-
-## Disclaimer: I am NOT a wine expert!
-
-```{r, out.width = "50%"}
-knitr::include_graphics("../images/wine-expert.png")
-```
-
-## Quality Distribution
+## Wine Quality Distribution
 
@@
   scale_x_continuous("Wine Quality", labels = 1:10, breaks = 1:10)
-```
-
-# How can we solve this problem?
-
-```{r, child="../chunks/what-is-ml.Rmd"}
-```
-
-# Step 1: Find data
-
-## 
-```{r}
-knitr::include_graphics("../images/checkmark.png")
 ```

### 2020-08-28 2020-ecml-xkdd/paper/paper.tex
copy to lncs template

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Interpretable Machine Learning -- State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No. 01IS18036A.
The authors of this work take full responsibilities for its content.
}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Christoph Molnar\inst{1}\orcidID{0000-0003-2331-868X}}
%\and
%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{Molnar}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{LMU Munich, Ludwigstr. 33, 80539 Munich, Germany
\email{christoph.molnar@gmail.com}\\
\url{https://www.slds.stat.uni-muenchen.de/people/molnar/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
I present a very short history of the field of interpretable machine learning, give an overview of popular methods and discuss challenges when interpreting machine learning models.

When machine learning (ML) is used, be it in a product, in a process or in research, interpretability of the model is often a deciding factor.
Interpretable machine learning (IML) is not an entirely new field, as rule-based ML started already in the 1950s and linear regression models go even back to the 1800s.
But I argue that it gained widespread attention in the formative last years starting around 2015.
In this paper which accompanies a talk, I give a brief overview of the developments in IML, popular IML methods and discuss challenges the field faces.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Background}
Today, ML is being used for knowledge discovery.
Scientists that would have used some (interpretable) statistical model with carefully crafted features and assumptions about the data distribution switched to the machine learning approach which select the model that fits the data best, even if that means that the model is not interpretable.
However, while this approach can yield much more accurate models, it became less obvious how to derive knowledge from the model.
Today, ML is used in many products, such as drones, games, object detection and so on.
To get a good product, the ML model is evaluated, usually using performance evaluation on test data.
But this might not be enough, since this can hide more severe modeling errors such as overlooking some structure in the training data.
Examples are Amazons hiring tool which made mysognyst decisions or an hospital system that wrongly assigns lower lung problem probability to asthma patients.
These examples have in common that the performance on test data was not affected and would go unnoticed without further looking into the model.
Today, ML is used in decision making in processes.
For example for pre-sorting fraudulent insurance claims.
Once a claim is flagged, it goes to a case worker.
A claim that is not only flagged, but that comes in with some explanations why it was flagged is much more useful to the case worker.

All these cases have in common that they have two goals: High predictive performance AND interpretability.
This was also recognized by the researcher community who is now very busy producing research and tools for intepreting ML models.
Learning interpretable prediction models from data -- that is not new and has been focus of, e.g. statistical learning and rule learning for a long time.
I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
I will shortly review how we got where we are, which methods are available and what the challenges are still.

\section{A Very Short History}
When we go back to the roots, I think this would be linear models and decision rules.
Things like cybernetics and expert systems are not of interest here, as we focus on a subset of AI technologies that learn from data, so only machine learning.

Linear models are very old, starting with Gauss and legegegegeg who used it for solar system equations.
qtetetetet then later popularized it for learning from data in the social sciences.

Until 1950s mostly statistical methods available, so I guess we can't speak yet of interpretable machine learning.
A side note: I don't want to distinguish too sharply between statistics and machine learning.
But for this context, I would say in statistical modeling the focus is on finding a suitable linear model using in-sample metrics such as R squared and AIC.
In machine learning we use an external test set to minimize the generalization loss and the model itself does not matter as long as the loss is low.

From 1950s on there is some first research on ML, with an AI winter following in the 1970s.
Again in 1980s / 1990s research picked up for SVMs, neural networks and so on.
Very interesting milestone is random forest in 1995, which comes with some built-in interpretability tools, 1995, and extended and popularized in 2001 by Breiman.
Still used to this day in many sciences such as medicine and psychology.

In 2010s then came the deep learning hype.

The big leap came in the mid 2010's, visible in Google Search trends [link] and Web of Science statistics when searching for 'Explainable AI' and 'Interpretable Machine Learning'.
These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion.
Research was done, big time in statistics for regression models in its various extensions, rule based stuff in computer science, random forest and its many extensions and interpretation.
What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.
Especially model-agnostic methods that work for any type of model have grown a lot in this new field.

\section{Where Do We Stand Today in 2020?}

We have software, big companies that offer solutions and many startups.
Research is heavily underway and people are already using the solutions.

\section{What Methods do We Have?}

There are many overview papers with categorizations of all these methods.
Here are the properties I find the most useful to categorize the various methods and allow a user to understand when to use which.

An interesting distinction is how the methods work:
They either explain individual 'parts' of the model, examples are the weights in a linear model, the structure of a tree.
This can be very straightforward such as the interpretation of a linear regression weight, which directly tells us how a feature influences the response.
Or this can be more complex, when we want to understand what the role of individual neurons in a convolutional neural network is.
Here we can do activation maximization which tells us to which image the neuron most strongly reacts to and interpret this neuron as, e.g., a detector of a certain pattern.
Methods that require to look at certain parts of the model are by definition model-specific, i.e. they cannot be applied to any model.
Methods that mostly observe the input - output behavior are often model-agnostic, i.e. they ignore whether the model is a neural network, linear model or random forest.

A second distinction is about whether a method explains an individual data point or the global model behavior.
Methods such as Shapley Values and LIME compute explanations for points, while there are many feature importance methods that quantify the importance of a feature globally for the model, meaning across all data points.

\subsection{Linear Models and Rule-based ML}

Linear models and rule can be seen as building blocks for many types of explanations and explanation methods.
Interpretations are often presented as a weighted sum of the inputs, such as LIME, Shapley Values, and many of the attribution methods for convolutional neural networks.
Even the entire model can -- in theory -- be deconstructed into and additive sum of components reaching from first-order effects to all interactions (cite functional ANOVA).
But also rule based explanations are very common types of explanations.
There are many new tree algorithms, such as XXX and XXX.

\paragraph{Linear Models}
Linear models have been around since 1800s.
The beauty of them is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
That makes estimation and interpretation simple.
But this is often not adequate since in the real world the effects are usually not linear.
Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
Other extensions allow for non-linear effect of a feature, by using splines.
Also interactions can be integrated.
All these extensions make the model more difficult to interpret, but also allow to more flexible model the target.

\paragraph{Rule-based ML}
There are different approaches.
They all have in common that they find decision rules in the form of conditions combined with AND conjunctions leading to some prediction if all conditions are matched.
There are different ways to find this rules and if they are allowed to overlap and so on.
Differences in how they are optimized and so on.

\subsection{Explain Individual Predictions}

\paragraph{Counterfactual Explanations}
Counterfactual explanations are rather simple explanations for individual predictions.
A counterfactual explanation tells you how the features should be changed to get to a different prediction or classification.
But when you look closer at how you find them, it gets more difficult.
There are many tradeoffs to make.
First, you want to get as close as possible to the desired outcome.
At the same time you don't want to change too many features, and the features that are changed should not be changed too much.
Also these changes should not create unrealistic combinations of feature values, think of a 2m person for which the weight was changed to 30kg.
The reason that there are many papers on the topic is that the search for counterfatuals can be expensive.
Some methods are only for specific types of models, and they all trade-off the various objectives in different ways.
We argued that it is necessary to treat this as a multi-objective problem and the result should be not one but many counterfactual explanations.

\paragraph{LIME}

\paragraph{Shapley Values}

\subsection{Global Model Interpretation}

\paragraph{Feature Effects}

\paragraph{Feature Importance}

\subsection{Other Methods}

\paragraph{Heatmaps}

\section{Challenges}

\subsection{Dependent Features}
Feature dependence comes in many forms, correlation being the simplest and most known.
When features are dependent -- and they usually are -- disentangling the effect of one from the other are difficult if not impossible.
For example, if you measure various markers of inflammation in a blood sample, they will be highly correlated.
Using these markers in a predictive model for classifying diseases for which these markers are highly predictive, the model can rely on any of them with arbitrary weighting.
Attributing which marker was the most important now becomes difficult.

Almost all interpretation methods are affected by this issue.
Most of them ignore this issue and pretend the features are independent of each other.
For example the permutation feature importance measure quantifies the importance as the drop in performance when the feature is permuted.
But permutation when features are dependent not only breaks the association with the target value, but also with the other features.
Now if we permute one blood marker, it might create new combination that might be very unlikely.
These combinations are now fed to the model to get the predictions, but these are outside of the training data distribution and the model never had to deal with similar combinations.
It is totally unclear how the model will behave and it might even be irrelevant how the model will behave for these data points as they will never occur in real life.
There are attempts to 'fix' this issue, namely to not permute the feature, but to draw from the conditional distribution.
This means we still sample new values, but only the ones that are conforming with the other feature values.
Unfortunately, this is not a simple 'fix' of the issue.
While it solves the extrapolation issue, it becomes a new measure altogether, since conditional feature importance measure the drop in performance "given that we know the other feature values".
This can mean that the two most important features can drop to almost zero importance when yhey are highly correlated.

Similar things happen to 'fixes' for e.g. Shapley Values.
Once we start sampling from the conditional distribution and not from the marginal, we don't get Shapley Values for our original question, but something else.

I fear that there is no easy solution.

\subsection{Causality}
In general, we are not allowed to make a causal interpretation of machine learning models.
There is an inherent conflict between prediction performance optimization and causal modeling.
A feature might give some additional predictive performance, but its effect may not be interpreted causally, because, e.g., we did not include the confounder of both this feature and the target.
The fact that the ground has some predictive value for tomorrow's weather, but it's not causal for it.
The common confounder is today's weather -- todays rain makes the ground wet.
And if we had no access to todays weather but only to the "wet ground"-feature, we have to decide between predictive performance and causal correct interpretation.

\subsection{Inference}
Example for confidence intervals for LRP https://arxiv.org/pdf/2008.01468.pdf.
Is model-specific for Deep Neural Networks using Dropout.

\subsection{Many More Topics}
This was not an exhaustive list, and especially focused on the methodological side, neglecting views of the user interface, societal impacts, psychological insights and so on.
These challenges are more from the methodological, mathematical view, i.e., what can be done given we have some data and model.
This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.

\section{Discussion}

While Interpretable Machine Learning can seem like a pretty young field, it has some old roots in Statistics and Computer Science.
Yet, many new methods have come in the recent years and we have witnessed a Cambrian explosion of methods, software and institutional interest.
We are seing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
These methods also are implemented now in many software packages.
Big leaps in the field came from old ideas in other fields.
The Shapley Value has been for a long time part of game theory, and its introduction to machine learning interpretability has spurred a lot of development.
We can learn a lot from Statistics and how it handles uncertainty.

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
Oct 2017
\end{thebibliography}
\end{document}

### 2020-09-02 2020-ecml-xkdd/paper/paper.tex
continue draft

@@
 \begin{abstract}
-I present a very short history of the field of interpretable machine learning, give an overview of popular methods and discuss challenges when interpreting machine learning models.
+  I present a very short history of the field of interpretable machine learning (IML), give an overview of popular interpretation methods and discuss challenges when interpreting machine learning models.
 
-When machine learning (ML) is used, be it in a product, in a process or in research, interpretability of the model is often a deciding factor.
-Interpretable machine learning (IML) is not an entirely new field, as rule-based ML started already in the 1950s and linear regression models go even back to the 1800s.
-But I argue that it gained widespread attention in the formative last years starting around 2015.
-In this paper which accompanies a talk, I give a brief overview of the developments in IML, popular IML methods and discuss challenges the field faces.
-
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{Interpretable Machine Learning \and Explainable AI}
 \end{abstract}
@@
 %
-\section{Background}
-Today, ML is being used for knowledge discovery.
-Scientists that would have used some (interpretable) statistical model with carefully crafted features and assumptions about the data distribution switched to the machine learning approach which select the model that fits the data best, even if that means that the model is not interpretable.
-However, while this approach can yield much more accurate models, it became less obvious how to derive knowledge from the model.
-Today, ML is used in many products, such as drones, games, object detection and so on.
-To get a good product, the ML model is evaluated, usually using performance evaluation on test data.
-But this might not be enough, since this can hide more severe modeling errors such as overlooking some structure in the training data.
-Examples are Amazons hiring tool which made mysognyst decisions or an hospital system that wrongly assigns lower lung problem probability to asthma patients.
-These examples have in common that the performance on test data was not affected and would go unnoticed without further looking into the model.
-Today, ML is used in decision making in processes.
-For example for pre-sorting fraudulent insurance claims.
-Once a claim is flagged, it goes to a case worker.
-A claim that is not only flagged, but that comes in with some explanations why it was flagged is much more useful to the case worker.
+Interpretability is often a deciding factor when machine learning (ML) is used in a product, a decision process or in research.
+Interpretable Machine learning gives the means to debug a model, extract insights and justify decisions.
+I want to give three very common use cases:
 
-All these cases have in common that they have two goals: High predictive performance AND interpretability.
-This was also recognized by the researcher community who is now very busy producing research and tools for intepreting ML models.
-Learning interpretable prediction models from data -- that is not new and has been focus of, e.g. statistical learning and rule learning for a long time.
-I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
-I will shortly review how we got where we are, which methods are available and what the challenges are still.
+%\paragraph{For science:} In applied sciences, data is often modeled with ML nowadays, where before statistical modeling approach would have been used.
+%The goal of science is not only to predict the world well, but also to generate further insights.
+%Examples are species distribution models in ecology which aim to explain why certain species can be found on certain areas based on e.g. the local climate.
+%Here, interpretability of the prediction model is needed to understand why a species, e.g., has a high probability to occur in a specific reason.
 
-\section{A Very Short History}
-When we go back to the roots, I think this would be linear models and decision rules.
-Things like cybernetics and expert systems are not of interest here, as we focus on a subset of AI technologies that learn from data, so only machine learning.
+\paragraph{A Very Short History of IML.}
+Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
+Statistics has a long tradition of learning interpretable prediction  models from data, starting in 1800 \cite{stigler1986history} with works on least squares by Legendre 1805, \cite{legendre1805nouvelles}, Gauss in 1809 \cite{gauss1809theoria} and Quetelet in 1827 \cite{quetelet1827recherches}.
+Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
+  This took off in the 80s with decision tree approaches such as CART CITE, ID3 CITE and rule learners like AQ CITE but reaches back to first research in the 1960s \cite{hajek1966guha}.
+Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g. model-based trees CITE).
+As a next stepping stone to the field of IML are random forests, an early approach to explain more complex machine learning models.
+I argue that one of the success factors of the random forest as a prediction model was (and is) the built-in feature importance measure.
+The high citation counts of the original paper (CITE), but also suggestions for how to improve the feature importance (CITE, CITE, CITE) are evidence for this.
+The IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
+Could it be that only the terms have changed, but the field has always been there and the same?
+I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time, see chapter (XXX).
 
-Linear models are very old, starting with Gauss and legegegegeg who used it for solar system equations.
-qtetetetet then later popularized it for learning from data in the social sciences.
+\paragraph{Today} we have reached a first state of maturity where some methods were established, implemented in software and used by practitioners and applied researchers.
+There is a lot of open source software that implement various techniques (CITE iml, CITE DALEX, CITE sklearn, ...).
+Big tech companies offer software (CITE ai360, CITE what-if tool, ...).
+Regulation such as GDPR has spurred a discussion around further needs of interpretability.
+Many startup are now focused on explainable / interpretable ML (TODO: Find good source).
 
+\section{Methods:}
+
+In this section I want to very broadly talk about methods that we have to explain models.
+A first useful distinction is between inherently interpretable models and usually model-agnostic explanation methods that describe the behavior of the model.
+This is a distinction between trying to make sense of a model by analyzing its parts vs. making sense of it by analyzing its prediction for certain inputs (sensitivity analysis).
+However we interpret a model, the interpretation is always projected back to the features themselves or some abstract transformation of them.
+
+\paragraph{Inherently interpretable models} are models where we can interpret its learned structure and parameters.
+In this category, I see all linear regression models of the form $Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$, where the weights $\beta$ are learned parameters of the model.
+Through the models structure (weighted sum of features) we can interpret the weights as the effect that a feature has on the prediction.
+Similarly, decision trees and other rule-based ML models have a learned structure which we can interpret as how the model makes a prediction.
+For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
+However, the more complex these structures get (e.g., a linear model with hundres of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
+Surprisingly, we can also try assing meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
+This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
+
+\paragraph{Sensitivity-analysis-like (model-agnostic) interpretation methods} often treats the model as a black-box (but not always).
+Here I also want to distinguis between local and global explanations.
+
+\paragraph{Local explanations} explain how individual predictions are derived.
+Here we have a great variety of methods, with very different methodological motivations.
+Popular methods are LIME, Shapley Values and Counterfactual Explanations.
+I want to highlight especially their very different approaches to the problem of local explanations.
+Counterfactual explanations are what-if scenarios.
+How does an instance have to change to get another (desired) prediction or classification.
+They build on a rich tradition in philosophy and also thrive in what the social sciences discovered:
+Good explanations are contrastive and focus on a few reasons.
+A very different approach originates from collaborative game theory: The Shapley Values.
+How do you fairly share some payout from a game among the players?
+The Shapley Values provide an answer for this question, by trying out all constellations of players and averaging the additional payout a player would add to the total payout when the player is added to a team.
+The same idea can be applied to ML: The predicted value is the payout, the player is a certain feature value.Many variations, extensions and more efficient implementations for specific ML models exist (CITE shap, tree shap, Other stuff).
+The third mention here is LIME.
+LIME is a surrogate model, i.e., the procedure approximates the original ML model with a simpler interpretable model.
+LIME does so locally, meaning only in some neighborhood around the data point that we want to explain.
+
+\paragraph{Surrogate Models} are an interesting and huge class of explanation method.
+This approach is very common in sensitivity analysis where we have emulation? models.
+Surrogate models combine the two strategies:
+First, approximate the predictions of the ML model with an interpretable model, such as a decision tree.
+This is a kind of sensitivity-like approach, where we see the original model as a black box and artificially probe the ML model and derive information from the behavior.
+Then we interpret the interpretable model, which now acts as a surrogate to the more complex ML model.
+
+\paragraph{Global, model-agnostic explanation methods} are also very popular.
+Here we have have feature importance, which rank features based on how relevant they were for the prediction and the feature effect, which expresses how a change in that feature changes the predicted outcome on average.
+Permutation Feature Importance, which is based on permuting features is popular importance measure, originally suggested for random forests, now available as model-agnostic version.
+Alternatives are variance based measures, see XXX for an overview of all the ways to measure importance.
+For feature effects we have Partial Dependence Plots (CITE), Individual Conditional Expectation Plots (ICE) and Accumulated Local Effect (ALE) Plots (CITE), with of course many variations (CITE, CITE).
+
+\section{old}
+
+\paragraph{A Very Short History of Machine Learning Interpretability}
+Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
+In statistics, there is a long tradition of learning interpretable models from data, starting in 1800 \cite{stigler1986history}.
+The least squares  method to fit linear models reaches back to 1805 by Legendre. \cite{legendre1805nouvelles}
+Later the linear regression model became a tool for astronomy and geodesy, e.g. by Gauss in 1809 \cite{gauss1809theoria}.
+Starting 1827 Quetelet used linear regression models (with least squares method) for (at that time) large amounts of data from social sciences, such as crime data \cite{quetelet1827recherches}.
+We can hardly call the least squares method interpretable machine learning, but it remains a highly relevant building block.
+Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
 Until 1950s mostly statistical methods available, so I guess we can't speak yet of interpretable machine learning.
@@
 
-From 1950s on there is some first research on ML, with an AI winter following in the 1970s.
-Again in 1980s / 1990s research picked up for SVMs, neural networks and so on.
-Very interesting milestone is random forest in 1995, which comes with some built-in interpretability tools, 1995, and extended and popularized in 2001 by Breiman.
-Still used to this day in many sciences such as medicine and psychology.
+I want to highlight the random forest CITE as a representative for early interpretable ML approach, which also lay out some more groundwork and a new paradigm for feature importance.
 
@@
 
+I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
+
 The big leap came in the mid 2010's, visible in Google Search trends [link] and Web of Science statistics when searching for 'Explainable AI' and 'Interpretable Machine Learning'.
 These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion.
+
 Research was done, big time in statistics for regression models in its various extensions, rule based stuff in computer science, random forest and its many extensions and interpretation.
@@
 
-\section{Where Do We Stand Today in 2020?}
+\paragraph{Where Do We Stand Today in 2020?}
 
+We have software that implements interpretability (CITE lots of stuff here).
+Many of the big IT companies
 We have software, big companies that offer solutions and many startups.
@@
 
-\section{What Methods do We Have?}
+\paragraph{What Methods do We Have?}
 
@@
 \paragraph{Linear Models}
-Linear models have been around since 1800s.
-The beauty of them is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
+The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
 That makes estimation and interpretation simple.
@@
 \subsection{Explain Individual Predictions}
+
+There was a great explosion in different methods to explain individual predictions of ML models in a model-agnostic way.
+Among the most popular are Local Interpretable Model-agnostic Explanations (LIME) CITE, Shapley Values from Game Theory (CITE x2) and Counterfactual Explanations.
+Their outputs of LIME and Shapley Value are weighted sums and are estimated using method from linear regression, again showing the deep roots with statistics.
+Counterfactual explanations are presented as decision rules, for LIME this is also possible
+
+\paragraph{Surrogate Models}
+Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.
 
@@
 %
-% \bibliographystyle{splncs04}
 % \bibliography{mybibliography}
 %
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
 
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
+\vskip 0.2in
+\bibliographystyle{splncs04}
+\bibliography{Bib}
 
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
 \end{document}

### 2020-09-03 2020-ecml-xkdd/paper/paper.tex
first bad draft

@@
 
+TODO: Mention sensitivity analysis
+
 \paragraph{A Very Short History of IML.}
@@
 Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
+Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
+The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
   This took off in the 80s with decision tree approaches such as CART CITE, ID3 CITE and rule learners like AQ CITE but reaches back to first research in the 1960s \cite{hajek1966guha}.
@@
 The high citation counts of the original paper (CITE), but also suggestions for how to improve the feature importance (CITE, CITE, CITE) are evidence for this.
-The IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
+
+In 2010s then came the deep learning hype.
+And a few years after that, the IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
 Could it be that only the terms have changed, but the field has always been there and the same?
+I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
 I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time, see chapter (XXX).
+These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion, since developments happened under different names and in different fields.
+What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.
 
@@
 In this section I want to very broadly talk about methods that we have to explain models.
-A first useful distinction is between inherently interpretable models and usually model-agnostic explanation methods that describe the behavior of the model.
-This is a distinction between trying to make sense of a model by analyzing its parts vs. making sense of it by analyzing its prediction for certain inputs (sensitivity analysis).
 However we interpret a model, the interpretation is always projected back to the features themselves or some abstract transformation of them.
+Often the distinction is made between model-specific (works only for one type of model) and model-agnostic (works for all models) interpretation methods.
+I want to distinguish between introspection and sensitivity analysis.
+An interpretation method that works by introspection assigns semantic meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model and describes its behavior}.
+Introspection is necessarily model-specific.
+Sensitivity analysis is mostly model-agnostic, but can also be done by leveraging model-specific knowledge such as gradient-based methods.
+FOOTNOTE: An example are the various heatmap-like explanation for CNNs for image classification (CITE many) which study the sensitivity of the prediction to the input pixels.
+Technically this is done by backpropagation and therefore model-specific.
+Model-agnostic versions (CITE) exist.
+The other useful category is local vs. global explanations.
+Local means that the method explains individual predictions, and global means it explains the average model behavior.
+For inherently interpretable models, global and local explanations are often the same: In linear models, the coefficients both how the model behaves globally, but also is used for individuals instances.
 
 \paragraph{Inherently interpretable models} are models where we can interpret its learned structure and parameters.
+As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
 In this category, I see all linear regression models of the form $Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$, where the weights $\beta$ are learned parameters of the model.
@@
 However, the more complex these structures get (e.g., a linear model with hundres of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
-Surprisingly, we can also try assing meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
+Surprisingly, we can also try assign meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
 This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
+Another example is the random forest minimal depth distribution TODO: CITE
+The distribution of minimal depth shows for a feature for all the trees in the forest at which depth it occurs, with the notation that it the feature is more important if it appears early in the tree, it is more important.
+This is again an example how we analyze the structure of a model to get an understanding of how it makes decisions, but the model is more complex (hundreds of decision trees) so there is more work to interpreting its learned structure.
+The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
+By using our distinction of introspection and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
+Gini importance collects for each feature and tree how the split decreased the Gini impurity.
+It does not rely on making predictions, but only on internal statistics.
+For mean decrease accuracy, we probe the random forest with perturbed data, i.e., we shuffle a feature and see how the prediction changes.
+It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
+Model-agnostic versions of this permutation feature importance exists ( TODO: CITE).
 
 \paragraph{Sensitivity-analysis-like (model-agnostic) interpretation methods} often treats the model as a black-box (but not always).
-Here I also want to distinguis between local and global explanations.
+Here I also want to distinguish between local and global explanations.
 
@@
 This approach is very common in sensitivity analysis where we have emulation? models.
+Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.
 Surrogate models combine the two strategies:
@@
 
-\section{old}
-
-\paragraph{A Very Short History of Machine Learning Interpretability}
-Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
-In statistics, there is a long tradition of learning interpretable models from data, starting in 1800 \cite{stigler1986history}.
-The least squares  method to fit linear models reaches back to 1805 by Legendre. \cite{legendre1805nouvelles}
-Later the linear regression model became a tool for astronomy and geodesy, e.g. by Gauss in 1809 \cite{gauss1809theoria}.
-Starting 1827 Quetelet used linear regression models (with least squares method) for (at that time) large amounts of data from social sciences, such as crime data \cite{quetelet1827recherches}.
-We can hardly call the least squares method interpretable machine learning, but it remains a highly relevant building block.
-Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
-Until 1950s mostly statistical methods available, so I guess we can't speak yet of interpretable machine learning.
-A side note: I don't want to distinguish too sharply between statistics and machine learning.
-But for this context, I would say in statistical modeling the focus is on finding a suitable linear model using in-sample metrics such as R squared and AIC.
-In machine learning we use an external test set to minimize the generalization loss and the model itself does not matter as long as the loss is low.
-
-I want to highlight the random forest CITE as a representative for early interpretable ML approach, which also lay out some more groundwork and a new paradigm for feature importance.
-
-In 2010s then came the deep learning hype.
-
-I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
-
-The big leap came in the mid 2010's, visible in Google Search trends [link] and Web of Science statistics when searching for 'Explainable AI' and 'Interpretable Machine Learning'.
-These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion.
-
-Research was done, big time in statistics for regression models in its various extensions, rule based stuff in computer science, random forest and its many extensions and interpretation.
-What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.
-Especially model-agnostic methods that work for any type of model have grown a lot in this new field.
-
-\paragraph{Where Do We Stand Today in 2020?}
-
-We have software that implements interpretability (CITE lots of stuff here).
-Many of the big IT companies
-We have software, big companies that offer solutions and many startups.
-Research is heavily underway and people are already using the solutions.
-
-\paragraph{What Methods do We Have?}
-
-There are many overview papers with categorizations of all these methods.
-Here are the properties I find the most useful to categorize the various methods and allow a user to understand when to use which.
-
-An interesting distinction is how the methods work:
-They either explain individual 'parts' of the model, examples are the weights in a linear model, the structure of a tree.
-This can be very straightforward such as the interpretation of a linear regression weight, which directly tells us how a feature influences the response.
-Or this can be more complex, when we want to understand what the role of individual neurons in a convolutional neural network is.
-Here we can do activation maximization which tells us to which image the neuron most strongly reacts to and interpret this neuron as, e.g., a detector of a certain pattern.
-Methods that require to look at certain parts of the model are by definition model-specific, i.e. they cannot be applied to any model.
-Methods that mostly observe the input - output behavior are often model-agnostic, i.e. they ignore whether the model is a neural network, linear model or random forest.
-
-A second distinction is about whether a method explains an individual data point or the global model behavior.
-Methods such as Shapley Values and LIME compute explanations for points, while there are many feature importance methods that quantify the importance of a feature globally for the model, meaning across all data points.
-
-\subsection{Linear Models and Rule-based ML}
-
-Linear models and rule can be seen as building blocks for many types of explanations and explanation methods.
-Interpretations are often presented as a weighted sum of the inputs, such as LIME, Shapley Values, and many of the attribution methods for convolutional neural networks.
-Even the entire model can -- in theory -- be deconstructed into and additive sum of components reaching from first-order effects to all interactions (cite functional ANOVA).
-But also rule based explanations are very common types of explanations.
-There are many new tree algorithms, such as XXX and XXX.
-
-\paragraph{Linear Models}
-The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
-That makes estimation and interpretation simple.
-But this is often not adequate since in the real world the effects are usually not linear.
-Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
-Other extensions allow for non-linear effect of a feature, by using splines.
-Also interactions can be integrated.
-All these extensions make the model more difficult to interpret, but also allow to more flexible model the target.
-
-\paragraph{Rule-based ML}
-There are different approaches.
-They all have in common that they find decision rules in the form of conditions combined with AND conjunctions leading to some prediction if all conditions are matched.
-There are different ways to find this rules and if they are allowed to overlap and so on.
-Differences in how they are optimized and so on.
-
-\subsection{Explain Individual Predictions}
-
-There was a great explosion in different methods to explain individual predictions of ML models in a model-agnostic way.
-Among the most popular are Local Interpretable Model-agnostic Explanations (LIME) CITE, Shapley Values from Game Theory (CITE x2) and Counterfactual Explanations.
-Their outputs of LIME and Shapley Value are weighted sums and are estimated using method from linear regression, again showing the deep roots with statistics.
-Counterfactual explanations are presented as decision rules, for LIME this is also possible
-
-\paragraph{Surrogate Models}
-Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.
-
-\paragraph{Counterfactual Explanations}
-Counterfactual explanations are rather simple explanations for individual predictions.
-A counterfactual explanation tells you how the features should be changed to get to a different prediction or classification.
-But when you look closer at how you find them, it gets more difficult.
-There are many tradeoffs to make.
-First, you want to get as close as possible to the desired outcome.
-At the same time you don't want to change too many features, and the features that are changed should not be changed too much.
-Also these changes should not create unrealistic combinations of feature values, think of a 2m person for which the weight was changed to 30kg.
-The reason that there are many papers on the topic is that the search for counterfatuals can be expensive.
-Some methods are only for specific types of models, and they all trade-off the various objectives in different ways.
-We argued that it is necessary to treat this as a multi-objective problem and the result should be not one but many counterfactual explanations.
-
-\paragraph{LIME}
-
-\paragraph{Shapley Values}
-
-\subsection{Global Model Interpretation}
-
-\paragraph{Feature Effects}
-
-\paragraph{Feature Importance}
-
-\subsection{Other Methods}
-
-\paragraph{Heatmaps}
-
 \section{Challenges}
 
-\subsection{Dependent Features}
-Feature dependence comes in many forms, correlation being the simplest and most known.
-When features are dependent -- and they usually are -- disentangling the effect of one from the other are difficult if not impossible.
-For example, if you measure various markers of inflammation in a blood sample, they will be highly correlated.
-Using these markers in a predictive model for classifying diseases for which these markers are highly predictive, the model can rely on any of them with arbitrary weighting.
-Attributing which marker was the most important now becomes difficult.
+There are also challenges for the field, that are already being tackled but still remain issues.
+This list is also based on this paper (CITE Pitfalls).
 
-Almost all interpretation methods are affected by this issue.
-Most of them ignore this issue and pretend the features are independent of each other.
-For example the permutation feature importance measure quantifies the importance as the drop in performance when the feature is permuted.
-But permutation when features are dependent not only breaks the association with the target value, but also with the other features.
-Now if we permute one blood marker, it might create new combination that might be very unlikely.
-These combinations are now fed to the model to get the predictions, but these are outside of the training data distribution and the model never had to deal with similar combinations.
-It is totally unclear how the model will behave and it might even be irrelevant how the model will behave for these data points as they will never occur in real life.
-There are attempts to 'fix' this issue, namely to not permute the feature, but to draw from the conditional distribution.
-This means we still sample new values, but only the ones that are conforming with the other feature values.
-Unfortunately, this is not a simple 'fix' of the issue.
-While it solves the extrapolation issue, it becomes a new measure altogether, since conditional feature importance measure the drop in performance "given that we know the other feature values".
-This can mean that the two most important features can drop to almost zero importance when yhey are highly correlated.
-
-Similar things happen to 'fixes' for e.g. Shapley Values.
-Once we start sampling from the conditional distribution and not from the marginal, we don't get Shapley Values for our original question, but something else.
-
+\paragraph{Feature dependence} makes interpretation more difficult.
+Inherently, when two features are dependent, e.g., correlated, attributing effects and importance to them is difficult.
+Most methods also suffer from extrapolation when features are dependent.
+When features are, e.g., permuted individually, this breaks the association with the target to predict (that is desirable) but it also breaks the association with the other features (which can cause problems).
+The problem here is that new data points are being generated that are unlikely.
+These points are then used for computing importance values, explanations for individual predictions and so on.
+Many model-agnostic methods have this issue: PFI, PDP, Shapley values, LIME, ...
+But also interpretation for inherently interpretable models can suffer:
+When features are strongly correlated, the variance of estimated weights in a linear model can become large.
+Also the interpretation of a weight is conditional that all other features remain fixed, which becomes the more unrealistic the more dependence there is between features.
+There are some "fixes" for it, which enforce that any manipulation of the data respects the joint distribution.
+All solutions have the problem that they now mix the effects and importance of features that switch the focus of interpretation or even break the original interpretation method in the case of Shapley values (CITE, CITE).
+Solutions exist for feature importance, called conditional feature importance CITE APPROACHES.
+Currently there is no satisfying solution that fixes extrapolation, but keeps the original marginal interpretation.
 I fear that there is no easy solution.
 
-\subsection{Causality}
-In general, we are not allowed to make a causal interpretation of machine learning models.
-There is an inherent conflict between prediction performance optimization and causal modeling.
-A feature might give some additional predictive performance, but its effect may not be interpreted causally, because, e.g., we did not include the confounder of both this feature and the target.
-The fact that the ground has some predictive value for tomorrow's weather, but it's not causal for it.
-The common confounder is today's weather -- todays rain makes the ground wet.
-And if we had no access to todays weather but only to the "wet ground"-feature, we have to decide between predictive performance and causal correct interpretation.
-
-\subsection{Inference}
+\paragraph{Inference}
+In statistics it is common procedure to quantify the uncertainties that come with estimates of, e.g., coefficients in a linear model.
+Quantification of uncertainty, be it in the form of error bars, confidence intervals, test statistics or p-values, helps to separate signal from noise.
+Many techniques in IML currently only give the point estimates, but don't quantify the uncertainty.
+For things like feature importance there are variations, but we need to have these things built into common software tools.
+Also it is unclear which uncertainty should be quantified and for what purpose.
+Should we treat the model as fixed?
+Or should we quantify the uncertainty of some feature effect in the same manner as we do for the generalization error, that is with for example cross-validation.
+Here we have to dig deeper into the statistics literature which can be our guide.
 Example for confidence intervals for LRP https://arxiv.org/pdf/2008.01468.pdf.
@@
 
-\subsection{Many More Topics}
+\paragraph{Causal interpretation}
+Usually, we desire a causal interpretation of an effect or importance.
+As a researcher you want to draw conclusion about the true underlying process that generated your observations and learn about the world.
+When you want to use ML for decision making, you need causal reasons that can change the target in the real world.
+For example, when modeling the likelihood that customers will quit their subscription, only a good prediction will not be enough, but you have the desire to prevent a customer from leaving.
+Only causal interpretation of the model will help you find reasonable real-world interventions.
+However, this is generally not possible and causality can even be a conflicting goal to predictive accuracy.
+For example when we don't have access to an important confounder of a (non-causal) feature and the target, it improve predictive performance if we include the non-causal feature.
+When we want to predict whether it will rain tomorrow, a useful feature might be whether the ground is wet today, assuming we don't have access to today's weather.
+Todays weather is the cause for both the wet ground and tomorrows weather.
+Including it in the model will improve performance, but we are not allowed to interpret the effect and importance for the wet ground causally for the weather tomorrow.
+Here, knowledge from the causality community is important.
+It's not enough to somehow fix causality within the interpretation method, but we must go back to the modeling process itself to consider causality already there.
+
+\paragraph{Lack of definition of interpretability and explainability} is a common critique of the field (TODO: CITE some papers, e.g. Lipton).
+First, I don't think the lack of definition is a problem for the "naming" of the field.
+I think of "Intepretable ML" and "Explainable AI" as useful terms to collect all topics that are related to analyzing ML models beyond performance metrics.
+There is also very valid critique in lack of definition.
+Because there is no single, clean definition of interpretability, quantitative research results can be hard to come by, e.g., how can you show that some interpretation method delivers better explanations than another?
+This is connected to a lack of groundtruth, we just don't know what the correct explanation is.
+To counter this argument, interpretability depends on audience and context (CITE PLEASE StOP?)
+It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity (CITE), interaction strength (CITE) and much more (CITE ...).
+It will take some time to establish some best practice for evaluation interpretation methods and the explanations they produce.
+
+\paragraph{And much more} research has to be done on which I did not touch.
 This was not an exhaustive list, and especially focused on the methodological side, neglecting views of the user interface, societal impacts, psychological insights and so on.
@@
 This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.
+How do we present multiple maybe conflicting explanations, e.g., when you generate multiple counterfactual explanations?
 
@@
 Yet, many new methods have come in the recent years and we have witnessed a Cambrian explosion of methods, software and institutional interest.
-We are seing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
+We are seeing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
 These methods also are implemented now in many software packages.

### 2020-09-04 2020-ecml-xkdd/paper/paper.tex
adds first citations

@@
 %
-\title{Interpretable Machine Learning -- State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
+\title{Interpretable Machine Learning -- A Short History, State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
 01IS18036A.
@@
 The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
-  This took off in the 80s with decision tree approaches such as CART CITE, ID3 CITE and rule learners like AQ CITE but reaches back to first research in the 1960s \cite{hajek1966guha}.
-Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g. model-based trees CITE).
+  This took off in the 80s with decision tree approaches such as CART and ID3 and rule learners like AQ \cite{furnkranz2012foundations} but reaches back to first research in the 1960s \cite{hajek1966guha}.
+Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g.
+model-based trees \cite{zeileis2008model}).
 As a next stepping stone to the field of IML are random forests, an early approach to explain more complex machine learning models.
 I argue that one of the success factors of the random forest as a prediction model was (and is) the built-in feature importance measure.
-The high citation counts of the original paper (CITE), but also suggestions for how to improve the feature importance (CITE, CITE, CITE) are evidence for this.
+The high citation counts (>60000 citation on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also suggestions for how to improve the random forest feature importance (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}) are evidence for this.
 
 In 2010s then came the deep learning hype.
-And a few years after that, the IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (see Figure XXX) and papers published with these terms (see Figure XXX).
+And a few years after that, the IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (Figure \ref{fig:count} right)and papers published with these terms (Figure \ref{fig:count} left).
+\begin{figure}
+  \includegraphics[width=\textwidth]{citation-search.png}
+  \caption{Left, Right}
+  \label{fig:count}
+\end{figure}
 Could it be that only the terms have changed, but the field has always been there and the same?
 I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
-I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time, see chapter (XXX).
+I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time.
 These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion, since developments happened under different names and in different fields.
@@
 \paragraph{Today} we have reached a first state of maturity where some methods were established, implemented in software and used by practitioners and applied researchers.
-There is a lot of open source software that implement various techniques (CITE iml, CITE DALEX, CITE sklearn, ...).
-Big tech companies offer software (CITE ai360, CITE what-if tool, ...).
+There is a lot of open source software that implement various techniques (\cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml}, ...).
+Big tech companies offer software \cite{exler2019if,arya2020ai,hall2017machine}...
 Regulation such as GDPR has spurred a discussion around further needs of interpretability.
@@
 Sensitivity analysis is mostly model-agnostic, but can also be done by leveraging model-specific knowledge such as gradient-based methods.
-FOOTNOTE: An example are the various heatmap-like explanation for CNNs for image classification (CITE many) which study the sensitivity of the prediction to the input pixels.
+FOOTNOTE: An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
 Technically this is done by backpropagation and therefore model-specific.
-Model-agnostic versions (CITE) exist.
+Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.
 The other useful category is local vs. global explanations.
@@
 For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
-However, the more complex these structures get (e.g., a linear model with hundres of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
+However, the more complex these structures get (e.g., a linear model with hundreds of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
 Surprisingly, we can also try assign meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
 This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
-Another example is the random forest minimal depth distribution TODO: CITE
+Another example is the random forest minimal depth distribution:\cite{randomForestExplainer}
 The distribution of minimal depth shows for a feature for all the trees in the forest at which depth it occurs, with the notation that it the feature is more important if it appears early in the tree, it is more important.
@@
 It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
-Model-agnostic versions of this permutation feature importance exists ( TODO: CITE).
+Model-agnostic versions of this permutation feature importance exists \cite{fisher2019all}.
 
@@
 
-\paragraph{Lack of definition of interpretability and explainability} is a common critique of the field (TODO: CITE some papers, e.g. Lipton).
+\paragraph{Lack of definition of interpretability and explainability} is a common critique of the field \cite{lipton2018mythos}.
 First, I don't think the lack of definition is a problem for the "naming" of the field.
@@
 This is connected to a lack of groundtruth, we just don't know what the correct explanation is.
-To counter this argument, interpretability depends on audience and context (CITE PLEASE StOP?)
-It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity (CITE), interaction strength (CITE) and much more (CITE ...).
+To counter this argument, interpretability depends on audience and context TODO:CITE.
+It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, simulability (CHECK) and much more \cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing},
 It will take some time to establish some best practice for evaluation interpretation methods and the explanations they produce.

### 2020-09-07 2020-ecml-xkdd/paper/paper.tex
adds citations

@@
 How do you fairly share some payout from a game among the players?
-The Shapley Values provide an answer for this question, by trying out all constellations of players and averaging the additional payout a player would add to the total payout when the player is added to a team.
-The same idea can be applied to ML: The predicted value is the payout, the player is a certain feature value.Many variations, extensions and more efficient implementations for specific ML models exist (CITE shap, tree shap, Other stuff).
-The third mention here is LIME.
+The Shapley values \cite{shapley1953value} provide an answer for this question, by trying out all constellations of players and averaging the additional payout a player would add to the total payout when the player is added to a team.
+The same idea can be applied to ML: The predicted value is the payout, the player is a certain feature value.Many variations, extensions and more efficient implementations for specific ML models exist \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}.
+The third mention here are local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}
 LIME is a surrogate model, i.e., the procedure approximates the original ML model with a simpler interpretable model.
@@
 
-\paragraph{Surrogate Models} are an interesting and huge class of explanation method.
+\paragraph{Surrogate Models} are an interesting and huge class of explanation method \cite{puri2017magix,molnar2019,ming2018rulematrix}.
 This approach is very common in sensitivity analysis where we have emulation? models.
@@
 Then we interpret the interpretable model, which now acts as a surrogate to the more complex ML model.
+There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which would be more the introspection approach.
 
@@
 Here we have have feature importance, which rank features based on how relevant they were for the prediction and the feature effect, which expresses how a change in that feature changes the predicted outcome on average.
-Permutation Feature Importance, which is based on permuting features is popular importance measure, originally suggested for random forests, now available as model-agnostic version.
-Alternatives are variance based measures, see XXX for an overview of all the ways to measure importance.
-For feature effects we have Partial Dependence Plots (CITE), Individual Conditional Expectation Plots (ICE) and Accumulated Local Effect (ALE) Plots (CITE), with of course many variations (CITE, CITE).
+Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
+Alternatives are variance based measures, see \cite{wei2015variable} for an overview of all the ways to measure importance.
+For feature effects we have Partial Dependence Plots \cite{friedman2001greedy}, Individual Conditional Expectation Plots \cite{goldstein2015peeking} and Accumulated Local Effect  (ALE)  Plots \cite{apley2016visualizing}.
 
@@
 There are also challenges for the field, that are already being tackled but still remain issues.
-This list is also based on this paper (CITE Pitfalls).
+This list is also based on this paper \cite{molnar2020pitfalls}.
 
@@
 There are some "fixes" for it, which enforce that any manipulation of the data respects the joint distribution.
-All solutions have the problem that they now mix the effects and importance of features that switch the focus of interpretation or even break the original interpretation method in the case of Shapley values (CITE, CITE).
-Solutions exist for feature importance, called conditional feature importance CITE APPROACHES.
+All solutions have the problem that they now mix the effects and importance of features that switch the focus of interpretation or even break the original interpretation method in the case of Shapley values \cite{sundararajan2019many,janzing2019feature}.
+Solutions exist for feature importance, called conditional feature importance \cite{watson2019testing,candes2018panning,strobl2008conditional,molnar2020model}
 Currently there is no satisfying solution that fixes extrapolation, but keeps the original marginal interpretation.
@@
 This is connected to a lack of groundtruth, we just don't know what the correct explanation is.
-To counter this argument, interpretability depends on audience and context TODO:CITE.
+To counter this argument, interpretability depends on audience and context.
 It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, simulability (CHECK) and much more \cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing},

### 2020-09-07 2020-ecml-xkdd/paper/paper.tex
improve challenges and discussion

@@
 % \renewcommand\UrlFont{\color{blue}\rmfamily}
+% For correct quotation marks (command \enquote)
+\usepackage{csquotes}
 
@@
 %
-%
 Interpretability is often a deciding factor when machine learning (ML) is used in a product, a decision process or in research.
-Interpretable Machine learning gives the means to debug a model, extract insights and justify decisions.
-I want to give three very common use cases:
-
+With interpretable machine learning (IML) \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangably} the user can find out problems, extract insights and justify decisions.
+%I want to give three very common use cases:
 %\paragraph{For science:} In applied sciences, data is often modeled with ML nowadays, where before statistical modeling approach would have been used.
@@
 Sensitivity analysis is mostly model-agnostic, but can also be done by leveraging model-specific knowledge such as gradient-based methods.
-FOOTNOTE: An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
+\footnote{An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
 Technically this is done by backpropagation and therefore model-specific.
-Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.
+Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.}
 The other useful category is local vs. global explanations.
@@
 
-There are also challenges for the field, that are already being tackled but still remain issues.
-This list is also based on this paper \cite{molnar2020pitfalls}.
+This is an incomplete list of some of the challenges of IML, mostly based on \cite{molnar2020pitfalls}.
 
-\paragraph{Feature dependence} makes interpretation more difficult.
-Inherently, when two features are dependent, e.g., correlated, attributing effects and importance to them is difficult.
-Most methods also suffer from extrapolation when features are dependent.
-When features are, e.g., permuted individually, this breaks the association with the target to predict (that is desirable) but it also breaks the association with the other features (which can cause problems).
-The problem here is that new data points are being generated that are unlikely.
-These points are then used for computing importance values, explanations for individual predictions and so on.
-Many model-agnostic methods have this issue: PFI, PDP, Shapley values, LIME, ...
-But also interpretation for inherently interpretable models can suffer:
-When features are strongly correlated, the variance of estimated weights in a linear model can become large.
-Also the interpretation of a weight is conditional that all other features remain fixed, which becomes the more unrealistic the more dependence there is between features.
-There are some "fixes" for it, which enforce that any manipulation of the data respects the joint distribution.
-All solutions have the problem that they now mix the effects and importance of features that switch the focus of interpretation or even break the original interpretation method in the case of Shapley values \cite{sundararajan2019many,janzing2019feature}.
-Solutions exist for feature importance, called conditional feature importance \cite{watson2019testing,candes2018panning,strobl2008conditional,molnar2020model}
-Currently there is no satisfying solution that fixes extrapolation, but keeps the original marginal interpretation.
-I fear that there is no easy solution.
+\paragraph{Feature dependence:} When two features share information (e.g., BMI and weight) separating their effects and importances is difficult.
+Some IML methods (e.g.LIME, permutation feature importance or Shapley values) simply \enquote{ignore} such dependence, which results in extrapolation, i.e., for the interpretation new, unrealistic data points are created (e.g., low weight but high BMI).
+Interpretations based on extrapolation into areas where the model never saw data can be very misleading \cite{hooker2019please}.
+There have been attempts to \enquote{fix} IML methods by creating only realistic data points (using the conditional instead of the marginal distribution).
+These conditional IML methods fix the extrapolation problem, but they also entangle the interpretations of a feature's effect and importance with all other features it depends on.
+In the case of Shapley values the conditional variant even breaks the interpretation \cite{sundararajan2019many,janzing2019feature}.
 
-\paragraph{Inference}
-In statistics it is common procedure to quantify the uncertainties that come with estimates of, e.g., coefficients in a linear model.
-Quantification of uncertainty, be it in the form of error bars, confidence intervals, test statistics or p-values, helps to separate signal from noise.
-Many techniques in IML currently only give the point estimates, but don't quantify the uncertainty.
-For things like feature importance there are variations, but we need to have these things built into common software tools.
-Also it is unclear which uncertainty should be quantified and for what purpose.
-Should we treat the model as fixed?
-Or should we quantify the uncertainty of some feature effect in the same manner as we do for the generalization error, that is with for example cross-validation.
+\paragraph{Uncertainty quantification} is a typical part of statistical modeling and becomes visible as error bars, confidence intervals, test statistics and p-values.
+Uncertainty quantification helps to separate signal from noise.
+Many IML methods only return the point estimates, but don't quantify the uncertainty.
+This reaches across most methods: rule-based ML , feature effect plots, feature importance, ...
+There is research about uncertainty quantification of model interpretations, e.g., feature importance \cite{watson2019testing,fisher2019all}, layer-wise relevance propagation \cite{fabi2020feature} and Shapley values \cite{williamson2020efficient}.
+However, it is not implemented as default in most IML software.
+Also we need a more united view on the topic, and specify which uncertainty is quantified (model bias, model variance, IML estimation error, ...) and to what end.
+The statistics community is very deep into statistical inference and IML can learn a lot.
 Here we have to dig deeper into the statistics literature which can be our guide.
-Example for confidence intervals for LRP https://arxiv.org/pdf/2008.01468.pdf.
-Is model-specific for Deep Neural Networks using Dropout.
+
+\paragraph{Statistical inference:}
+When uncertainty quantification is combined with assumptions about the data generating process, we can draw conclusions about the real world.
+But for that we have to better understand the properties and assumptions of the various IML methods.
+For example, methods such as partial dependence plots require that features are independent.
+To enable statistical inference, there are at least two challenges to overcome: 1) define the assumptions needed that allow a model (and an IML method) to be interpreted as the \enquote{true} model and 2) tools and procedures to check these assumptions.
 
 \paragraph{Causal interpretation}
-Usually, we desire a causal interpretation of an effect or importance.
-As a researcher you want to draw conclusion about the true underlying process that generated your observations and learn about the world.
-When you want to use ML for decision making, you need causal reasons that can change the target in the real world.
-For example, when modeling the likelihood that customers will quit their subscription, only a good prediction will not be enough, but you have the desire to prevent a customer from leaving.
-Only causal interpretation of the model will help you find reasonable real-world interventions.
-However, this is generally not possible and causality can even be a conflicting goal to predictive accuracy.
-For example when we don't have access to an important confounder of a (non-causal) feature and the target, it improve predictive performance if we include the non-causal feature.
-When we want to predict whether it will rain tomorrow, a useful feature might be whether the ground is wet today, assuming we don't have access to today's weather.
-Todays weather is the cause for both the wet ground and tomorrows weather.
-Including it in the model will improve performance, but we are not allowed to interpret the effect and importance for the wet ground causally for the weather tomorrow.
-Here, knowledge from the causality community is important.
-It's not enough to somehow fix causality within the interpretation method, but we must go back to the modeling process itself to consider causality already there.
+Ideally, a model relies on the true causes to make a prediction.
+Then we could also interpret it causally, making statements about the underlying phenomenon, which is the focus of ML in research, but also makes models robust against adversarial attacks, but also makes the model more usefule when used as a basis for decision making.
+Unfortunately, predictive performance and causality can be conflicting goals.
+For example, today's weather directly causes tomorrow's weather, but we might only have access to feature \enquote{wet ground}.
+Using \enquote{wet ground} in the prediction model for \enquote{tomorrow's weather} is useful as it has information about \enquote{today's weather}, but we are not allowed to interpret it causally, because the confounder \enquote{today's weather} is missing.
+Further research is needed to understand when we are allowed to make causal interpretations of an ML model.
 
-\paragraph{Lack of definition of interpretability and explainability} is a common critique of the field \cite{lipton2018mythos}.
-First, I don't think the lack of definition is a problem for the "naming" of the field.
-I think of "Intepretable ML" and "Explainable AI" as useful terms to collect all topics that are related to analyzing ML models beyond performance metrics.
-There is also very valid critique in lack of definition.
-Because there is no single, clean definition of interpretability, quantitative research results can be hard to come by, e.g., how can you show that some interpretation method delivers better explanations than another?
-This is connected to a lack of groundtruth, we just don't know what the correct explanation is.
-To counter this argument, interpretability depends on audience and context.
-It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, simulability (CHECK) and much more \cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing},
-It will take some time to establish some best practice for evaluation interpretation methods and the explanations they produce.
+\paragraph{Lack of definition} is a common critique of the field \cite{lipton2018mythos,doshi2017towards}.
+Without a definition of "interpretability", how can we decide if a new method is better at explanation an ML method?
+To evaluate the \textbf{predictive performance} of an ML model, we simply compute the prediction error on test data given the groundtruth label.
+To evaluate the \textbf{interpretability} of that same ML model is not as easy.
+We do not know what the groundtruth explanation looks like and have no straightforward mean to quantify how interpretable a model is or how correct an explanation is.
+Instead of having one groundtruth explanation, various \textbf{quantifiable aspects of interpretability} are emerging.\cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing}
+Sparsity of the explanation (or the model itself), interaction strength of features within the ML model and
+It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength and a user's ability to run a model on a given input (simulatability) are examples of such aspects.
+The challenge ahead remains to establish a best practice on how to evaluate interpretation methods and the explanations they produce.
 
-\paragraph{And much more} research has to be done on which I did not touch.
-This was not an exhaustive list, and especially focused on the methodological side, neglecting views of the user interface, societal impacts, psychological insights and so on.
-These challenges are more from the methodological, mathematical view, i.e., what can be done given we have some data and model.
+We have focused mostly on the methodological, mathematical challenges in a rather static setting, where you are given a machine learning model and data.
+\paragraph{Many more} challenges lie ahead to improve interpretability of machine learning.
+What is the best way to explain a machine learning model for people with a specific background and how do we present the explanations?
+How are people affected by explanations, how are explanations maybe misused (CITE fairwashing by HIma)?
 This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.
@@
 
-While Interpretable Machine Learning can seem like a pretty young field, it has some old roots in Statistics and Computer Science.
-Yet, many new methods have come in the recent years and we have witnessed a Cambrian explosion of methods, software and institutional interest.
-We are seeing both ways of making models themselves more interpretable, but also model-agnostic methods that work for any type of model.
-These methods also are implemented now in many software packages.
-Big leaps in the field came from old ideas in other fields.
-The Shapley Value has been for a long time part of game theory, and its introduction to machine learning interpretability has spurred a lot of development.
-We can learn a lot from Statistics and how it handles uncertainty.
+Interpretable Machine Learning is a young field, but has deeper roots in Statistics and Computer Science and also draws from other fields such as the social sciences.
+There has been Cambrian explosion of methods since 2015, including model-agnostic interpretation methods and many specialised methods for, e.g., deep neural networks.
+Many of these methods have found their way into open source software and products of startups and tech companies.
+Interpretable Machine Learning is used in science, company products and processes.
+While we have reached a first maturity, there a still some challenges ahead, such as feature dependence, causality and inference.
+Many IML methods are imported from other fields, so thinking outside the box (pun intended) might help in the challenges ahead.
+We believe that the field has to both reach out horizontally -- to other domains -- and vertically -- draw from the rich research in statistics and computer science.
 

### 2020-09-08 2020-ecml-xkdd/paper/paper.tex
finish first draft

@@
 %
-\title{Interpretable Machine Learning -- A Short History, State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
+\title{Interpretable Machine Learning Methods -- A Brief History, State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
 01IS18036A.
@@
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{IML - History, State of the Art, Challenges}
 % If the paper title is too long for the running head, you can set
@@
   I present a very short history of the field of interpretable machine learning (IML), give an overview of popular interpretation methods and discuss challenges when interpreting machine learning models.
+  This view might be a bit biased through the eyes of a statistician.
 
@@
 Interpretability is often a deciding factor when machine learning (ML) is used in a product, a decision process or in research.
-With interpretable machine learning (IML) \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangably} the user can find out problems, extract insights and justify decisions.
+With interpretable machine learning (IML) \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangably} the user can find out problems of the ML model, extract insights and justify decisions.
 %I want to give three very common use cases:
@@
 
-TODO: Mention sensitivity analysis
-
 \paragraph{A Very Short History of IML.}
-Learning interpretable prediction models from data is not new, but has been the focus of statistical learning and rule-based ML for a long time.
-Statistics has a long tradition of learning interpretable prediction  models from data, starting in 1800 \cite{stigler1986history} with works on least squares by Legendre 1805, \cite{legendre1805nouvelles}, Gauss in 1809 \cite{gauss1809theoria} and Quetelet in 1827 \cite{quetelet1827recherches}.
-Another methodological pillow are rule-based machine learning approaches, which includes models such as decision rules and decision trees.
-Many extensions exists, such as non-linear transformations for the case that the target follows a different distribtuion (GLMs, logistic regression).
-The beauty of linear models is that they are additive, meaning that the influences of each feature on the target can be added together to give the full effect.
-  This took off in the 80s with decision tree approaches such as CART and ID3 and rule learners like AQ \cite{furnkranz2012foundations} but reaches back to first research in the 1960s \cite{hajek1966guha}.
+A lot of IML research just happened in the last five years.
+But learning interpretable models from data has a much longer tradition.
+Linear regression models have been used back in 1800 by Gauss, Legendre and Quetelet \cite{stigler1986history,legendre1805nouvelles,gauss1809theoria,quetelet1827recherches}.
+Rule-based machine learning (e.g., decision rules and trees) has been an active research areas since the 1980s \cite{furnkranz2012foundations} with some roots reaching back into the 1960s \cite{hajek1966guha}.
 Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g.
-model-based trees \cite{zeileis2008model}).
-As a next stepping stone to the field of IML are random forests, an early approach to explain more complex machine learning models.
-I argue that one of the success factors of the random forest as a prediction model was (and is) the built-in feature importance measure.
-The high citation counts (>60000 citation on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also suggestions for how to improve the random forest feature importance (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}) are evidence for this.
-
-In 2010s then came the deep learning hype.
-And a few years after that, the IML field really took off in around 2015, judging on usage of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (Figure \ref{fig:count} right)and papers published with these terms (Figure \ref{fig:count} left).
+model-based trees \cite{zeileis2008model} or RuleFit \cite{friedman2008predictive}).
+Machine learning research in general also started around that time, e.g., support vector machines in 1974 \cite{vapnik1974theory}, early important work on neural networks in the 1960s and boosting in 1990 \cite{schapire1990strength}.
+Interpretability has always been a companion to ML.
+A good example for this is the random forest, which I also see as an IML milestone for making ML practical.
+Random forests, a black box machine learning model that often works well without tuning, came with a built-in feature importance measure, which contributed to the success.
+The evidence that interpretability was success factor are the many citations (>60000 citation on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also of papers improving the importance measure (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}).
+In the 2010s came the deep learning hype, after a deep neural network won the ImageNet challenge.
+A few years after that, the IML field really took off (around 2015), judging by frequency of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (Figure \ref{fig:count} right)and papers published with these terms (Figure \ref{fig:count} left).
 \begin{figure}
@@
 \end{figure}
-Could it be that only the terms have changed, but the field has always been there and the same?
-I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
-I would say no, since a lot of commonly used methods, especially model-agnostic ones and the ones for deep neural network also came about this time.
-These figures look like there has not been much research on explaining prediction models before, which would be a wrong conclusion, since developments happened under different names and in different fields.
-What has changed is that Interpretable Machine Learning or Explainable Artificial Intelligence has become a field that now ties all these topics together.
+Since then, many approaches have been introduced to the field, many of them model-agnostic explanation methods, i.e., working for different types of ML models, but also model-specific explanation methods to interpret, e.g., deep neural networks and tree ensembles.
+Rule-based ML and statistical regression models continue to be developed.
+%Could it be that only the terms have changed, but the field has always been there and the same?
+%I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
 
-\paragraph{Today} we have reached a first state of maturity where some methods were established, implemented in software and used by practitioners and applied researchers.
-There is a lot of open source software that implement various techniques (\cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml}, ...).
-Big tech companies offer software \cite{exler2019if,arya2020ai,hall2017machine}...
+\paragraph{Today} IML reached a first state of usability.
+Some methods have been established with a better understanding of weaknesses and strengths.
+There is a lot of open source software that implement various techniques (\cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml}, ...) and also big tech companies offer software \cite{exler2019if,arya2020ai,hall2017machine}.
+These software tools gives ML practitioners the means to interpret their models.
 Regulation such as GDPR has spurred a discussion around further needs of interpretability.
-Many startup are now focused on explainable / interpretable ML (TODO: Find good source).
+Many startup are now focused on interpretable ML.
 
-\section{Methods:}
+\section*{IML Methods}
 
-In this section I want to very broadly talk about methods that we have to explain models.
-However we interpret a model, the interpretation is always projected back to the features themselves or some abstract transformation of them.
-Often the distinction is made between model-specific (works only for one type of model) and model-agnostic (works for all models) interpretation methods.
-I want to distinguish between introspection and sensitivity analysis.
-An interpretation method that works by introspection assigns semantic meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model and describes its behavior}.
-Introspection is necessarily model-specific.
-Sensitivity analysis is mostly model-agnostic, but can also be done by leveraging model-specific knowledge such as gradient-based methods.
-\footnote{An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
+In this section we broadly cover some popular IML methods.
+%Often a distinction is made between model-specific and model-agnostic explanation methods.
+%An interpretation method is model-specific if it only works for a specific ML model, it is model-agnostic if it can be applied to any model.
+We distinguish IML methods between introspection and sensitivity analysis\footnote{Not to be confused with the research field of sensitivity analysis, which studies the uncertainty of outputs in mathematical models and systems.
+But there are methodological overlaps (e.g., Shapley values)}.
+An IML method that works by introspection assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.
+The introspection is always model-specific, e.g., a linear regression model can be interpreted by analyzing its coefficients, while a decision tree can be interpreted by visualizing the learned decision structure.
+Sensitivity analysis is mostly model-agnostic and works by manipulating input data and analyzing ML model  predictions.
+Some IML sensitivity analysis methods rely on model-specific knowledge such as gradients of neural networks.
+An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
 Technically this is done by backpropagation and therefore model-specific.
-Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.}
-The other useful category is local vs. global explanations.
-Local means that the method explains individual predictions, and global means it explains the average model behavior.
-For inherently interpretable models, global and local explanations are often the same: In linear models, the coefficients both how the model behaves globally, but also is used for individuals instances.
+Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.
 
-\paragraph{Inherently interpretable models} are models where we can interpret its learned structure and parameters.
-As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
-In this category, I see all linear regression models of the form $Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$, where the weights $\beta$ are learned parameters of the model.
-Through the models structure (weighted sum of features) we can interpret the weights as the effect that a feature has on the prediction.
-Similarly, decision trees and other rule-based ML models have a learned structure which we can interpret as how the model makes a prediction.
-For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
-However, the more complex these structures get (e.g., a linear model with hundreds of features and complex interaction terms or very deep decision trees), the more difficult it becomes to see this as interpretable.
-Surprisingly, we can also try assign meaning to structures of more complex models such as individual neurons or layers in a convolutional neural network.
-This leads to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
-Another example is the random forest minimal depth distribution:\cite{randomForestExplainer}
-The distribution of minimal depth shows for a feature for all the trees in the forest at which depth it occurs, with the notation that it the feature is more important if it appears early in the tree, it is more important.
-This is again an example how we analyze the structure of a model to get an understanding of how it makes decisions, but the model is more complex (hundreds of decision trees) so there is more work to interpreting its learned structure.
-The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
-By using our distinction of introspection and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
-Gini importance collects for each feature and tree how the split decreased the Gini impurity.
-It does not rely on making predictions, but only on internal statistics.
-For mean decrease accuracy, we probe the random forest with perturbed data, i.e., we shuffle a feature and see how the prediction changes.
-It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
-Model-agnostic versions of this permutation feature importance exists \cite{fisher2019all}.
+%The other useful category is local vs. global explanations.
+%Local means that the method explains individual predictions, and global means it explains the average model behavior.
+%For inherently interpretable models, global and local explanations are often the same: In linear models, the coefficients both how the model behaves globally, but also is used for individuals instances.
 
-\paragraph{Sensitivity-analysis-like (model-agnostic) interpretation methods} often treats the model as a black-box (but not always).
-Here I also want to distinguish between local and global explanations.
+\paragraph{Inherently interpretable models} are models with interpretable, learned structures and parameters (introspection).
+%As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
+Linear regression models ($Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$) can be interpreted by introspection:
+The model structure, a weighted sum of features, allows that we interpret the $\beta$-weights as the effect that a feature has on the prediction.
+Many extensions exists and linear regression models are remain an active field of research.
+Decision trees and other rule-based ML models have a learned structure (e.g.,\enquote{IF feature $x_1 > 0$ and feature $x_2 \in \{A,B\}$, THEN predict 0.6}).
+We can interpret the learned structure to trace how the model makes predictions.
+Rule-based ML also remains an active area of research.
+%For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
+The more complex interpretable models get (e.g., linear models with hundreds of features and complex interaction terms or deep decision trees), the more difficult the interpretation becomes.
 
-\paragraph{Local explanations} explain how individual predictions are derived.
-Here we have a great variety of methods, with very different methodological motivations.
-Popular methods are LIME, Shapley Values and Counterfactual Explanations.
-I want to highlight especially their very different approaches to the problem of local explanations.
-Counterfactual explanations are what-if scenarios.
-How does an instance have to change to get another (desired) prediction or classification.
-They build on a rich tradition in philosophy and also thrive in what the social sciences discovered:
-Good explanations are contrastive and focus on a few reasons.
+Even for complex models such as deep neural networks, introspection can be a means of interpretation.
+For example, the abstract features learned by deep convolutional neural networks can be visualized by finding or generating images that activate a feature map of the CNN the most \cite{olah2017feature}.
+Feature visualization showed that CNNs learn low-level features such as edge-detectors at the early neural layers and gradually more abstract detectors for textures and objects in higher layers.
+%CNN feature visualization lead to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
+In random forests, the minimal depth distribution \cite{randomForestExplainer,ishwaran2010high}, analyzes at which tree depth a feature is used, which can be interpreted as importance of a feature.
+Minimal tree depth is another example of analyzing the structure of a model to get an understanding of how it makes decisions.
+%The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
+%By using our distinction of introspection and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
+%Gini importance collects for each feature and tree how the split decreased the Gini impurity.
+%It does not rely on making predictions, but only on internal statistics.
+%For mean decrease accuracy, we probe the random forest with perturbed data, i.e., we shuffle a feature and see how the prediction changes.
+%It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
+%Model-agnostic versions of this permutation feature importance exists \cite{fisher2019all}.
+Since all IML methods that rely on introspection are model-specific, the explanations are tied to that ML model.
+If a different ML model is used, the IML method has to change, which is a disadvantage.
+But if a model is well understood and frequently used in community, like random forest in ecology research \cite{cutler2007random}, model introspection can be a powerful tool.
+
+\paragraph{IML method based on sensitivity analysis} often treat the ML model as a closed system that receives an input and produces an output, the prediction.
+We further distinguish between local and global explanations that sensitivity based IML methods can produce.
+
+\paragraph{Local explanations} are used to explain individual predictions of an ML model.
+Local explanation methods have received a lot of attention and there has been a lot of innovation in the last years.
+Popular local IML methods are LIME \cite{ribeiro2016should}, Shapley Values \cite{lundberg2017unified,vstrumbelj2014explaining} and Counterfactual Explanations \cite{wachter2017counterfactual,dandl2020multi}.
+Counterfactual explanations explain predictions in the form of what-if scenarios, which builds on a rich tradition in philosophy.
+According to findings in the social sciences \cite{miller2019explanation}, counterfactual explanations are \enquote{good} explanations because they are contrastive and focus on a few reasons.
 A very different approach originates from collaborative game theory: The Shapley Values.
 How do you fairly share some payout from a game among the players?
-The Shapley values \cite{shapley1953value} provide an answer for this question, by trying out all constellations of players and averaging the additional payout a player would add to the total payout when the player is added to a team.
-The same idea can be applied to ML: The predicted value is the payout, the player is a certain feature value.Many variations, extensions and more efficient implementations for specific ML models exist \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}.
-The third mention here are local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}
-LIME is a surrogate model, i.e., the procedure approximates the original ML model with a simpler interpretable model.
-LIME does so locally, meaning only in some neighborhood around the data point that we want to explain.
+The Shapley values \cite{shapley1953value} provide an answer to how to fairly share a payout among the players of a collaborative game.
+The collaborative game idea can be applied to ML \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}: The predicted value is the payout, each player is a certain feature value.
+Another popular method is local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}
+LIME uses surrogate models, i.e., the method trains an interpretable model such as a liner regression model to approximate the predictions of original ML model.
 
-\paragraph{Surrogate Models} are an interesting and huge class of explanation method \cite{puri2017magix,molnar2019,ming2018rulematrix}.
-This approach is very common in sensitivity analysis where we have emulation? models.
-Many methods -- a confusing amount, really -- are nothing but simple models that approximate the behavior of the (complex) model to be explained.
-Surrogate models combine the two strategies:
-First, approximate the predictions of the ML model with an interpretable model, such as a decision tree.
-This is a kind of sensitivity-like approach, where we see the original model as a black box and artificially probe the ML model and derive information from the behavior.
-Then we interpret the interpretable model, which now acts as a surrogate to the more complex ML model.
+\paragraph{Global, model-agnostic explanation methods} are used to explain how the model behaves on average for some given dataset.
+A useful distinction of global explanations are feature importance and feature effect.
+Feature importance ranks features based on how relevant they were for the prediction.
+Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
+An alternative are variance based measures.
+See \cite{wei2015variable} for an overview of all the ways to measure importance.
+The feature effect expresses how a change in a feature changes the predicted outcome.
+Popular feature effect plots are Partial Dependence Plots \cite{friedman2001greedy}, Individual Conditional Expectation Plots \cite{goldstein2015peeking} and Accumulated Local Effect  (ALE)  Plots \cite{apley2016visualizing}.
+
+\paragraph{Surrogate models}\footnote{Surrogate models are related to knowledge distillation and the teacher-student model} are interpretable models designed to \enquote{copy} the behavior of the ML model.
+They do not neatly fit into the categorization of sensitivity analysis or introspection, as they combine both:
+The surrogate approach treats the ML model as a black box, only requiring the input / output data  (similar to sensitivity analysis), but the interpretation is based on introspection of the interpretable surrogate model.
+Many methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting}.
+They differ in the ML model they can be used for, the data sampling strategy, the interpretable model that is used and so on.
 There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which would be more the introspection approach.
 
-\paragraph{Global, model-agnostic explanation methods} are also very popular.
-Here we have have feature importance, which rank features based on how relevant they were for the prediction and the feature effect, which expresses how a change in that feature changes the predicted outcome on average.
-Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
-Alternatives are variance based measures, see \cite{wei2015variable} for an overview of all the ways to measure importance.
-For feature effects we have Partial Dependence Plots \cite{friedman2001greedy}, Individual Conditional Expectation Plots \cite{goldstein2015peeking} and Accumulated Local Effect  (ALE)  Plots \cite{apley2016visualizing}.
+%\paragraph{Blending all together.} This list was not exhaustive. For example, there are also ways to add interpretability constraints to a model, such as monotonicity or sparsity.
+%This can helps introspection, e.g., a linear model with most coefficients at zero is easier to interpret.
+%But this can also help with sensitivity analysis, as PDPs of monotonic features will only show in one direction.
 
-\section{Challenges}
+\section*{Challenges}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 

### 2020-09-10 2020-ecml-xkdd/paper/paper.tex
adds some improvements

@@
 %
-\title{Interpretable Machine Learning Methods -- A Brief History, State of the Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
+\title{Interpretable Machine Learning -- A Brief History, State-Of-The-Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
 01IS18036A.
@@
 %
-\titlerunning{IML - History, State of the Art, Challenges}
+\titlerunning{IML - History, Methods, Challenges}
 % If the paper title is too long for the running head, you can set
@@
 With interpretable machine learning (IML) \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangably} the user can find out problems of the ML model, extract insights and justify decisions.
+This short paper views IML through a methodological viewpoint, with an emphasis on \textbf{what can we, mathematically and algorithmically, derive from an ML model} and less on the human part.
 %I want to give three very common use cases:
@@
 %An interpretation method is model-specific if it only works for a specific ML model, it is model-agnostic if it can be applied to any model.
-We distinguish IML methods between introspection and sensitivity analysis\footnote{Not to be confused with the research field of sensitivity analysis, which studies the uncertainty of outputs in mathematical models and systems.
+We distinguish IML methods between component and sensitivity analysis\footnote{Not to be confused with the research field of sensitivity analysis, which studies the uncertainty of outputs in mathematical models and systems.
 But there are methodological overlaps (e.g., Shapley values)}.
-An IML method that works by introspection assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.
+An IML method that works by component analysis assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.\footnote{Some surveys \cite{vilone2020explainable,guidotti2018survey} divide methods into \textit{ante-hoc (alternatively: transparent design, white box models)} and \textit{post-hoc} IML method, depending on whether interpretability is considered at model design and training or after training, leaving the (black-box) model unchanged.
+Another category divides model-agnostic and model-specific.
+Component vs.
+sensitivity analysis is, in our opinion, another interesting view and why only repeat what others already have said?}
+The idea of component analysis is closely related to decomposability:
+You can only analyse components of a model that can be decomposed into individual components that we can interpret individually.
+Component analysis does not necessarily require that the user understands the model in its entirety (simulatability).
 The introspection is always model-specific, e.g., a linear regression model can be interpreted by analyzing its coefficients, while a decision tree can be interpreted by visualizing the learned decision structure.
@@
 \paragraph{Inherently interpretable models} are models with interpretable, learned structures and parameters (introspection).
+Linear regression models, decision trees and decision rules are considered to be interpetable \cite{freitas2014comprehensible,huysmans2011empirical}.
 %As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
@@
 Rule-based ML also remains an active area of research.
-%For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "If feature $x_1$ larger than two and feature $x_4$ in category A, then prediction is 10".
+%For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "IF condition 1 AND condition 2, THEN predict \textit{outcome}".
 The more complex interpretable models get (e.g., linear models with hundreds of features and complex interaction terms or deep decision trees), the more difficult the interpretation becomes.
 
-Even for complex models such as deep neural networks, introspection can be a means of interpretation.
+Component analysis is how we interpret inherently interpretable models, but component analysis can be used to explain black box models.
+Even for complex models such as deep neural networks, component analysis can be a means of interpretation.
 For example, the abstract features learned by deep convolutional neural networks can be visualized by finding or generating images that activate a feature map of the CNN the most \cite{olah2017feature}.
@@
 %The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
-%By using our distinction of introspection and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
+%By using our distinction of component and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
 %Gini importance collects for each feature and tree how the split decreased the Gini impurity.
@@
 %Model-agnostic versions of this permutation feature importance exists \cite{fisher2019all}.
-Since all IML methods that rely on introspection are model-specific, the explanations are tied to that ML model.
+Since all IML methods that rely on component analysis are model-specific, the explanations are tied to that ML model.
 If a different ML model is used, the IML method has to change, which is a disadvantage.
-But if a model is well understood and frequently used in community, like random forest in ecology research \cite{cutler2007random}, model introspection can be a powerful tool.
+But if a model is well understood and frequently used in community, like random forest in ecology research \cite{cutler2007random}, model component analysis can be a powerful tool.
+A component analysis can explain the overall model behavior (global) or individiual predictions (local).
+In many cases of component analysis global and local analysis collapses to the same thing.
+For example in linear models the weights explain both how the features affect an individual prediction, but also the average, global behavior.
 
@@
 \paragraph{Surrogate models}\footnote{Surrogate models are related to knowledge distillation and the teacher-student model} are interpretable models designed to \enquote{copy} the behavior of the ML model.
-They do not neatly fit into the categorization of sensitivity analysis or introspection, as they combine both:
-The surrogate approach treats the ML model as a black box, only requiring the input / output data  (similar to sensitivity analysis), but the interpretation is based on introspection of the interpretable surrogate model.
-Many methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting}.
+They do not neatly fit into the categorization of component or sensitivity analysis, as they combine both:
+The surrogate approach treats the ML model as a black box, only requiring the input / output data  (similar to sensitivity analysis), but the interpretation is based on analyzing components of the interpretable surrogate model.
+Many methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting,craven1996extracting,krishnan2017palm}.
 They differ in the ML model they can be used for, the data sampling strategy, the interpretable model that is used and so on.
-There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which would be more the introspection approach.
+There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which would be more the component analysis approach.
+
+What I missed out on:
+
+- Human-Computer Interaction, e.g. (Explainable Artificial Intelligence: a Systematic Review) sees Explainable AI as the intersection between Artificial Intelligence and Human-Computer Interaction.
+- Ethics and Philosophy.
+- Explanations can have different \textit{structure}: text, image, tables, ...
+- Methods that are model-agnostic might not be input data agnostic, e.g., PDP does not make sense for pixel data as used in image classification
+- Some methods are designed or adapted so that we can more easily understand what its components do.
+For example for neural networks we can \textit{disentangle} the invidual neurons TODO:CITE, so that we can interpret them better.
+- TODO: Add more to the idea of making components of a (black-box) ML model more interpretable. This can be e.g. tree pruning, adding sparsity to linear regression models, disentangle neurons.
+- We can use sensitivity analysis also to interpretable models
+- We did not touch topics where data is analyzed before modeling to understand the data better
+- Idea that we can decompose the function with functional ANOVA and so on
 
 %\paragraph{Blending all together.} This list was not exhaustive. For example, there are also ways to add interpretability constraints to a model, such as monotonicity or sparsity.
-%This can helps introspection, e.g., a linear model with most coefficients at zero is easier to interpret.
+%This can helps component analysis, e.g., a linear model with most coefficients at zero is easier to interpret.
 %But this can also help with sensitivity analysis, as PDPs of monotonic features will only show in one direction.
@@
 We do not know what the groundtruth explanation looks like and have no straightforward mean to quantify how interpretable a model is or how correct an explanation is.
-Instead of having one groundtruth explanation, various \textbf{quantifiable aspects of interpretability} are emerging.\cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing}
+Instead of having one groundtruth explanation, various \textbf{quantifiable aspects of interpretability} are emerging \cite{poursabzi2018manipulating,philipp2018measuring,molnar2019quantifying,hauenstein2018computing,zhou2018measuring,akaike1998information,schwarz1978estimating,poursabzi2018manipulating,dhurandhar2017tip,friedler2019assessing}.
+
+Two ways of evaluation of interpretability have emerged: \textit{objective evaluations} which are mathematically quantifiable metrics, and {human-centered evaluations} which involve studies with users.
 Sparsity of the explanation (or the model itself), interaction strength of features within the ML model and
-It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength and a user's ability to run a model on a given input (simulatability) are examples of such aspects.
+It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, fidelity (how well explanation approximates the ML model), sensitivity to perturbations,  and a user's ability to run a model on a given input (simulatability) are examples of such aspects.
 The challenge ahead remains to establish a best practice on how to evaluate interpretation methods and the explanations they produce.
+Here we should also look to the field of Human-Computer Interaction who do exactly that.
 

### 2020-09-13 2020-ecml-xkdd/paper/paper.tex
Updates from Overleaf

@@
 \usepackage{csquotes}
-
+% TODO: Strongly short the paper.
+% TODO: Add some images. Add sensitivity analysis vs. component analysis
 \begin{document}
@@
 \begin{abstract}
-  I present a very short history of the field of interpretable machine learning (IML), give an overview of popular interpretation methods and discuss challenges when interpreting machine learning models.
-  This view might be a bit biased through the eyes of a statistician.
+  We present a brief history of interpretable machine learning (IML), give an overview of state-of-the-art interpretation methods and discuss challenges when interpreting machine learning models.
+  We focus on the IML methods, putting emphasis on what interpretations can be derived mathematically and algorithmically from a machine learning model.
 
-\keywords{Interpretable Machine Learning \and Explainable AI}
+\keywords{Interpretable Machine Learning \and Explainable Artificial Intelligence}
 \end{abstract}
@@
 %
-Interpretability is often a deciding factor when machine learning (ML) is used in a product, a decision process or in research.
-With interpretable machine learning (IML) \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangably} the user can find out problems of the ML model, extract insights and justify decisions.
-This short paper views IML through a methodological viewpoint, with an emphasis on \textbf{what can we, mathematically and algorithmically, derive from an ML model} and less on the human part.
-%I want to give three very common use cases:
-%\paragraph{For science:} In applied sciences, data is often modeled with ML nowadays, where before statistical modeling approach would have been used.
-%The goal of science is not only to predict the world well, but also to generate further insights.
-%Examples are species distribution models in ecology which aim to explain why certain species can be found on certain areas based on e.g. the local climate.
-%Here, interpretability of the prediction model is needed to understand why a species, e.g., has a high probability to occur in a specific reason.
+Interpretability is often a deciding factor when a machine learning (ML) model is used in a product, a decision process or in research.
+Interpretable machine learning (IML) methods \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangeably} can be used to discover knowledge, justify the model and its predictions, control and improve models \cite{adadi2018peeking}.
+This paper reviews IML from a methodological viewpoint, with an emphasis on \textbf{what can we derive mathematically and algorithmically from an ML model}.
 
-\paragraph{A Very Short History of IML.}
-A lot of IML research just happened in the last five years.
+\section{A Brief History of IML}
+A lot of IML research happened in the last couple of years.
 But learning interpretable models from data has a much longer tradition.
 Linear regression models have been used back in 1800 by Gauss, Legendre and Quetelet \cite{stigler1986history,legendre1805nouvelles,gauss1809theoria,quetelet1827recherches}.
-Rule-based machine learning (e.g., decision rules and trees) has been an active research areas since the 1980s \cite{furnkranz2012foundations} with some roots reaching back into the 1960s \cite{hajek1966guha}.
-Both research on linear regression models and its extensions and rule-based ML remain important and busy research areas to this day and are even blending together (e.g.
-model-based trees \cite{zeileis2008model} or RuleFit \cite{friedman2008predictive}).
-Machine learning research in general also started around that time, e.g., support vector machines in 1974 \cite{vapnik1974theory}, early important work on neural networks in the 1960s and boosting in 1990 \cite{schapire1990strength}.
+Rule-based machine learning (e.g., decision rules and trees) has been an active research area since the 1980s \cite{furnkranz2012foundations}.
+Both research on regression analysis and rule-based ML remain important and busy research areas to this day and are even blending together (e.g., model-based trees \cite{zeileis2008model}, RuleFit \cite{friedman2008predictive}).
+Machine learning research in general also started around that time, e.g., support vector machines in 1974 \cite{vapnik1974theory}, early important work on neural networks in the 1960s \cite{schmidhuber2015deep} and boosting in 1990 \cite{schapire1990strength}.
 Interpretability has always been a companion to ML.
 A good example for this is the random forest, which I also see as an IML milestone for making ML practical.
-Random forests, a black box machine learning model that often works well without tuning, came with a built-in feature importance measure, which contributed to the success.
-The evidence that interpretability was success factor are the many citations (>60000 citation on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also of papers improving the importance measure (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}).
+Random forests, a black box machine learning model that often works well without tuning, came with a built-in feature importance measure, which contributed to its success.
+The evidence that interpretability was success factor are the many citations ($>60000$ citations on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also of papers improving the importance measure (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}).
 In the 2010s came the deep learning hype, after a deep neural network won the ImageNet challenge.
@@
 Since then, many approaches have been introduced to the field, many of them model-agnostic explanation methods, i.e., working for different types of ML models, but also model-specific explanation methods to interpret, e.g., deep neural networks and tree ensembles.
-Rule-based ML and statistical regression models continue to be developed.
+Rule-based ML and statistical regression models continue to be developed and serves both as full ML models, but also as building blocks for many of the newer IML approaches.
 %Could it be that only the terms have changed, but the field has always been there and the same?
 %I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
+% TODO: Find more evaluation papers
+\textbf{Today} IML reached a first state of usability.
+Some methods have been established with a better understanding of weaknesses and strengths.
+There are dozens of surveys of the IML field \cite{molnar2019,guidotti2018survey,vilone2020explainable,rosenfeld2019explainability,adadi2018peeking,anjomshoae2019explainable,du2019techniques,carvalho2019machine}.
+The field is starting to further consolidate terms and knowledge \cite{hall2019systematic,doshi2017towards,murdoch2019definitions,samek2019towards,preece2018stakeholders} and we have some ideas how to evaluate explanations and interpretability methods \cite{mohseni2018multidisciplinary}.
+Many open source software packages implement various techniques \cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml} .
+These software tools give ML practitioners the means to interpret their models.
+Regulation such as GDPR has spurred a discussion around further needs of interpretability \cite{wachter2017counterfactual}.
+IML has also arrived in industry \cite{gade2019explainable}, there are startups that focus on machine learning interpretability and also big tech companies offer software \cite{exler2019if,arya2020ai,hall2017machine}.
 
-\paragraph{Today} IML reached a first state of usability.
-Some methods have been established with a better understanding of weaknesses and strengths.
-There is a lot of open source software that implement various techniques (\cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml}, ...) and also big tech companies offer software \cite{exler2019if,arya2020ai,hall2017machine}.
-These software tools gives ML practitioners the means to interpret their models.
-Regulation such as GDPR has spurred a discussion around further needs of interpretability.
-Many startup are now focused on interpretable ML.
-
-\section*{IML Methods}
+\section{IML Methods}
 
 In this section we broadly cover some popular IML methods.
-%Often a distinction is made between model-specific and model-agnostic explanation methods.
-%An interpretation method is model-specific if it only works for a specific ML model, it is model-agnostic if it can be applied to any model.
 We distinguish IML methods between component and sensitivity analysis\footnote{Not to be confused with the research field of sensitivity analysis, which studies the uncertainty of outputs in mathematical models and systems.
-But there are methodological overlaps (e.g., Shapley values)}.
-An IML method that works by component analysis assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.\footnote{Some surveys \cite{vilone2020explainable,guidotti2018survey} divide methods into \textit{ante-hoc (alternatively: transparent design, white box models)} and \textit{post-hoc} IML method, depending on whether interpretability is considered at model design and training or after training, leaving the (black-box) model unchanged.
+There are methodological overlaps (e.g., Shapley values), but also differences in methods and how input data distributions are handled}.
+An IML method that works by component analysis assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.\footnote{Some surveys \cite{vilone2020explainable,guidotti2018survey,du2019techniques} divide methods into \textit{ante-hoc (alternatively: transparent design, white box models)} and \textit{post-hoc} IML method, depending on whether interpretability is considered at model design and training or after training, leaving the (black-box) model unchanged.
 Another category divides model-agnostic and model-specific.
@@
 The idea of component analysis is closely related to decomposability:
-You can only analyse components of a model that can be decomposed into individual components that we can interpret individually.
-Component analysis does not necessarily require that the user understands the model in its entirety (simulatability).
-The introspection is always model-specific, e.g., a linear regression model can be interpreted by analyzing its coefficients, while a decision tree can be interpreted by visualizing the learned decision structure.
-Sensitivity analysis is mostly model-agnostic and works by manipulating input data and analyzing ML model  predictions.
-Some IML sensitivity analysis methods rely on model-specific knowledge such as gradients of neural networks.
-An example are the various heatmap-like explanation for CNNs for image classification \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not} which study the sensitivity of the prediction to the input pixels.
-Technically this is done by backpropagation and therefore model-specific.
-Model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist.
-
-%The other useful category is local vs. global explanations.
-%Local means that the method explains individual predictions, and global means it explains the average model behavior.
-%For inherently interpretable models, global and local explanations are often the same: In linear models, the coefficients both how the model behaves globally, but also is used for individuals instances.
-
-\paragraph{Inherently interpretable models} are models with interpretable, learned structures and parameters (introspection).
-Linear regression models, decision trees and decision rules are considered to be interpetable \cite{freitas2014comprehensible,huysmans2011empirical}.
+You can only analyze components of a model that can be decomposed into individual components that we can interpret individually, like weights in a linear model.
+Component analysis does not necessarily require that the user understands the model in its entirety (simulatability), but only individual parts.
+Component analysis is always model-specific, e.g., a linear regression model can be interpreted by analyzing its coefficients, while a decision tree can be interpreted by visualizing the learned decision structure.
+Sensitivity analysis is mostly model-agnostic and works by manipulating input data and analyzing the respective model predictions.
+Usually, we analyze the components of ML models that are said to be inherently interpretable, but there are also ways to analyze components of more complex models, but this often requires more work.
+\textit{Inherently interpretable models} are models with learned structures and learned parameters which can be assigned some meaning.
+Linear regression models, decision trees and decision rules are considered to be interpretable \cite{freitas2014comprehensible,huysmans2011empirical}.
 %As a rule of thumb for when we do an introspection is that we are not probing the model with various (manipulated data points).
-Linear regression models ($Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$) can be interpreted by introspection:
+Linear regression models ($Y = \beta_0 + \beta_1 x_1 + \ldots + \beta_p x_p$) can be interpreted by analyzing components:
 The model structure, a weighted sum of features, allows that we interpret the $\beta$-weights as the effect that a feature has on the prediction.
-Many extensions exists and linear regression models are remain an active field of research.
+Many extensions exist \cite{hastie1990generalized,fahrmeir2013multivariate,gelman2006data} and linear regression models remain an active field of research \cite{fasiolo2020scalable,caruana2015intelligible,fasiolo2020fast,ustun2016supersparse}
 Decision trees and other rule-based ML models have a learned structure (e.g.,\enquote{IF feature $x_1 > 0$ and feature $x_2 \in \{A,B\}$, THEN predict 0.6}).
 We can interpret the learned structure to trace how the model makes predictions.
-Rule-based ML also remains an active area of research.
-%For example, an individual prediction can be traced through a decision tree and we retrieve an explanation of the form "IF condition 1 AND condition 2, THEN predict \textit{outcome}".
+Rule-based ML also remains an active area of research (for example, \cite{wang2015falling,letham2015interpretable,hothorn2015ctree}).
+
 The more complex interpretable models get (e.g., linear models with hundreds of features and complex interaction terms or deep decision trees), the more difficult the interpretation becomes.
@@
 Component analysis is how we interpret inherently interpretable models, but component analysis can be used to explain black box models.
-Even for complex models such as deep neural networks, component analysis can be a means of interpretation.
-For example, the abstract features learned by deep convolutional neural networks can be visualized by finding or generating images that activate a feature map of the CNN the most \cite{olah2017feature}.
+For example, the abstract features learned by a deep convolutional neural network (CNN) can be visualized by finding or generating images that activate a feature map of the CNN \cite{olah2017feature}.
+% TODO: Add image
 Feature visualization showed that CNNs learn low-level features such as edge-detectors at the early neural layers and gradually more abstract detectors for textures and objects in higher layers.
-%CNN feature visualization lead to the famous results where we find out that CNNs learn edge detectors at the lowest levels and increasingly abstract concepts at higher layers.
-In random forests, the minimal depth distribution \cite{randomForestExplainer,ishwaran2010high}, analyzes at which tree depth a feature is used, which can be interpreted as importance of a feature.
-Minimal tree depth is another example of analyzing the structure of a model to get an understanding of how it makes decisions.
-%The random forest is also a useful example for another measure, the feature importance, for which various measures exist, such as decrease in accuracy and decrease in Gini impurity.
-%By using our distinction of component and sensitivity analysis Gini impurity would fall into the former while mean decrease accuracy falls into the latter.
-%Gini importance collects for each feature and tree how the split decreased the Gini impurity.
-%It does not rely on making predictions, but only on internal statistics.
-%For mean decrease accuracy, we probe the random forest with perturbed data, i.e., we shuffle a feature and see how the prediction changes.
-%It's a model-specific methods since we do this tree-wise and use the out-of-bag sample of the tree, so this procedure is cleverly adapted to the random forest.
-%Model-agnostic versions of this permutation feature importance exists \cite{fisher2019all}.
+As one other example among many, there are various feature importance measures for the random forest that are based on analyzing components.
+For example, the minimal depth distribution \cite{randomForestExplainer,ishwaran2010high} and the Gini importance \cite{breiman2001random} analyze the structure of the trees of the forest.
+.
 Since all IML methods that rely on component analysis are model-specific, the explanations are tied to that ML model.
@@
 But if a model is well understood and frequently used in community, like random forest in ecology research \cite{cutler2007random}, model component analysis can be a powerful tool.
-A component analysis can explain the overall model behavior (global) or individiual predictions (local).
-In many cases of component analysis global and local analysis collapses to the same thing.
-For example in linear models the weights explain both how the features affect an individual prediction, but also the average, global behavior.
+Component analysis can explain the overall model behavior (global) or individual predictions (local).
+In many cases the components of a model has the same global and local interpretation.
+For example, in linear regression models the weights explain both how the features affect an individual prediction, but also summarize the global behavior.
 
@@
 \paragraph{Local explanations} are used to explain individual predictions of an ML model.
-Local explanation methods have received a lot of attention and there has been a lot of innovation in the last years.
+Local explanation methods have received much attention and there has been a lot of innovation in the last years.
 Popular local IML methods are LIME \cite{ribeiro2016should}, Shapley Values \cite{lundberg2017unified,vstrumbelj2014explaining} and Counterfactual Explanations \cite{wachter2017counterfactual,dandl2020multi}.
@@
 According to findings in the social sciences \cite{miller2019explanation}, counterfactual explanations are \enquote{good} explanations because they are contrastive and focus on a few reasons.
-A very different approach originates from collaborative game theory: The Shapley Values.
-How do you fairly share some payout from a game among the players?
-The Shapley values \cite{shapley1953value} provide an answer to how to fairly share a payout among the players of a collaborative game.
-The collaborative game idea can be applied to ML \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}: The predicted value is the payout, each player is a certain feature value.
-Another popular method is local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}
-LIME uses surrogate models, i.e., the method trains an interpretable model such as a liner regression model to approximate the predictions of original ML model.
+A different approach originates from collaborative game theory:
+The Shapley values \cite{shapley1953value} provide an answer on how to fairly share a payout among the players of a collaborative game.
+The collaborative game idea can be applied to ML \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}: The predicted value is the payout, each feature value is a player.
+Another popular method are local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}.
+LIME uses surrogate models, i.e., the method trains an interpretable model such as a linear regression model to approximate the predictions of original ML model.
+Numerous extensions of LIME exist, that try to fix issues with the original method, extend it to other tasks and data or analyze its properties \cite{hu2018locally,rabold2018explaining,rabold2019enriching,visani2020optilime,haunschmid2020audiolime,rahnama2019study,shankaranarayana2019alime}.
+% TODO: Add the many extensions of Shapley Values
+Some IML methods rely on model-specific knowledge to analyze how changes in the input features changes the output.
+Saliency maps for CNNs make use of the network gradients to explain individual classifications.
+The explanations are in the form of heatmaps that show how changing a pixel can change the classificaiton.
+The saliency map methods differ in how they backpropagate \cite{sundararajan2017axiomatic,lundberg2017unified,montavon2017explaining,simonyan2013deep,shrikumar2016not}.
+Additionally model-agnostic versions \cite{ribeiro2016should,lundberg2017unified,zeiler2014visualizing} exist for analyzing image classifiers.
 
@@
 Feature importance ranks features based on how relevant they were for the prediction.
-Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
+Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is a popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
 An alternative are variance based measures.
@@
 
-\paragraph{Surrogate models}\footnote{Surrogate models are related to knowledge distillation and the teacher-student model} are interpretable models designed to \enquote{copy} the behavior of the ML model.
+\paragraph{Surrogate models}\footnote{Surrogate models are related to knowledge distillation and the teacher-student model.} are interpretable models designed to \enquote{copy} the behavior of the ML model.
 They do not neatly fit into the categorization of component or sensitivity analysis, as they combine both:
@@
 Many methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting,craven1996extracting,krishnan2017palm}.
-They differ in the ML model they can be used for, the data sampling strategy, the interpretable model that is used and so on.
-There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which would be more the component analysis approach.
+They differ in which ML model they can be used for, the data sampling strategy, the interpretable model that is used and so on.
+There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which resembles more the component analysis approach.
 
-What I missed out on:
+% TODO: Add that sensitivity analysis can also be used for inherently interpreteable models
 
-- Human-Computer Interaction, e.g. (Explainable Artificial Intelligence: a Systematic Review) sees Explainable AI as the intersection between Artificial Intelligence and Human-Computer Interaction.
-- Ethics and Philosophy.
-- Explanations can have different \textit{structure}: text, image, tables, ...
-- Methods that are model-agnostic might not be input data agnostic, e.g., PDP does not make sense for pixel data as used in image classification
-- Some methods are designed or adapted so that we can more easily understand what its components do.
-For example for neural networks we can \textit{disentangle} the invidual neurons TODO:CITE, so that we can interpret them better.
-- TODO: Add more to the idea of making components of a (black-box) ML model more interpretable. This can be e.g. tree pruning, adding sparsity to linear regression models, disentangle neurons.
-- We can use sensitivity analysis also to interpretable models
-- We did not touch topics where data is analyzed before modeling to understand the data better
-- Idea that we can decompose the function with functional ANOVA and so on
+%What I missed out on:
+%
+%- Human-Computer Interaction, e.g. (Explainable Artificial Intelligence: a Systematic Review) sees Explainable AI as the intersection between Artificial Intelligence and Human-Computer Interaction.
+%- Ethics and Philosophy.
+%- Explanations can have different \textit{structure}: text, image, tables, ...
+%- Methods that are model-agnostic might not be input data agnostic, e.g., PDP does not make sense for pixel data as used in image classification
+%- Some methods are designed or adapted so that we can more easily understand what its components do.
+For example for neural networks we can \textit{disentangle} the individual neurons TODO:CITE, so that we can interpret them better.
+%- TODO: Add more to the idea of making components of a (black-box) ML model more interpretable. This can be e.g. tree pruning, adding sparsity to linear regression models, disentangle neurons.
+%- We can use sensitivity analysis also to interpretable models
+%- We did not touch topics where data is analyzed before modelling to understand the data better
+%- Idea that we can decompose the function with functional ANOVA and so on
 
@@
 
-\section*{Challenges}
+\section{Challenges}
 
@@
 
-\paragraph{Feature dependence:} When two features share information (e.g., BMI and weight) separating their effects and importances is difficult.
-Some IML methods (e.g.LIME, permutation feature importance or Shapley values) simply \enquote{ignore} such dependence, which results in extrapolation, i.e., for the interpretation new, unrealistic data points are created (e.g., low weight but high BMI).
-Interpretations based on extrapolation into areas where the model never saw data can be very misleading \cite{hooker2019please}.
-There have been attempts to \enquote{fix} IML methods by creating only realistic data points (using the conditional instead of the marginal distribution).
-These conditional IML methods fix the extrapolation problem, but they also entangle the interpretations of a feature's effect and importance with all other features it depends on.
-In the case of Shapley values the conditional variant even breaks the interpretation \cite{sundararajan2019many,janzing2019feature}.
+\paragraph{Statistical Uncertainty and Inference}
+Many IML methods such as permutation feature importance or Shapley values provide explanations without quantifying the uncertainty of that explanation.
+The the model itself, but also the explanations are usually computed from data and are subject to variance.
+We have to become more rigorous in reporting both reporting the explanations, but also their uncertainty.
+Some research is working towards the goal of quantifying uncertainty of explanations, for example for feature importance \cite{watson2019testing,fisher2019all}, layer-wise relevance propagation \cite{fabi2020feature} and Shapley values \cite{williamson2020efficient}.
+Quantifying uncertainty and expressing it as error bars, confidence intervals or also test statistics and p-values is an important part of statistical modeling.
+The goal of statistical inference is to infer properties of the underlying data distribution of the phenomenon that is being studied.
+In this process, a statistical model is proposed with a set of assumptions (for example, a linear model with assumptions that the features only have additive effects).
+When all assumptions are met, the statistical model may be interpreted to reflect the true data generating process.
+Statistical modeling is used for testing scientific hypothesis.
+Machine learning modeling takes a different approach:
+Instead of assuming a model of the data generating process, various flexible models are compared on test data.
+It is desirable to draw also conclusions from these models about the underlying data generating process.
+With IML, with have method to compute meaningful properties of the model. 
+But to infer properties about the data generating process from the ML model, we have to specify the assumptions needed that allow a combination of ML model and IML method to be interpreted as the \enquote{true} model.
+We also need tools and procedures to check these assumptions.
+Hopefully we can avoid problems in statistical testing, such as p-hacking \cite{head2015extent} and the resulting publication bias \cite{begg1994publication}.
 
-\paragraph{Uncertainty quantification} is a typical part of statistical modeling and becomes visible as error bars, confidence intervals, test statistics and p-values.
-Uncertainty quantification helps to separate signal from noise.
-Many IML methods only return the point estimates, but don't quantify the uncertainty.
-This reaches across most methods: rule-based ML , feature effect plots, feature importance, ...
-There is research about uncertainty quantification of model interpretations, e.g., feature importance \cite{watson2019testing,fisher2019all}, layer-wise relevance propagation \cite{fabi2020feature} and Shapley values \cite{williamson2020efficient}.
-However, it is not implemented as default in most IML software.
-Also we need a more united view on the topic, and specify which uncertainty is quantified (model bias, model variance, IML estimation error, ...) and to what end.
-The statistics community is very deep into statistical inference and IML can learn a lot.
-Here we have to dig deeper into the statistics literature which can be our guide.
-
-\paragraph{Statistical inference:}
-When uncertainty quantification is combined with assumptions about the data generating process, we can draw conclusions about the real world.
-But for that we have to better understand the properties and assumptions of the various IML methods.
-For example, methods such as partial dependence plots require that features are independent.
-To enable statistical inference, there are at least two challenges to overcome: 1) define the assumptions needed that allow a model (and an IML method) to be interpreted as the \enquote{true} model and 2) tools and procedures to check these assumptions.
-
-\paragraph{Causal interpretation}
+\paragraph{Causal interpretation.}
 Ideally, a model relies on the true causes to make a prediction.
-Then we could also interpret it causally, making statements about the underlying phenomenon, which is the focus of ML in research, but also makes models robust against adversarial attacks, but also makes the model more usefule when used as a basis for decision making.
+Then we could  interpret it causally, making statements about the underlying phenomena, which is the focus of ML in research, but also makes models robust against adversarial attacks, and more useful when used as a basis for decision making.
 Unfortunately, predictive performance and causality can be conflicting goals.
@@
 Further research is needed to understand when we are allowed to make causal interpretations of an ML model.
+First steps have been made for permutation feature importance \cite{konig2020relative} and Shapley values \cite{ma2020predictive}.
 
 \paragraph{Lack of definition} is a common critique of the field \cite{lipton2018mythos,doshi2017towards}.
-Without a definition of "interpretability", how can we decide if a new method is better at explanation an ML method?
+Without a definition of "interpretability", how can we decide if a new method explains ML models better?
 To evaluate the \textbf{predictive performance} of an ML model, we simply compute the prediction error on test data given the groundtruth label.
@@
 
-Two ways of evaluation of interpretability have emerged: \textit{objective evaluations} which are mathematically quantifiable metrics, and {human-centered evaluations} which involve studies with users.
-Sparsity of the explanation (or the model itself), interaction strength of features within the ML model and
+Two main ways of evaluation of interpretability have emerged: \textit{objective evaluations} which are mathematically quantifiable metrics, and {human-centered evaluations} which involve studies with users.
 It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, fidelity (how well explanation approximates the ML model), sensitivity to perturbations,  and a user's ability to run a model on a given input (simulatability) are examples of such aspects.
@@
 We have focused mostly on the methodological, mathematical challenges in a rather static setting, where you are given a machine learning model and data.
-\paragraph{Many more} challenges lie ahead to improve interpretability of machine learning.
+\textit{Many more} challenges lie ahead to improve interpretability of machine learning.
 What is the best way to explain a machine learning model for people with a specific background and how do we present the explanations?
@@
 
-\section*{Discussion}
+\section{Discussion}
 
-Interpretable Machine Learning is a young field, but has deeper roots in Statistics and Computer Science and also draws from other fields such as the social sciences.
-There has been Cambrian explosion of methods since 2015, including model-agnostic interpretation methods and many specialised methods for, e.g., deep neural networks.
+Interpretable Machine Learning is a young field, but has deeper roots in Statistics and Computer Science and also draws from other fields such as the social sciences, human-computer interaction and game theory.
+There has been Cambrian explosion of methods in the 2010s, including model-agnostic interpretation methods and many specialized methods for, e.g., deep neural networks.
 Many of these methods have found their way into open source software and products of startups and tech companies.
@@
 Many IML methods are imported from other fields, so thinking outside the box (pun intended) might help in the challenges ahead.
-We believe that the field has to both reach out horizontally -- to other domains -- and vertically -- draw from the rich research in statistics and computer science.
+We believe that the field has to reach out horizontally -- to other domains -- and vertically -- drawing from the rich research in statistics and computer science.
 

### 2020-09-22 2020-ecml-xkdd/paper/paper.tex
Updates from Overleaf

@@
 \usepackage{csquotes}
-% TODO: Strongly short the paper.
-% TODO: Add some images. Add sensitivity analysis vs. component analysis
 \begin{document}
 %
-\title{Interpretable Machine Learning -- A Brief History, State-Of-The-Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
+\title{Interpretable Machine Learning -- A Brief History, State-of-the-Art and Challenges\thanks{This project is funded by the Bavarian State Ministry of Science and the Arts and coordinated by the Bavarian Research Institute for Digital Transformation (bidt) and supported by the German Federal Ministry of Education and Research (BMBF) under Grant No.
 01IS18036A.
@@
   We present a brief history of interpretable machine learning (IML), give an overview of state-of-the-art interpretation methods and discuss challenges when interpreting machine learning models.
-  We focus on the IML methods, putting emphasis on what interpretations can be derived mathematically and algorithmically from a machine learning model.
+  This paper reviews IML from a methodological viewpoint, with an emphasis on what can we derive mathematically and algorithmically from an ML model.
 
@@
 Interpretability is often a deciding factor when a machine learning (ML) model is used in a product, a decision process or in research.
-Interpretable machine learning (IML) methods \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangeably} can be used to discover knowledge, justify the model and its predictions, control and improve models \cite{adadi2018peeking}.
-This paper reviews IML from a methodological viewpoint, with an emphasis on \textbf{what can we derive mathematically and algorithmically from an ML model}.
+Interpretable machine learning (IML) methods \footnote{We will be using Interpretable Machine Learning and Explainable AI exchangeably} can be used to discover knowledge, justify the model and its predictions and to control and improve the models \cite{adadi2018peeking}.
+In this paper, we take a look at the historical building blocks of interpretable machine learning and give an overview of methods to interpret models.
+We argue that interpretable machine learning has reached a state of usability, but some challenges remain.
 
@@
 Rule-based machine learning (e.g., decision rules and trees) has been an active research area since the 1980s \cite{furnkranz2012foundations}.
-Both research on regression analysis and rule-based ML remain important and busy research areas to this day and are even blending together (e.g., model-based trees \cite{zeileis2008model}, RuleFit \cite{friedman2008predictive}).
-Machine learning research in general also started around that time, e.g., support vector machines in 1974 \cite{vapnik1974theory}, early important work on neural networks in the 1960s \cite{schmidhuber2015deep} and boosting in 1990 \cite{schapire1990strength}.
-Interpretability has always been a companion to ML.
-A good example for this is the random forest, which I also see as an IML milestone for making ML practical.
+A lot of machine learning research began in the second half of the 20th century with research on, for example, support vector machines in 1974 \cite{vapnik1974theory}, early important work on neural networks in the 1960s \cite{schmidhuber2015deep} and boosting in 1990 \cite{schapire1990strength}.
+Interpretability has always been a companion of ML.
+The random forest algorithm  \cite{breiman2001random} is an example of the early importance of interpretability and also an IML milestone.
 Random forests, a black box machine learning model that often works well without tuning, came with a built-in feature importance measure, which contributed to its success.
-The evidence that interpretability was success factor are the many citations ($>60000$ citations on Google Scholar as of September 2020) of the original paper \cite{breiman2001random}, but also of papers improving the importance measure (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}).
+\footnote{The random forest paper has been cited  $>60000$ times (Google Scholar; September 2020) and there are many papers improving the importance measure (\cite{strobl2008conditional,strobl2007bias,hapfelmeier2014new,ishwaran2007variable}) which are also cited frequently.}
 In the 2010s came the deep learning hype, after a deep neural network won the ImageNet challenge.
-A few years after that, the IML field really took off (around 2015), judging by frequency of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (Figure \ref{fig:count} right)and papers published with these terms (Figure \ref{fig:count} left).
+A few years after that, the IML field really took off (around 2015), judging by frequency of the search terms "Interpretable Machine Learning" and "Explainable AI" on Google (Figure \ref{fig:count} right) and papers published with these terms (Figure \ref{fig:count} left).
 \begin{figure}
@@
 \end{figure}
-Since then, many approaches have been introduced to the field, many of them model-agnostic explanation methods, i.e., working for different types of ML models, but also model-specific explanation methods to interpret, e.g., deep neural networks and tree ensembles.
-Rule-based ML and statistical regression models continue to be developed and serves both as full ML models, but also as building blocks for many of the newer IML approaches.
-%Could it be that only the terms have changed, but the field has always been there and the same?
-%I would still argue that the field has reached a new form, as evidenced by new keywords "Interpretable Machine Learning" and "Explainable Artificial Intelligence", but also by bringing in new ideas from other fields, pulling many unconnected research fields together and an additional focus on model-agnostic ML interpretation methods.
-% TODO: Find more evaluation papers
-\textbf{Today} IML reached a first state of usability.
-Some methods have been established with a better understanding of weaknesses and strengths.
-There are dozens of surveys of the IML field \cite{molnar2019,guidotti2018survey,vilone2020explainable,rosenfeld2019explainability,adadi2018peeking,anjomshoae2019explainable,du2019techniques,carvalho2019machine}.
-The field is starting to further consolidate terms and knowledge \cite{hall2019systematic,doshi2017towards,murdoch2019definitions,samek2019towards,preece2018stakeholders} and we have some ideas how to evaluate explanations and interpretability methods \cite{mohseni2018multidisciplinary}.
-Many open source software packages implement various techniques \cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml} .
-These software tools give ML practitioners the means to interpret their models.
+Since then, many approaches have been introduced to the field, many of them model-agnostic explanation methods, which work for different types of ML models, but also model-specific explanation methods to interpret,for example, deep neural networks and tree ensembles.
+Regression analysis and rule-based ML remain important and active research areas to this day and are blending together (e.g., model-based trees \cite{zeileis2008model}, RuleFit \cite{friedman2008predictive}).
+Many extensions of the linear regression model exist \cite{hastie1990generalized,fahrmeir2013multivariate,gelman2006data} and linear regression models remain an active field of research \cite{fasiolo2020scalable,caruana2015intelligible,fasiolo2020fast,ustun2016supersparse}
+Rule-based ML also remains an active area of research (for example, \cite{wang2015falling,letham2015interpretable,hothorn2015ctree}).
+Both regression models and rule-based ML serve as stand-alone ML algorithms, but also as building blocks for many other IML approaches.
+
+\section{Today}
+
+IML has reached a first state of usability.
+Research-wise, the field is maturing in terms of methods surveys  \cite{molnar2019,guidotti2018survey,vilone2020explainable,rosenfeld2019explainability,adadi2018peeking,anjomshoae2019explainable,du2019techniques,carvalho2019machine}, further consolidation of terms and knowledge \cite{hall2019systematic,doshi2017towards,murdoch2019definitions,samek2019towards,preece2018stakeholders,chromik2020taxonomy} and work about defining interpretability or evaluation of IML methods \cite{mohseni2018multidisciplinary,mohseni2018human,ribeiro2016should,hoffman2018metrics}.
+We have a better understanding of weaknesses of IML methods \cite{laugel2019dangers,molnar2019,molnar2020pitfalls,hooker2019please,adebayo2018sanity,janzing2019feature,sundararajan2019many,strobl2007bias,strobl2008conditional,apley2016visualizing}.
+Open source software  with implementations of various IML methods is available \cite{iml,biecek2018dalex,pedregosa2011scikit,klaise2020alibi,nori2019interpretml} and gives ML practitioners the means to interpret their models.
 Regulation such as GDPR has spurred a discussion around further needs of interpretability \cite{wachter2017counterfactual}.
@@
 
-In this section we broadly cover some popular IML methods.
 We distinguish IML methods between component and sensitivity analysis\footnote{Not to be confused with the research field of sensitivity analysis, which studies the uncertainty of outputs in mathematical models and systems.
-There are methodological overlaps (e.g., Shapley values), but also differences in methods and how input data distributions are handled}.
-An IML method that works by component analysis assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.\footnote{Some surveys \cite{vilone2020explainable,guidotti2018survey,du2019techniques} divide methods into \textit{ante-hoc (alternatively: transparent design, white box models)} and \textit{post-hoc} IML method, depending on whether interpretability is considered at model design and training or after training, leaving the (black-box) model unchanged.
+There are methodological overlaps (e.g., Shapley values), but also differences in methods and how input data distributions are handled.}, illustrated in Figure~\ref{fig:iml-type}.
+An IML method that works by component analysis assigns meaning to \textit{(learned) model parameters and structures} while sensitivity analysis \textit{probes the model with artificial data points and describes its behavior}.\footnote{Some surveys distinguish between \textit{ante-hoc (alternatively: transparent design, white box models, inherently interpretable model)} and \textit{post-hoc} IML method, depending on whether interpretability is considered at model design and training or after training, leaving the (black-box) model unchanged.
 Another category divides model-agnostic and model-specific.
 Component vs.
-sensitivity analysis is, in our opinion, another interesting view and why only repeat what others already have said?}
-The idea of component analysis is closely related to decomposability:
-You can only analyze components of a model that can be decomposed into individual components that we can interpret individually, like weights in a linear model.
-Component analysis does not necessarily require that the user understands the model in its entirety (simulatability), but only individual parts.
+sensitivity analysis is, in our opinion, another interesting view.}
+
+\begin{figure}
+    \centering
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{specific-black-box.png}
+    \end{minipage}\hfill
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{agnostic-black-box.png}
+    \end{minipage}
+    \label{fig:iml-type}
+    \caption{Some IML approaches work by assigning meaning to individual model components (left), some by analyzing the model predictions for perturbations of the data.}
+\end{figure}
+
+\subsubsection{Analyzing Components of \enquote{Interpretable} Models}
+Component analysis requires that a model can be decomposed into individual components that we can interpret individually, like weights in a linear model, but it does not necessarily require that the user understands the model in its entirety (simulatability).
 Component analysis is always model-specific, e.g., a linear regression model can be interpreted by analyzing its coefficients, while a decision tree can be interpreted by visualizing the learned decision structure.
-Sensitivity analysis is mostly model-agnostic and works by manipulating input data and analyzing the respective model predictions.
-Usually, we analyze the components of ML models that are said to be inherently interpretable, but there are also ways to analyze components of more complex models, but this often requires more work.
-\textit{Inherently interpretable models} are models with learned structures and learned parameters which can be assigned some meaning.
+\textbf{Inherently interpretable models} are models with learned structures and learned parameters which can be assigned some meaning.
 Linear regression models, decision trees and decision rules are considered to be interpretable \cite{freitas2014comprehensible,huysmans2011empirical}.
@@
 The model structure, a weighted sum of features, allows that we interpret the $\beta$-weights as the effect that a feature has on the prediction.
-Many extensions exist \cite{hastie1990generalized,fahrmeir2013multivariate,gelman2006data} and linear regression models remain an active field of research \cite{fasiolo2020scalable,caruana2015intelligible,fasiolo2020fast,ustun2016supersparse}
 Decision trees and other rule-based ML models have a learned structure (e.g.,\enquote{IF feature $x_1 > 0$ and feature $x_2 \in \{A,B\}$, THEN predict 0.6}).
 We can interpret the learned structure to trace how the model makes predictions.
-Rule-based ML also remains an active area of research (for example, \cite{wang2015falling,letham2015interpretable,hothorn2015ctree}).
-
 The more complex interpretable models get (e.g., linear models with hundreds of features and complex interaction terms or deep decision trees), the more difficult the interpretation becomes.
 
-Component analysis is how we interpret inherently interpretable models, but component analysis can be used to explain black box models.
+\subsubsection{Analyzing Components of \enquote{Black-Box} Models}
+
+With a bit more effort, we can also analyze components of black box models.
+\footnote{This blurs the line between \enquote{inherently interpretable} and a \enquote{black-box} model.}
 For example, the abstract features learned by a deep convolutional neural network (CNN) can be visualized by finding or generating images that activate a feature map of the CNN \cite{olah2017feature}.
-% TODO: Add image
-Feature visualization showed that CNNs learn low-level features such as edge-detectors at the early neural layers and gradually more abstract detectors for textures and objects in higher layers.
-As one other example among many, there are various feature importance measures for the random forest that are based on analyzing components.
-For example, the minimal depth distribution \cite{randomForestExplainer,ishwaran2010high} and the Gini importance \cite{breiman2001random} analyze the structure of the trees of the forest.
-.
-Since all IML methods that rely on component analysis are model-specific, the explanations are tied to that ML model.
-If a different ML model is used, the IML method has to change, which is a disadvantage.
-But if a model is well understood and frequently used in community, like random forest in ecology research \cite{cutler2007random}, model component analysis can be a powerful tool.
-Component analysis can explain the overall model behavior (global) or individual predictions (local).
-In many cases the components of a model has the same global and local interpretation.
-For example, in linear regression models the weights explain both how the features affect an individual prediction, but also summarize the global behavior.
+For the random forest, the minimal depth distribution \cite{randomForestExplainer,ishwaran2010high} and the Gini importance \cite{breiman2001random} analyze the structure of the trees of the forest and can be used to quantify feature importance.
+% TODO: Finde more examples to make black-box model structures interpretable.
+Also, e.g., the logistic regression model needs \enquote{post-hoc} analysis bc.
+we have to compute log odds from the weights.
+and for gam we have to visualize the function learned by splines
+If a ML algorithm is well understood and frequently used in a community, like random forests in ecology research \cite{cutler2007random}, model component analysis can be powerful, but it has the disadvantage that it is tied to that specific model.
 
-\paragraph{IML method based on sensitivity analysis} often treat the ML model as a closed system that receives an input and produces an output, the prediction.
-We further distinguish between local and global explanations that sensitivity based IML methods can produce.
+\subsection{Local Sensitivity Analysis}
 
-\paragraph{Local explanations} are used to explain individual predictions of an ML model.
+Sensitivity analysis is mostly model-agnostic and works by manipulating input data and analyzing the respective model predictions.
+IML methods based on sensitivity analysis often treat the ML model as a closed system that receives features as an input and produces the prediction as output.
+In the following, we distinguish between local and global explanations.
+Local methods explain individual predictions of ML models.
 Local explanation methods have received much attention and there has been a lot of innovation in the last years.
@@
 The Shapley values \cite{shapley1953value} provide an answer on how to fairly share a payout among the players of a collaborative game.
-The collaborative game idea can be applied to ML \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}: The predicted value is the payout, each feature value is a player.
+The collaborative game idea can be applied to ML \cite{vstrumbelj2014explaining,lundberg2017unified,lundberg2018consistent}:
+The predicted value is the payout, each feature value is a player.
+% TODO: Write about extensions and evaluations, also cite kumar2020problems. Check MOC paper for references
 Another popular method are local interpretable model-agnostic explanations, short LIME \cite{ribeiro2016should}.
 LIME uses surrogate models, i.e., the method trains an interpretable model such as a linear regression model to approximate the predictions of original ML model.
-Numerous extensions of LIME exist, that try to fix issues with the original method, extend it to other tasks and data or analyze its properties \cite{hu2018locally,rabold2018explaining,rabold2019enriching,visani2020optilime,haunschmid2020audiolime,rahnama2019study,shankaranarayana2019alime}.
+Numerous extensions of LIME exist, that try to fix issues with the original method, extend it to other tasks and data or analyze its properties \cite{hu2018locally,rabold2018explaining,rabold2019enriching,visani2020optilime,haunschmid2020audiolime,rahnama2019study,shankaranarayana2019alime,botari2020melime}.
 % TODO: Add the many extensions of Shapley Values
+% TODO: Add extensions and implementations of Counterfactual explanations
 Some IML methods rely on model-specific knowledge to analyze how changes in the input features changes the output.
@@
 
-\paragraph{Global, model-agnostic explanation methods} are used to explain how the model behaves on average for some given dataset.
+\subsection{Global Sensitivity Analysis}
+
+Global model-agnostic explanation methods are used to explain how the model behaves on average for some given dataset.
 A useful distinction of global explanations are feature importance and feature effect.
 Feature importance ranks features based on how relevant they were for the prediction.
-Permutation Feature Importance \cite{fisher2019all}, which is based on permuting features is a popular importance measure, originally suggested for random forests \cite{breiman2001random}, now available as model-agnostic version.
+Permutation feature importance \cite{fisher2019all,casalicchio2018visualizing} is a popular importance measure, originally suggested for random forests \cite{breiman2001random}.
 An alternative are variance based measures.
@@
 The feature effect expresses how a change in a feature changes the predicted outcome.
-Popular feature effect plots are Partial Dependence Plots \cite{friedman2001greedy}, Individual Conditional Expectation Plots \cite{goldstein2015peeking} and Accumulated Local Effect  (ALE)  Plots \cite{apley2016visualizing}.
+Popular feature effect plots are partial dependence plots \cite{friedman2001greedy}, individual conditional expectation curves \cite{goldstein2015peeking}, accumulated local effect plots \cite{apley2016visualizing} and the functional ANOVA \cite{hooker2007generalized}.
+Analyzing influential data instances, inspired by statistics, provides a different view into the model and describes how influential a data point was for a prediction \cite{koh2017understanding}.
 
-\paragraph{Surrogate models}\footnote{Surrogate models are related to knowledge distillation and the teacher-student model.} are interpretable models designed to \enquote{copy} the behavior of the ML model.
+\subsection{Surrogate Models}
+
+Surrogate models\footnote{Surrogate models are related to knowledge distillation and the teacher-student model.} are interpretable models designed to \enquote{copy} the behavior of the ML model.
 They do not neatly fit into the categorization of component or sensitivity analysis, as they combine both:
 The surrogate approach treats the ML model as a black box, only requiring the input / output data  (similar to sensitivity analysis), but the interpretation is based on analyzing components of the interpretable surrogate model.
-Many methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting,craven1996extracting,krishnan2017palm}.
-They differ in which ML model they can be used for, the data sampling strategy, the interpretable model that is used and so on.
-There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}, which resembles more the component analysis approach.
-
-% TODO: Add that sensitivity analysis can also be used for inherently interpreteable models
+Many IML methods are surrogate model approaches \cite{puri2017magix,molnar2019,ming2018rulematrix,ribeiro2016should,frosst2017distilling,bastani2017interpreting,craven1996extracting,krishnan2017palm} and differ in the targeted ML model, the data sampling strategy, the interpretable model that is used and so on.
+There are also methods for extracting, e.g., decision rules from specific models based on their internal components such as weights \cite{andrews1995survey,augasta2012rule}.
 
@@
 %- Ethics and Philosophy.
-%- Explanations can have different \textit{structure}: text, image, tables, ...
-%- Methods that are model-agnostic might not be input data agnostic, e.g., PDP does not make sense for pixel data as used in image classification
 %- Some methods are designed or adapted so that we can more easily understand what its components do.
@@
 %- TODO: Add more to the idea of making components of a (black-box) ML model more interpretable. This can be e.g. tree pruning, adding sparsity to linear regression models, disentangle neurons.
-%- We can use sensitivity analysis also to interpretable models
 %- We did not touch topics where data is analyzed before modelling to understand the data better
-%- Idea that we can decompose the function with functional ANOVA and so on
-
-%\paragraph{Blending all together.} This list was not exhaustive. For example, there are also ways to add interpretability constraints to a model, such as monotonicity or sparsity.
+% - Talk about social science: miller2017explainable
 %This can helps component analysis, e.g., a linear model with most coefficients at zero is easier to interpret.
 %But this can also help with sensitivity analysis, as PDPs of monotonic features will only show in one direction.
-
+% TODO: Write about dependent features and how they are affecting interpretation. also cite kumar2020problems here
 \section{Challenges}
 
-This is an incomplete list of some of the challenges of IML, mostly based on \cite{molnar2020pitfalls}.
+This section presents an incomplete of challenges for IML, mostly based on \cite{molnar2020pitfalls}.
 
-\paragraph{Statistical Uncertainty and Inference}
+\subsection{Statistical Uncertainty and Inference}
+
 Many IML methods such as permutation feature importance or Shapley values provide explanations without quantifying the uncertainty of that explanation.
-The the model itself, but also the explanations are usually computed from data and are subject to variance.
-We have to become more rigorous in reporting both reporting the explanations, but also their uncertainty.
-Some research is working towards the goal of quantifying uncertainty of explanations, for example for feature importance \cite{watson2019testing,fisher2019all}, layer-wise relevance propagation \cite{fabi2020feature} and Shapley values \cite{williamson2020efficient}.
-Quantifying uncertainty and expressing it as error bars, confidence intervals or also test statistics and p-values is an important part of statistical modeling.
-The goal of statistical inference is to infer properties of the underlying data distribution of the phenomenon that is being studied.
-In this process, a statistical model is proposed with a set of assumptions (for example, a linear model with assumptions that the features only have additive effects).
-When all assumptions are met, the statistical model may be interpreted to reflect the true data generating process.
-Statistical modeling is used for testing scientific hypothesis.
-Machine learning modeling takes a different approach:
-Instead of assuming a model of the data generating process, various flexible models are compared on test data.
-It is desirable to draw also conclusions from these models about the underlying data generating process.
-With IML, with have method to compute meaningful properties of the model. 
-But to infer properties about the data generating process from the ML model, we have to specify the assumptions needed that allow a combination of ML model and IML method to be interpreted as the \enquote{true} model.
+The model itself, but also the explanations are computed from data and are subject to uncertainty.
+We have to become more rigorous in reporting explanations including their uncertainty.
+First research is working towards quantifying uncertainty of explanations, for example, for feature importance \cite{watson2019testing,fisher2019all}, layer-wise relevance propagation \cite{fabi2020feature} and Shapley values \cite{williamson2020efficient}.
+
+With IML, with have methods to compute meaningful properties of an ML model. 
+But to infer properties about the data generating process from the ML model, we have to specify the required assumptions that would allow us to do so.
 We also need tools and procedures to check these assumptions.
@@
 
-\paragraph{Causal interpretation.}
+\subsection{Causal Interpretation}
 Ideally, a model relies on the true causes to make a prediction.
 Then we could  interpret it causally, making statements about the underlying phenomena, which is the focus of ML in research, but also makes models robust against adversarial attacks, and more useful when used as a basis for decision making.
+% TODO: Find citation for this statement
 Unfortunately, predictive performance and causality can be conflicting goals.
 For example, today's weather directly causes tomorrow's weather, but we might only have access to feature \enquote{wet ground}.
-Using \enquote{wet ground} in the prediction model for \enquote{tomorrow's weather} is useful as it has information about \enquote{today's weather}, but we are not allowed to interpret it causally, because the confounder \enquote{today's weather} is missing.
+Using \enquote{wet ground} in the prediction model for \enquote{tomorrow's weather} is useful as it has information about \enquote{today's weather}, but we are not allowed to interpret it causally, because the confounder \enquote{today's weather} is missing from the ML model.
 Further research is needed to understand when we are allowed to make causal interpretations of an ML model.
@@
 
-\paragraph{Lack of definition} is a common critique of the field \cite{lipton2018mythos,doshi2017towards}.
+\subsection{Feature Dependence}
+
+Feature dependence introduces problems with attribution and extrapolation.
+% Attribution problem
+Attribution of importance and effects to features becomes difficult when features are, for example, correlated and therefore share information.
+It can be shown that correlated features in random forests are preferred and attributed a higher importance \cite{strobl2008conditional,hooker2019please}.
+% Extrapolation problem
+Many sensitivity analysis based methods permute features.
+When the permuted feature has some dependence with another, this association is broken and the resulting data point extrapolates to areas outside the distribution.
+The ML model was never trained on such combinations and will likely not be confronted with similar data points for realistic prediction.
+Therefore extrapolation can cause misleading interpretations.
+% Fixes
+There have been attempts to \enquote{fix} permutation-based method, by using a conditional permutation scheme that respects the joint distribution of the data.
+The change from unconditional to conditional permutation changes the respective interpretation methods \cite{molnar2020model,apley2016visualizing} or in worst case can break it \cite{janzing2019feature,sundararajan2019many}.
+
+\subsection{Definition of Interpretability}
+A lack of definition is a common critique of the field \cite{lipton2018mythos,doshi2017towards}.
 Without a definition of "interpretability", how can we decide if a new method explains ML models better?
 To evaluate the \textbf{predictive performance} of an ML model, we simply compute the prediction error on test data given the groundtruth label.
-To evaluate the \textbf{interpretability} of that same ML model is not as easy.
+To evaluate the \textbf{interpretability} of that same ML model is more difficult.
 We do not know what the groundtruth explanation looks like and have no straightforward mean to quantify how interpretable a model is or how correct an explanation is.
@@
 
-Two main ways of evaluation of interpretability have emerged: \textit{objective evaluations} which are mathematically quantifiable metrics, and {human-centered evaluations} which involve studies with users.
-It seems like we are on our way to have many definitions of \textit{various aspects} of interpretability such as sparsity, interaction strength, fidelity (how well explanation approximates the ML model), sensitivity to perturbations,  and a user's ability to run a model on a given input (simulatability) are examples of such aspects.
+The two main ways of evaluation of interpretability are: \textbf{objective evaluations} which are mathematically quantifiable metrics, and {human-centered evaluations} which involve studies with users.
+Examples of aspects of interpretability are sparsity, interaction strength, fidelity (how well explanation approximates the ML model), sensitivity to perturbations, and a user's ability to run a model on a given input (simulatability).
+% TODO: check with IML book
 The challenge ahead remains to establish a best practice on how to evaluate interpretation methods and the explanations they produce.
@@
 
+\subsection{More Challenges Ahead}
+% TODO: Rework section
 We have focused mostly on the methodological, mathematical challenges in a rather static setting, where you are given a machine learning model and data.
-\textit{Many more} challenges lie ahead to improve interpretability of machine learning.
+But ML models are not used in a static and isolated way.
+Rather they are embedded in a process or product.
 What is the best way to explain a machine learning model for people with a specific background and how do we present the explanations?
-How are people affected by explanations, how are explanations maybe misused (CITE fairwashing by HIma)?
+How are people affected by explanations, how are explanations maybe misused?
 This is also a rather static view point in the sense that we can do a lot if we allow a more interactive way of modeling and "having a conversation" between a person and the model or even the modeling process.
@@
 
-Interpretable Machine Learning is a young field, but has deeper roots in Statistics and Computer Science and also draws from other fields such as the social sciences, human-computer interaction and game theory.
-There has been Cambrian explosion of methods in the 2010s, including model-agnostic interpretation methods and many specialized methods for, e.g., deep neural networks.
-Many of these methods have found their way into open source software and products of startups and tech companies.
-Interpretable Machine Learning is used in science, company products and processes.
-While we have reached a first maturity, there a still some challenges ahead, such as feature dependence, causality and inference.
-Many IML methods are imported from other fields, so thinking outside the box (pun intended) might help in the challenges ahead.
-We believe that the field has to reach out horizontally -- to other domains -- and vertically -- drawing from the rich research in statistics and computer science.
+Interpretable machine learning is a young field, but has deeper roots in statistics and computer science and also draws from other fields such as the social sciences, human-computer interaction and game theory.
+There has been a Cambrian explosion of methods in the 2010s, including model-agnostic interpretation methods and many specialized methods for, e.g., deep neural networks.
+Many IML methods have found their way into open source software and products of startups and tech companies.
+Interpretable machine learning is used in science, company products and processes.
+While we have reached a first state of usability and productivity, there a still some challenges ahead, such as feature dependence, causality and uncertainty quantification.
+To solve the challenges ahead, we believe that the field has to reach out horizontally -- to other domains -- and vertically -- drawing from the rich research in statistics and computer science.
 

### 2021-06-21 2021-bliz-summer-school/2021-talk-bliz.Rmd
add bliz talk

---
title: "Interpretable Machine Learning"
author: "Christoph Molnar, Julia Herbinger & Giuseppe Casalicchio"
date: "June 5th, 2021"
output: 
  ioslides_presentation:
    widescreen: false
    smaller: false
css: ../styles.css
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
  warning = FALSE,
  message = FALSE, 
  fig.align='center', 
  out.width = '90%',
  dev = "CairoPNG")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 15))
```

```{r, child="../chunks/story-wines.Rmd"}
```

# Why interpretability?

## {.center data-background=../images/explain.png data-background-size=contain}

```{r, child="../chunks/tools.Rmd"}
```

```{r child = "../chunks/book.Rmd"}
```

# Backup

## Units in Wine dataset

- fixed acidity g(tartaric acid)/dm^3^
- volatile acidity: g(acetric acid/dm^3^)
- citric acid: g/dm^3^
- residual sugar: g/dm^3^
- chlorides: g(sodium chloride)/dm^3^
- free sulfur dioxide: mg/dm^3^
- total sulfur dioxide: mg/dm^3^
- density> g/cm^3^
- pH
- sulphates: g(postassium sulphate) / dm^3^
- alcohol vol.%

### 2021-06-21 chunks/story-wines.Rmd
add bliz talk

@@
 
-## Accumulated Local Effects
+## ICE and PDP
 
-```{r,  out.width = '78%'}
-knitr::include_graphics("../images/ale-intuition-1.jpg")
-```
-
-## Accumulated Local Effects
-
-```{r,  out.width = '78%'}
-knitr::include_graphics("../images/ale-intuition-2.jpg")
-```
-
-## Accumulated Local Effects
-
-```{r,  out.width = '78%'}
-knitr::include_graphics("../images/ale-intuition-3.jpg")
-```
-
-## Accumulated Local Effects
-```{r,  out.width = '78%'}
-knitr::include_graphics("../images/ale-intuition-4.jpg")
-```
-
-## Accumulated Local Effects
-```{r,  out.width = '78%'}
-knitr::include_graphics("../images/ale-intuition-5.jpg")
+```{r,  out.width = '72%'}
+knitr::include_graphics("../images/pdp.gif")
 ```
@@
 ```{r}
-eff = FeatureEffect$new(pred, "alcohol", method = 'ale')
+eff = FeatureEffect$new(pred, "alcohol", method = 'pdp+ice')
 plot(eff)
@@
 ```{r}
-eff = FeatureEffect$new(pred, "volatile.acidity", method = 'ale')
+eff = FeatureEffect$new(pred, "volatile.acidity", method = 'pdp+ice')
 plot(eff)
@@
 ```{r}
-effs = FeatureEffects$new(pred)
+effs = FeatureEffects$new(pred, method = "pdp")
 plot(effs, ncols = 4)

### 2021-06-21 code/prepare-wine-data.R

@@ -4,4 +4,5 @@ wine = na.omit(wine)
 # Not sure if there is some order to it
 set.seed(42)
 wine = wine[sample(1:nrow(wine)),]
+wine$type = as.factor(wine$type)
 n = nrow(wine)

### 2021-06-30 chunks/story-wines.Rmd
finalizes talk

@@
 ```{r load-data,include=FALSE}
+library('patchwork')
 library('mlr')
@@
 
+## Linear Regression {.center data-background=../images/lm.png data-background-size=contain}
+
+## Decision Tree {.center data-background=../images/tree.png data-background-size=contain}
+
 ## Random Forest
@@
 
+## Comparing Model Performance
+
+Model performance was measured as mean absolute error using 5-fold cross-validation.
+
 ```{r benchmark}
@@
 lrns = list(lrn.ranger, lrn.lm, lrn.rpart)
+test_size = 300
+test_index = sample(1:nrow(wine), size = test_size)
+wine_test = wine[test_index,]
+wine_train = wine[setdiff(1:nrow(wine), test_index), ]
 
-tsk = makeRegrTask(data = wine, target = "quality")
+tsk = makeRegrTask(data = wine_train, target = "quality")
 
@@
 bmr_tab = getBMRAggrPerformances(bmr, as.df = TRUE)
-#knitr::kable(bmr_tab[-1])
+bmr_tab$mae.test.mean = round(bmr_tab$mae.test.mean, 2)
+bmr_tab$learner.id[bmr_tab$learner.id == "regr.ranger"] =  "Random Forest"
+bmr_tab$learner.id[bmr_tab$learner.id == "regr.lm"] =  "Linear Regression"
+bmr_tab$learner.id[bmr_tab$learner.id == "regr.rpart"] =  "Decision Tree"
+colnames(bmr_tab) = c("", "Model", "Mean Absolute Error")
+knitr::kable(bmr_tab[-1])
 ```
@@
 set.seed(42)
-sample_size = 500
-wine_subset = wine[sample(1:nrow(wine), size = sample_size),]
-
-pred = Predictor$new(mod, data = wine_subset, y = "quality")
+pred = Predictor$new(mod, data = wine_test, y = "quality")
 ```
-
-## Train Random Forest to Predict Quality {.emphasizedabit}
-
-Mean absolute error on test data (cross-validated): `r sprintf('%.2f', bmr_tab$mae.test.mean[bmr_tab$learner.id == "regr.ranger"])`
 
@@
 ```{r check-model}
-preds = pred$predict(wine)
-preds$actual = wine$quality
+preds = pred$predict(wine_test)
+preds$actual = wine_test$quality
 ggplot(preds, aes(x = actual, y = .prediction, group = actual)) + 
@@
 
-## Which features are important?
+# Which features are important?
 
@@
 
-## Which features are important?
+1. Measure model error on test data
+1. Shuffle feature of interest
+1. Measure model error again
+1. Compute difference or ratio
+
+## Permutation Feature Importance
 
 ```{r feature-importance}
-imp = FeatureImp$new(pred, loss = "mae")
+imp = FeatureImp$new(pred, loss = "mae", compare = "difference")
 plot(imp)
@@
 
-## How do features affect predictions?
+# How do features affect predictions?
+
+## ICE and PDP
+
+ICE: Individual Conditional Expectation curves
+
+PDP: Partial Dependence Plots
+
+1. Choose a feature
+1. Define a grid of values along that feature
+1. For each value of the grid
+    1. Replace value for all data points the feature value with the grid value
+    1. Get the model predictions
+1. Draw a line through the grid predictions -> ICE
+1. Average the lines -> PDP
 
@@
 
-## Rule of thumb for wine quality?
+## Interactions
+
+```{r, out.width = "100%"}
+library("patchwork")
+p1 = FeatureEffect$new(pred, method = "pdp", feature = "alcohol")$plot()
+p2 = FeatureEffect$new(pred, method = "pdp", feature = "volatile.acidity")$plot()
+p3 = FeatureEffect$new(pred, method = "pdp", feature = c("alcohol", "volatile.acidity"))$plot()
+(p1 + p2) / p3 + plot_layout(heights = c(1,3))
+```
+
+## Interactions (H-Statistic)
+
+H-Statistic tells us how much of the effect of two is due to interactions.
+
+```{r}
+ntrct = Interaction$new(pred, feature = "alcohol")
+plot(ntrct, ncols = 4)
+```
+
+# Rule of thumb for wine quality?
 
@@
 
-## Explain individual predictions
+# Explain individual predictions
 
@@
 
-## Improve worst wine?
+# How Can the Worst Wine be Improved?
 

### 2021-07-05 chunks/story-wines.Rmd
adds some formulas

@@
 
+$$MAE(y,f(X)) = abs(y - f(X))$$
+
 ```{r final-model}
@@
 
-1. Measure model error on test data
-1. Shuffle feature of interest
+1. Measure model error on test data (e.g. MAE)
+1. Permute feature of interest
 1. Measure model error again
@@
 
-```{r,  out.width = '72%'}
+```{r,  out.width = '50%'}
 knitr::include_graphics("../images/pdp.gif")
 ```
+
+$$PDP(x) = \frac{1}{n} \sum_{i=1}^n f(x, x_C^{(i)})$$
 
@@
 
-H-Statistic tells us how much of the effect of two is due to interactions.
+H-Statistic tells us how much of the effect of two is due to interactions: $H^2_{jk}= \frac{\sum_{i=1}^n\left[PD_{jk}(x^{(i)})-PD_j(x^{(i)})-PD_k(x^{(i)})\right]^2}{\sum_{i=1}^n{PD}^2_{jk}(x^{(i)})}$
 
-```{r}
+```{r, out.width = "80%"}
 ntrct = Interaction$new(pred, feature = "alcohol")

### 2021-07-21 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
start talk

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output: 
  beamer_presentation:
    includes:
      in_header: "header_pagenrs.tex"
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
  warning = FALSE,
  message = FALSE, 
  fig.align='center', 
  out.width = '90%',
  dev = "CairoPNG")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 15))
```

# Some ideas

- Cite pitfalls at many places in the talk
- INtroduction with History: INterweave motivation, and also cite history paper
-

# Outline

- A Brief History of Interpretability
- A Closer Look at Model-agnostic interpretability
- methods:
  - Feature Effects: PDP, ICE and ALE
  - Feature Importance: PFI
- Overview of Stuff I did in my PhD 
- Measuring Model Complexity
- Conditional Feature Importance Using Subgroups
- Inference of Effects

# Summary of Thesis

In my thesis I did a lot of consolidation work:

- Book
- R package
- History paper
- Pitfalls paper

What if we embrace black box view of models?

Special focus on old, established tools, and make them fit for

### 2021-07-29 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
structure talk

@@
 
-- A Brief History of Interpretability
+- A Brief History of Interpretable Modeling
 - A Closer Look at Model-agnostic interpretability
@@
 
+# Before You Pin Me Down
+
+To get this out of the way, here is how I use "Interpretable Machine Learning":
+
+It's an umbrella term for approaches that analyze or describe prediction mechanisms in models that were learned on data.
+
+# A Brief History of Interpretable Modeling
+
+Interpretable Models with data have a longer history, but the field of interpretable machine learning is still very young as consolidated field.
+
+* 1800s: Linear regression models
+* 1960: Rule-based ML (AID, CHAID, CART, association rules, ...)
+* 2001: Random Forest
+* 2021: Deep Learning Boom kicked off
+* 2016: Interpretable ML / Explainable AI boom
+
+While 200 years of history, consolidation as larger field only started in 2016.
+
+TODO: Add citation report and or google trends image
+
+# Why Interpretability?
+
+* The more non-linear and complex a model, the more difficult to understand how it operates
+* Trust
+* Debugging
+* Control 
+* ...
+
+# Interpretable Models
+
+* GAM and other linear models
+* Trees and decision rules
+* More complex models, but interpretation requires more work or post-processing
+
+# Model-agnostic Interpretation
+
+Basic idea: Treat the model as a black box, not allowed to look inside.
+
+Idea very powerful: Allows modularity, works for most models, results are comparable across models
+
+TODO: CITE SIPA paper here
+
+# PDP and ICE
+
+# PFI
+
+TODO: Explain how PFI works
+
+# PFI and PDP similarity
+
+CITE paper GC here
+
+# Difference between model-agnostic and model-specific interpetation
+
+* Analyzing components vs. analyzing behavior
+* Components not very powerful, unless they directly describe behavior
+* For complex model, components do not directly tell us about behavior of model
+* Behavior more comparable
+
 # Summary of Thesis
+
+TODO: Draw image with Ipad
 
@@
 
-Special focus on old, established tools, and make them fit for
+Special focus on old, established tools, and make them fit for right use in practice
+
+Also some more specific papers
+
+* subgroups
+* inference
+* quantifying
+
+and some contributions to
+
+* MOC
+* relative feature importance
+* SIPA
+* feature importance and PDP (GC)
+
+# Paper: Subgroups
+
+## Problem: Dependent features
+
+Show example with dependent features for PFI
+
+## Soluation: Conditional variants
+
+Show solution
+
+* But: Not meaningful for PDP, or other measure more useful (LOESS)
+* But: Conditioning unclear
+* But: Intepretation difficult
+
+## Our solution: Permute in subgroups
+
+TODO: Show algorithm
+
+## Some results from our paper
+
+TODO: Show results or application
+
+# Paper: IML Inference
+
+## Problem: Uncertainty quantification
+
+PDP and PFI and most other just descriptive tools
+Often practictioner draw confidence intervals
+but only based on monte carlo approximation
+
+* But what is even the ground truth?
+* And what about the uncertainty in the model fit?
+* How could we estimate the uncertainty?
+
+## Solution: Treat as statistical estimator
+
+## How to Compute the variance
+
+## Problem of too low variance estimates
+
+# Paper: Quantifying
+
+TODO: Copy as much as possible from the talk that already exists for this paper
+
+Motivation: want to quantify in a model-agnostic way how complex a model is
+
+## Functional Decomposition
+
+Explain how functional decomposition works
+
+## ALE
+
+Formula:
+
+TODO: For estimation show the images
+
+## The three measures
+
+* Mean effect complexity
+* non linearity
+* number of features
+
+## Using in benchmarking (multi-objective)
+
+## Thank You for Your Attention

### 2021-08-16 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
continue draft

@@
 institute: "BIPS bremen"
-output: 
-  beamer_presentation:
-    includes:
-      in_header: "header_pagenrs.tex"
+output: ioslides_presentation
 ---
@@
 
-# Some ideas
+## Some ideas
 
@@
 - INtroduction with History: INterweave motivation, and also cite history paper
--
 
-# Outline
+## Outline
 
 - A Brief History of Interpretable Modeling
-- A Closer Look at Model-agnostic interpretability
-- methods:
-  - Feature Effects: PDP, ICE and ALE
-  - Feature Importance: PFI
-- Overview of Stuff I did in my PhD 
+- Model-agnostic interpretability
+- Methods PDP and PFI
+- What I did in my PhD 
+    - Conditional Feature for PFI and PDP Importance Using Subgroups
+    - Statistical Inference for PDP and PFI
     - Measuring Model Complexity
-- Conditional Feature Importance Using Subgroups
-- Inference of Effects
 
-# Before You Pin Me Down
+## Before You Pin Me Down
 
@@
 
-# A Brief History of Interpretable Modeling
-
-Interpretable Models with data have a longer history, but the field of interpretable machine learning is still very young as consolidated field.
+## A Brief History of Interpretable Modeling
 
 * 1800s: Linear regression models
-* 1960: Rule-based ML (AID, CHAID, CART, association rules, ...)
+* 1960s: Rule-based ML
 * 2001:  Random Forest
-* 2021: Deep Learning Boom kicked off
+* 2012:  Deep Learning Boom kicked off
 * 2016:  Interpretable ML / Explainable AI boom
 
-While 200 years of history, consolidation as larger field only started in 2016.
+Interpretable ML has 200 year old roots, but consolidation only started in 2016.
 
@@
 
-# Why Interpretability?
+## Why Interpretability?
 
@@
 
-# Interpretable Models
+## Interpretable Models
 
@@
 * More complex models, but interpretation requires more work or post-processing
+
+## Interpretable Models
+
+```{r, echo = FALSE}
+knitr::include_graphics("../images/white-box.png")
+```
 
@@
 
-# PDP and ICE
+## Interpreting Models
 
-# PFI
+```{r, echo=FALSE}
+knitr::include_graphics("../images/black-box.png")
+```
 
-TODO: Explain how PFI works
+## Model-agnostic Methods
 
-# PFI and PDP similarity
+```{r, echo=FALSE}
+knitr::include_graphics("../images/agnostic-black-box.png")
+```
+
+# Methodological Background
+
+## Individual Conditional Expectations (ICE)
+
+maybe use ice + pdp animation, but split into individual frames so they are clickable
+
+## Partial Dependence Plot (PDP)
+
+The theoretical construct:
+$$ PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})] $$
+
+The estimator:
+$$\widehat{PD} = \frac{1}{n_2}\sum_{i=1}^{n_2} \hat{f}(x, x_{C}^{(i)})$$
+
+What it looks like:
+
+TODO: 1D-plot and 2D-plot
+
+## Permutation Feature Importance PFI
+
+The construct:
+$$PFI =  \mathbb{E}_{Y \tilde{X}_S X_C}(L(Y, \hat{f}(\tilde{X}_S, X_C))) - \mathbb{E}_{Y X_S X_C}(L(Y,\hat{f}(X_S, X_C)))$$
+
+The algorithm (simples form): CITE Breiman
+
+- Measure original loss $L$ on test data.
+- Permute features $X_S$ to get sample $\tilde{X}_S$.
+- Get prediction with $\tilde{X}_S$, measure loss again.
+- Measure difference between losses.
+- Interpretation: PFI is performance drop when "destroying" information about feature.
+
+## PFI and PDP similarity
 
@@
 
-## Soluation: Conditional variants
+## Solution: Conditional variants
 
@@
 
-## Thank You for Your Attention
+# Inference of Feature Effects and Importance in Machine Learning
+
+## Status-Quo
+
+- For inference, practitioner rely on interpretable models
+- remember: interpretable models have parameters that map to data-generating process
+- but: situations where more flexible models match DGP better (measured in test loss)
+- we have: model-agnostic interpretation
+- But: only describes the model behavior
+- Question: Can we link PD and PFI to the DGP? and even do inference?
+
+## Example
+
+TODO: Show PDP and PFI from some papers?
+
+## Linking external tools to DGP
+
+- Hope: Lift PDP and PFI to status of statistical estimator
+- Needed: estimand. A definition of what it is supposed to estimate in the DGP (not the model)
+- Idea: PDP and PFI are expectation of a function
+- Assumption: True function $f$ exists in DGP
+- Then: We can apply same interpretation tools to DGP. Very simple.
+
+TODO: simple image
+
+## Source of uncertainty
+
+- The ground truth estimand allows use to do the usual statistical estimator stuff:
+- For example: Measure MSE
+- Or: split up MSE into bias (systematic deviation of estimator to estimand)
+- Variance: Two sources in the estimation process:
+  - Model is a random variable bc. training data is a sample + model-specific inherent randomness (such as subset sampling in random forests)
+  - Expectations are based on Monte carlo sampling
+
+## Model-PD and Model-PFI
+
+Show definition plus confidence intervals plus image
+
+## Learner-PD and Learner-PFI
+
+Show definition plus confidence intervals plus image
+
+But there is a problem when estimating this variance
+
+## Variance correction
+
+- Variance is underestimated as training data is shared by models.
+- Same for test data.
+- Underestimating variance also means confidence intervals are too short, appearing more certain than they really are.
+- Same problem exists for variance estimates for model performance, as needed for comparing models
+- Nadeau et. al (CITE) proposed correction term
+- term very naive, but MUCH better than without correction
+
+## Show simulation results
+
+# Thank You for Your Attention

### 2021-08-19 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
add ice and pdp explanations

@@
 ```
-
-## Some ideas
-
-- Cite pitfalls at many places in the talk
-- INtroduction with History: INterweave motivation, and also cite history paper
 
@@
 
+## Model-agnostic Interpretation
+
 Basic idea: Treat the model as a black box, not allowed to look inside.
@@
 
-## Individual Conditional Expectations (ICE)
+## Individual Conditional Expectation (ICE)
 
-maybe use ice + pdp animation, but split into individual frames so they are clickable
+How does a feature influence the prediction for an individual observation?
+
+```{r}
+load(url("https://github.com/christophM/interpretable-ml-book/blob/master/data/bike.RData?raw=true"))
+library("mlr")
+library("iml")
+library("ggplot2")
+library(dplyr)
+library(tidyr)
+
+bike.task = makeRegrTask(data = bike, target = "cnt")
+mod.bike = mlr::train(mlr::makeLearner(cl = 'regr.randomForest', id = 'bike-rf'), bike.task)
+
+set.seed(42)
+bike.sub = bike[sample(1:nrow(bike), size = 15),]
+
+pred.bike = Predictor$new(mod.bike, data = bike.sub)
+pdp = FeatureEffect$new(pred.bike, "temp", method = "ice", grid.size = 7)
+
+res.raw = pdp$results
+theme_set(theme_classic(base_size = 20))
+
+## Step 1: Define grid over feature
+grid.dat = data.frame(temp = unique(res.raw$temp), .id = 1, temp2 = 1)
+p = ggplot(grid.dat) + geom_vline(aes(xintercept = temp)) +
+  scale_y_continuous("Prediction", limits = c(min(res.raw$.y.hat), max(res.raw$.y.hat))) +
+  ggtitle("1) Define grid along feature") +
+  scale_x_continuous("Feature")
+  #scale_y_continuous("Prediction")
+p
+```
+
+## Individual Conditional Expectation (ICE)
+
+How does a feature influence the prediction for an individual observation?
+
+```{r}
+## Step 2: Get predictions for data set when at grid
+grid.dat = data.frame(temp = unique(res.raw$temp))
+res.raw$tt = res.raw$temp
+p = ggplot(res.raw, aes(x = temp, y = .value)) +
+  geom_vline(data = grid.dat, aes(xintercept = temp)) +
+  geom_point(aes(color = factor(.id)), size = 3)+
+  scale_color_discrete(guide = "none") +
+  ggtitle("2) Model predictions at grid points") +
+  scale_x_continuous("Feature") +
+  scale_y_continuous("Prediction", breaks = NULL)
+p
+```
+
+## Individual Conditional Expectation (ICE)
+
+How does a feature influence the prediction for an individual observation?
+
+```{r}
+res.raw$.id2 = factor(res.raw$.id)
+p = ggplot(res.raw, aes(x = temp, y = .value))  +
+  #geom_vline(aes(xintercept = temp), data = grid.dat) +
+  geom_line(aes(group = .id), size  = 1)  +
+  geom_point(data = res.raw[c("temp", ".value", ".id2")], aes(color = .id2), size = 3) +
+  scale_color_discrete(guide = "none") +
+  ggtitle("3) Line per data instance -> ICE curves") +
+  scale_x_continuous("Feature") +
+  scale_y_continuous("Prediction", breaks = NULL)
+p
+```
+
+## Partial Dependence Plot (PDP)
+
+How does a feature influence the prediction for an individual observation **on average**?
+
+## Partial Dependence Plot (PDP)
+
+How does a feature influence the prediction for an individual observation **on average**?
+
+```{r}
+p
+```
+
+## Partial Dependence Plot (PDP)
+
+How does a feature influence the prediction for an individual observation **on average**?
+
+```{r}
+res = res.raw %>% group_by(temp) %>%
+  summarize(pdp = mean(.value))
+
+p = ggplot(res) +
+  geom_line(aes(x = temp, y = pdp), size = 1.5) +
+  geom_point(data = res.raw[c("temp", ".value", ".id2")], aes(color = .id2, x = temp, y = .value), size = 1, alpha = 0.5) +
+  ggtitle("4) Average curves to get a PDP") +
+  scale_color_discrete(guide = "none") +
+  scale_x_continuous("Feature") +
+  scale_y_continuous("Prediction", breaks = NULL)
+p
+```
 
@@
 The estimator:
-$$\widehat{PD} = \frac{1}{n_2}\sum_{i=1}^{n_2} \hat{f}(x, x_{C}^{(i)})$$
-
-What it looks like:
-
-TODO: 1D-plot and 2D-plot
+$$\widehat{PD} = \frac{1}{n}\sum_{i=1}^{n} \hat{f}(x, x_{C}^{(i)})$$
 

### 2021-08-21 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
continues draft

@@
 
-- A Brief History of Interpretable Modeling
-- Model-agnostic interpretability
-- Methods PDP and PFI
+- Introduction
+  - Brief History of Interpretability
+  - What Is Model-Agnostic Interpretability?
+  - Partial Dependence Plot (PDP)
+  - Permutation Feature Importance (PFI)
 - What I did in my PhD
-    - Conditional Feature for PFI and PDP Importance Using Subgroups
 - Statistical Inference for PDP and PFI
-    - Measuring Model Complexity
 
-## Before You Pin Me Down
-
-To get this out of the way, here is how I use "Interpretable Machine Learning":
-
-It's an umbrella term for approaches that analyze or describe prediction mechanisms in models that were learned on data.
+# Introduction
 
@@
 
-* 1800s: Linear regression models
-* 1960s: Rule-based ML
-* 2001:  Random Forest
-* 2012:  Deep Learning Boom kicked off
-* 2016:  Interpretable ML / Explainable AI boom
+Interpretable ML has 200 year old roots, but consolidation only started in 2016.
+
+>- 1800s: Linear regression models
+>- 1960s: Rule-based ML
+>- 2001:  Random Forest
+>- 2012:  Deep Learning Boom kicked off
+>- 2016:  Interpretable ML / Explainable AI boom
+
+## A Brief History of Interpretable Modeling
 
@@
 
-TODO: Add citation report and or google trends image
+TODO: Web Of Science for keywords XAI and IML (written out)
 
@@
 
-* The more non-linear and complex a model, the more difficult to understand how it operates
-* Trust
-* Debugging
-* Control
-* ...
+```{r, out.width = 500}
+knitr::include_graphics("../images/explain.png")
+```
 
@@
 
-* GAM and other linear models
-* Trees and decision rules
-* More complex models, but interpretation requires more work or post-processing
+* Decision trees and decision rules
+* Generalized additive models and the whole gang of statistical modeling
+* Interpretable models allow interpretation of internal structure and parameters.
 
@@
 
+## Black Box Models
+
+```{r, echo=FALSE}
+knitr::include_graphics("../images/black-box.png")
+```
+
+Random forest, gradient-boosted trees, neural networks, ...
+
 ## Model-agnostic Interpretation
@@
 
-## Interpreting Models
-
-```{r, echo=FALSE}
-knitr::include_graphics("../images/black-box.png")
-```
-
 ## Model-agnostic Methods
@@
 # Methodological Background
+
+## Individual Conditional Expectation (ICE)
+
+How does a feature influence the prediction for an individual observation?
 
@@
 
-## Permutation Feature Importance PFI
+## Permutation Feature Importance (PFI)
+
+How important is a feature for making correct predictions?
+
+## Permutation Feature Importance (PFI)
+
+How important is a feature for making correct predictions?
+
+Idea: Shuffle the feature, and see how much worse the predictions are.
+
+```{r}
+knitr::include_graphics("../images/permimp-shuffle.png")
+```
+
+## Permutation Feature Importance (PFI)
 
 The construct:
+
 $$PFI =  \mathbb{E}_{Y \tilde{X}_S X_C}(L(Y, \hat{f}(\tilde{X}_S, X_C))) - \mathbb{E}_{Y X_S X_C}(L(Y,\hat{f}(X_S, X_C)))$$
@@
 
-## PFI and PDP similarity
-
-CITE paper GC here
-
 # Difference between model-agnostic and model-specific interpetation
@@
 
-# Summary of Thesis
+# Stuff I Did for My Thesis
 
-TODO: Draw image with Ipad
+## Most Important
 
-In my thesis I did a lot of consolidation work:
+## Not Part of My PhD Though
 
+```{r, out.width = 540}
+knitr::include_graphics("../images/academic-credit.jpg")
+```
+
+## Summary of Thesis
+
+- Consolidation Work
   - Book
@@
   - Pitfalls paper
-
-What if we embrace black box view of models?
-
-Special focus on old, established tools, and make them fit for right use in practice
-
-Also some more specific papers
-
-* subgroups
-* inference
-* quantifying
-
-and some contributions to
-
-* MOC
-* relative feature importance
-* SIPA
-* feature importance and PDP (GC)
+- Improving Methods
+  - subgroups
+  - iml inference
+  - quantifying
+- Contributed to
+  - MOC
+  - relative feature importance
+  - SIPA
+  -  feature importance and PDP (GC)
 

### 2021-08-21 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
use xaringan

@@
 institute: "BIPS bremen"
-output: ioslides_presentation
+output:
+  xaringan::moon_reader
 ---
@@
 
-## Outline
+# Outline
 
@@
 
+---
+
 # Introduction
+
+---
 
@@
 Interpretable ML has 200 year old roots, but consolidation only started in 2016.
+--
 
->- 1800s: Linear regression models
->- 1960s: Rule-based ML
->- 2001:  Random Forest
->- 2012:  Deep Learning Boom kicked off
->- 2016:  Interpretable ML / Explainable AI boom
+- 1800s: Linear regression models
+
+--
+
+- 1960s: Rule-based ML
+
+--
+
+- 2001:  Random Forest
+
+--
+- 2012:  Deep Learning Boom kicked off
+
+--
+
+- 2016:  Interpretable ML / Explainable AI boom
+
+???
+
+Now impress audience with history
+
+---
 
@@
 
+---
+
 ## Why Interpretability?
@@
 ```
+
+---
 
@@
 
+---
+
 ## Interpretable Models
@@
 
+---
+
 # Model-agnostic Interpretation
+
+---
 
@@
 
+---
+
 ## Model-agnostic Interpretation
@@
 
+---
+
 ## Model-agnostic Methods
@@
 
+---
+
 # Methodological Background
+
+---
 
@@
 
-## Individual Conditional Expectation (ICE)
+--
 
-How does a feature influence the prediction for an individual observation?
-
-```{r}
+```{r, out.width = 500}
 load(url("https://github.com/christophM/interpretable-ml-book/blob/master/data/bike.RData?raw=true"))
@@
 ```
+---
 
@@
 
-```{r}
+```{r, out.width = 500}
 ## Step 2: Get predictions for data set when at grid
@@
 
+---
+
 ## Individual Conditional Expectation (ICE)
@@
 
-```{r}
+```{r, out.width = 500}
 res.raw$.id2 = factor(res.raw$.id)
@@
 
-## Partial Dependence Plot (PDP)
+--
 
-How does a feature influence the prediction for an individual observation **on average**?
-
-```{r}
+```{r, out.width = 500}
 p
 ```
+
+---
 
@@
 
-```{r}
+```{r, out.width = 500}
 res = res.raw %>% group_by(temp) %>%
@@
 
+---
+
 ## Partial Dependence Plot (PDP)
@@
 
+---
+
 ## Permutation Feature Importance (PFI)
@@
 How important is a feature for making correct predictions?
+
+---
 
@@
 
-# Difference between model-agnostic and model-specific interpetation
+---
+
+## Difference between model-agnostic and model-specific interpetation
 
@@
 
+---
+
 # Stuff I Did for My Thesis
 
-## Most Important
+---
+
+## Most Impactful Work
+
+```{r, out.width = 300}
+knitr::include_graphics("../images/book.jpg")
+```
+
+---
 
@@
 
-# Paper: Subgroups
-
-## Problem: Dependent features
-
-Show example with dependent features for PFI
-
-## Solution: Conditional variants
-
-Show solution
-
-* But: Not meaningful for PDP, or other measure more useful (LOESS)
-* But: Conditioning unclear
-* But: Intepretation difficult
-
-## Our solution: Permute in subgroups
-
-TODO: Show algorithm
-
-## Some results from our paper
-
-TODO: Show results or application
-
-# Paper: IML Inference
-
-## Problem: Uncertainty quantification
-
-PDP and PFI and most other just descriptive tools
-Often practictioner draw confidence intervals
-but only based on monte carlo approximation
-
-* But what is even the ground truth?
-* And what about the uncertainty in the model fit?
-* How could we estimate the uncertainty?
-
-## Solution: Treat as statistical estimator
-
-## How to Compute the variance
-
-## Problem of too low variance estimates
-
-# Paper: Quantifying
-
-TODO: Copy as much as possible from the talk that already exists for this paper
-
-Motivation: want to quantify in a model-agnostic way how complex a model is
-
-## Functional Decomposition
-
-Explain how functional decomposition works
-
-## ALE
-
-Formula:
-
-TODO: For estimation show the images
-
-## The three measures
-
-* Mean effect complexity
-* non linearity
-* number of features
-
-## Using in benchmarking (multi-objective)
+---
 
 # Inference of Feature Effects and Importance in Machine Learning
+
+---
+
+## Goal of Paper
+
+Treat permutation feature importance (PFI) and partial dependence plots (PDP) as statistical estimators.
+
+---
+
+## Statistical Modeling
+
+- Assumptions made about the data-generating process
+- Mapping between model parameters and properties of data-generating process
+- Statistical Inference possible
+
+TODO: Draw image which shows link between model and DGP
+
+---
 
@@
 - For inference, practitioner rely on interpretable models
-- remember: interpretable models have parameters that map to data-generating process
-- but: situations where more flexible models match DGP better (measured in test loss)
-- we have: model-agnostic interpretation
-- But: only describes the model behavior
+- But: situations where more flexible models match DGP better (measured in test loss)
+- Interpretation possible with PDP and PFI (among others)
+- However only describes the model behavior
 - Question: Can we link PD and PFI to the DGP? and even do inference?
 
-## Example
-
-TODO: Show PDP and PFI from some papers?
+---
 
@@
 
-TODO: simple image
+TODO: simple image with DGP + model. External tools, PDP and PFI, applied externally and compared.
+
+---
 
@@
 
+---
+
 ## Model-PD and Model-PFI
@@
 Show definition plus confidence intervals plus image
+
+---
 
@@
 But there is a problem when estimating this variance
+
+---
 
@@
 
-## Show simulation results
+---
 
-# Thank You for Your Attention
+## Thank You for Your Attention
+
+```{r, out.width = 500}
+knitr::include_graphics("../images/applause.gif")
+```

### 2021-08-23 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
finish slides

@@
 output:
-  xaringan::moon_reader
+  xaringan::moon_reader:
+    css: ["default", "default-fonts", "references.css"]
 ---
@@
 - Statistical Inference for PDP and PFI
-
----
-
-# Introduction
 
@@
 
-## A Brief History of Interpretable Modeling
+## Interpretable Models
 
-Interpretable ML has 200 year old roots, but consolidation only started in 2016.
-
-TODO: Web Of Science for keywords XAI and IML (written out)
-
----
-
-## Why Interpretability?
-
-```{r, out.width = 500}
-knitr::include_graphics("../images/explain.png")
+```{r, echo = FALSE}
+knitr::include_graphics("../images/white-box.png")
 ```
 
----
-
-## Interpretable Models
+???
 
@@
 ---
-
-## Interpretable Models
-
-```{r, echo = FALSE}
-knitr::include_graphics("../images/white-box.png")
-```
-
----
+class: center, middle
+background-color: darkgrey
 
@@
 
-## Model-agnostic Interpretation
+## Model-agnostic Methods
+
+```{r, echo=FALSE}
+knitr::include_graphics("../images/agnostic-black-box.png")
+```
+
+???
 
@@
 ---
-
-## Model-agnostic Methods
-
-```{r, echo=FALSE}
-knitr::include_graphics("../images/agnostic-black-box.png")
-```
-
----
+class: center, middle
+background-color: darkgrey
 
@@
 
-How does a feature influence the prediction for an individual observation?
+How does a feature influence the prediction for an **individual data point**?
 
@@
 
-How does a feature influence the prediction for an individual observation **on average**?
+How does a feature influence the prediction **on average**?
 
@@
 The theoretical construct:
-$$ PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})] $$
+$$PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})]$$
 
@@
 
-The algorithm (simples form): CITE Breiman
+The algorithm (simples form):
 
@@
 ---
-
-## Difference between model-agnostic and model-specific interpetation
-
-* Analyzing components vs. analyzing behavior
-* Components not very powerful, unless they directly describe behavior
-* For complex model, components do not directly tell us about behavior of model
-* Behavior more comparable
-
----
+class: middle, center
+background-color: darkgrey
 
@@
 ---
+background-color: lightgrey
 
@@
 ---
+background-image:url(../images/thesis-summary.jpeg)
+background-size: contain
 
-## Summary of Thesis
+???
 
@@
 ---
+class: center, middle
+background-color: darkgrey
 
-# Inference of Feature Effects and Importance in Machine Learning
-
----
-
-## Goal of Paper
-
-Treat permutation feature importance (PFI) and partial dependence plots (PDP) as statistical estimators.
+# Treating PDP and PFI as Statistical Estimators
 
@@
 
+```{r, out.width = 500}
+knitr::include_graphics("../images/dgp-model-mapping.jpeg")
+```
+
+???
+
 - Assumptions made about the data-generating process
@@
 
-TODO: Draw image which shows link between model and DGP
-
 ---
 
-## Status-Quo
+## Linking external tools to DGP
 
+```{r, out.width = 500}
+knitr::include_graphics("../images/no-dgp-model-mapping.jpeg")
+```
+
+???
 - For inference, practitioner rely on interpretable models
@@
 - Question: Can we link PD and PFI to the DGP? and even do inference?
-
----
-
-## Linking external tools to DGP
-
 - Hope: Lift PDP and PFI to status of statistical estimator
@@
 
-TODO: simple image with DGP + model. External tools, PDP and PFI, applied externally and compared.
+---
+
+## Linking External Tools to DGP
+
+```{r, out.width = 500}
+knitr::include_graphics("../images/external-dgp-model-mapping.jpeg")
+```
 
@@
 ## Source of uncertainty
+
+```{r, out.width = 900}
+knitr::include_graphics("../images/pdp-pfi-deviations.png")
+```
+
+???
 
@@
 
-## Model-PD and Model-PFI
+## Learner-PD and Learner-PFI
 
-Show definition plus confidence intervals plus image
-
+```{r, out.width = 640}
+knitr::include_graphics("../images/pdp-pfi-ci.png")
+```
 ---
 
-## Learner-PD and Learner-PFI
+## One Problem: Variance Underestimation
 
-Show definition plus confidence intervals plus image
+```{r, out.width = 500}
+knitr::include_graphics("../images/overlapping-training-data.jpeg")
+```
 
-But there is a problem when estimating this variance
+-> Variance correction needed
 
----
-
-## Variance correction
+???
 
@@
 
+class: small
+
+## References
+
+- Molnar, C., Casalicchio, G., & Bischl, B. (2018). iml: An R package for Interpretable Machine Learning. Journal of Open Source Software, 3, 786.
+- Molnar, C.. (2018). Interpretable Machine Learning - A Guide for Making Black Box Models Explainable.
+- Molnar, C., Casalicchio, G., & Bischl, B.
+(2019).
+Quantifying Model Complexity via Functional Decomposition for Better Post-Hoc Interpretability.
+In Joint European Conference on Machine Learning and Knowledge Discovery in Databases (pp.
+193–204).
+- Dandl, S., Molnar, C., Binder, M., & Bischl, B.
+(2020).
+Multi-objective counterfactual explanations.
+In Bäck T.
+et al.
+(eds) Parallel Problem Solving from Nature – PPSN XVI.
+PPSN 2020.
+Lecture Notes in Computer Science (pp.
+448–469).
+- Molnar, C., König, G., Bischl, B., & Casalicchio, G. (2020). Model-agnostic Feature Importance and Effects with Dependent Features–A Conditional Subgroup Approach. arXiv preprint arXiv:2006.04628.
+- Molnar, C., König, G., Herbinger, J., Freiesleben, T., Dandl, S., Scholbeck, C., Casalicchio, G., Grosse-Wentrup, M., & Bischl, B.
+(2020).
+Pitfalls to Avoid when Interpreting Machine Learning Models.
+In ICML workshop XXAI: Extending Explainable AI Beyond Deep Models and Classifiers.
+- Molnar, C. (2013). Recursive partitioning by conditional inference. Department of Statistics, University of Munich: Munich, Germany.
+- Molnar, C., Casalicchio, G., & Bischl, B. (2020). Interpretable Machine Learning - A Brief History, State-of-the-Art and Challenges. arXiv preprint arXiv:2010.09337.
+- Molnar, C., Freiesleben, T., König, G., Casalicchio, G., Wright, M., & Bischl, B.
+(2021), Relating Partial Dependence Plot and Permutation Feature Importance to the Data-Generating Process (Working Title).
+Not even on arxiv yet.
+- Casalicchio, G., Molnar, C., & Bischl, B.
+(2018).
+Visualizing the feature importance for black box models.
+In Joint European Conference on Machine Learning and Knowledge Discovery in Databases (pp.
+655–670).
+- Scholbeck, C., Molnar, C., Heumann, C., Bischl, B., & Casalicchio, G.
+(2019).
+Sampling, intervention, prediction, aggregation: A generalized framework for model-agnostic interpretations.
+arXiv preprint arXiv:1904.03959.
+- König, G., Molnar, C., Bischl, B., & Grosse-Wentrup, M. (2021). Relative Feature Importance. In 2020 25th International Conference on Pattern Recognition (ICPR) (pp. 9318–9325).
+
+---
+
 ## Thank You for Your Attention
